---
title: "hurricane code"
author: "Jessica Broach"
date: "2026-08-14"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## packages

```{r}
library(dplyr)
library(mice)
```


## data cleaning

### 1st trimester hurricane exposure

```{r, read in file}
hurricane1 <- read.csv("/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/All_1st_Trimester_Storm_Effectsv2.csv")
```

```{r, make missing data NA}
hurricane1[hurricane1 == 99] <- NA

hurricane1$DATE_OF_BIRTH <- as.character(hurricane1$DATE_OF_BIRTH)
hurricane1$MENSES_DATE <- as.character(hurricane1$MENSES_DATE)
# Only replace "NULL" with NA in character columns
hurricane1[] <- lapply(hurricane1, function(col) {
  if (is.character(col)) {
    col[col == "NULL"] <- NA
  }
  return(col)
})

hurricane1[] <- lapply(hurricane1, function(col) {
  if (is.character(col)) {
    col[col == "U"] <- NA
  }
  return(col)
})
hurricane1$DATE_OF_BIRTH <- as.Date(hurricane1$DATE_OF_BIRTH, format = "%Y-%m-%d") 
hurricane1$MENSES_DATE <- as.Date(hurricane1$MENSES_DATE, format = "%Y-%m-%d")

hurricane1$FATHER_EDCODE[hurricane1$FATHER_EDCODE == 9] <- NA
hurricane1$MOTHER_EDCODE[hurricane1$MOTHER_EDCODE == 9] <- NA
hurricane1$PRINCIPAL_SRCPAY_CODE[hurricane1$PRINCIPAL_SRCPAY_CODE == 9] <- NA
```

#### create covariates

```{r, combine father/mother race}
#father
hurricane1 <- hurricane1 %>%
  mutate(father_race = case_when(
    Father_CalculatedRace %in% c(01, 1) ~ 1,
    Father_CalculatedRace %in% c(02, 2) ~ 2,
    Father_CalculatedRace %in% c(03, 3) ~ 3,
    Father_CalculatedRace %in% c(04, 4, 05, 5, 07, 7, 08, 8, 10, 11, 12) ~ 4,
    Father_CalculatedRace %in% c(06, 6, 13, 14, 15) ~ 5,
    Father_CalculatedRace %in% c(09, 9) ~ 6,
    Father_CalculatedRace == 20 ~ 7,
  ))

#mother
hurricane1 <- hurricane1 %>%
  mutate(mother_race = case_when(
    Mother_CalculatedRace %in% c(01, 1) ~ 1,
    Mother_CalculatedRace %in% c(02, 2) ~ 2,
    Mother_CalculatedRace %in% c(03, 3) ~ 3,
    Mother_CalculatedRace %in% c(04, 4, 05, 5, 07, 7, 08, 8, 10, 11, 12) ~ 4,
    Mother_CalculatedRace %in% c(06, 6, 13, 14, 15) ~ 5,
    Mother_CalculatedRace %in% c(09, 9) ~ 6,
    Mother_CalculatedRace == 20 ~ 7,
  ))

hurricane1 <- hurricane1 %>%
  mutate(mother_his = case_when(
    Mother_CalculatedHisp %in% c(0, 6) ~ 0,
    Mother_CalculatedHisp %in% c(1, 2, 3, 5) ~ 1,
  ))
```

```{r, categorize BMI}
hurricane1$PrePregnancy_BMI <- as.numeric(hurricane1$PrePregnancy_BMI)
hurricane1$PPBMI_cat <- cut(hurricane1$PrePregnancy_BMI,
                              breaks = c(-Inf, 18.5, 25, 30, Inf),
                              labels = c(1, 2, 3, 4),
)

```


```{r, change class type}
hurricane1$FATHER_EDCODE <- as.character(hurricane1$FATHER_EDCODE)
hurricane1$father_race <- as.character(hurricane1$father_race)
hurricane1$mother_race <- as.character(hurricane1$mother_race)
hurricane1$MOTHER_EDCODE <- as.character(hurricane1$MOTHER_EDCODE)
hurricane1$Father_CalculatedHisp <- as.character(hurricane1$Father_CalculatedHisp)
hurricane1$Mother_CalculatedHisp <- as.character(hurricane1$Mother_CalculatedHisp)
```


#### create hurricane exposure variables


```{r, any hurricane impact}
hurricane1 <- hurricane1 %>%
  rowwise() %>%
  mutate(any_storm = ifelse(
    any(c_across(c(
      "Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
      "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
      "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
      "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12"
    )) %in% c("Threat", "Impact", "Cleanup"), na.rm = TRUE),
    1, 0
  )) %>%
  ungroup()
```

```{r, threat}
library(dplyr)

hurricane1 <- hurricane1 %>%
  rowwise() %>%
  mutate(threat = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Threat", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```

```{r, impact}
hurricane1 <- hurricane1 %>%
  rowwise() %>%
  mutate(impact = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Impact", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```

```{r, cleanup}
hurricane1 <- hurricane1 %>%
  rowwise() %>%
  mutate(cleanup = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Cleanup", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```



#### create ptb and lbw

```{r, restrict lbw to term births}
hurricane1_lbw <- hurricane1
hurricane1_lbw <- subset(hurricane1_lbw, GESTATION_WEEKS >= 37)
```

```{r, create outcome variables}
hurricane1$ptb <- ifelse(hurricane1$GESTATION_WEEKS < 37, 1, 0)
hurricane1$vptb <- ifelse(hurricane1$GESTATION_WEEKS < 32, 1, 0)
hurricane1$lbw <- ifelse(hurricane1$BIRTH_WEIGHT_GRAMS < 2500, 1, 0)
hurricane1$vlbw <- ifelse(hurricane1$BIRTH_WEIGHT_GRAMS < 1500, 1, 0)
hurricane1_lbw$lbw <- ifelse(hurricane1_lbw$BIRTH_WEIGHT_GRAMS < 2500, 1, 0)
```

#### save data

```{r}
save(hurricane1, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane/hurricane1.Rdata")
save(hurricane1_lbw, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane/hurricane1_lbw.Rdata")
```


### 2 trimester hurricane exposure


```{r, read in file}
hurricane2 <- read.csv("/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/All_2nd_Trimester_Storm_Effects.csv")
```

```{r, make missing data NA}
hurricane2[hurricane2 == 99] <- NA

hurricane2$DATE_OF_BIRTH <- as.character(hurricane2$DATE_OF_BIRTH)
hurricane2$MENSES_DATE <- as.character(hurricane2$MENSES_DATE)
# Only replace "NULL" with NA in character columns
hurricane2[] <- lapply(hurricane2, function(col) {
  if (is.character(col)) {
    col[col == "NULL"] <- NA
  }
  return(col)
})

hurricane2[] <- lapply(hurricane2, function(col) {
  if (is.character(col)) {
    col[col == "U"] <- NA
  }
  return(col)
})
hurricane2$DATE_OF_BIRTH <- as.Date(hurricane2$DATE_OF_BIRTH, format = "%Y-%m-%d") 
hurricane2$MENSES_DATE <- as.Date(hurricane2$MENSES_DATE, format = "%Y-%m-%d")

hurricane2$FATHER_EDCODE[hurricane2$FATHER_EDCODE == 9] <- NA
hurricane2$MOTHER_EDCODE[hurricane2$MOTHER_EDCODE == 9] <- NA
hurricane2$PRINCIPAL_SRCPAY_CODE[hurricane2$PRINCIPAL_SRCPAY_CODE == 9] <- NA
```

#### create covariates

```{r, combine father/mother race}
#father
hurricane2 <- hurricane2 %>%
  mutate(father_race = case_when(
    Father_CalculatedRace %in% c(01, 1) ~ 1,
    Father_CalculatedRace %in% c(02, 2) ~ 2,
    Father_CalculatedRace %in% c(03, 3) ~ 3,
    Father_CalculatedRace %in% c(04, 4, 05, 5, 07, 7, 08, 8, 10, 11, 12) ~ 4,
    Father_CalculatedRace %in% c(06, 6, 13, 14, 15) ~ 5,
    Father_CalculatedRace %in% c(09, 9) ~ 6,
    Father_CalculatedRace == 20 ~ 7,
  ))

#mother
hurricane2 <- hurricane2 %>%
  mutate(mother_race = case_when(
    Mother_CalculatedRace %in% c(01, 1) ~ 1,
    Mother_CalculatedRace %in% c(02, 2) ~ 2,
    Mother_CalculatedRace %in% c(03, 3) ~ 3,
    Mother_CalculatedRace %in% c(04, 4, 05, 5, 07, 7, 08, 8, 10, 11, 12) ~ 4,
    Mother_CalculatedRace %in% c(06, 6, 13, 14, 15) ~ 5,
    Mother_CalculatedRace %in% c(09, 9) ~ 6,
    Mother_CalculatedRace == 20 ~ 7,
  ))

hurricane2 <- hurricane2 %>%
  mutate(mother_his = case_when(
    Mother_CalculatedHisp %in% c(0, 6) ~ 0,
    Mother_CalculatedHisp %in% c(1, 2, 3, 5) ~ 1,
  ))
```

```{r, categorize BMI}
hurricane2$PrePregnancy_BMI <- as.numeric(hurricane2$PrePregnancy_BMI)
hurricane2$PPBMI_cat <- cut(hurricane2$PrePregnancy_BMI,
                              breaks = c(-Inf, 18.5, 25, 30, Inf),
                              labels = c(1, 2, 3, 4),
)

```


```{r, change class type}
hurricane2$FATHER_EDCODE <- as.character(hurricane2$FATHER_EDCODE)
hurricane2$father_race <- as.character(hurricane2$father_race)
hurricane2$mother_race <- as.character(hurricane2$mother_race)
hurricane2$MOTHER_EDCODE <- as.character(hurricane2$MOTHER_EDCODE)
hurricane2$Father_CalculatedHisp <- as.character(hurricane2$Father_CalculatedHisp)
hurricane2$Mother_CalculatedHisp <- as.character(hurricane2$Mother_CalculatedHisp)
```


#### create hurricane exposure


```{r, any hurricane impact}
hurricane2 <- hurricane2 %>%
  rowwise() %>%
  mutate(any_storm = ifelse(
    any(c_across(c(
      "Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
      "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
      "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
      "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12"
    )) %in% c("Threat", "Impact", "Cleanup"), na.rm = TRUE),
    1, 0
  )) %>%
  ungroup()
```

```{r, threat}
library(dplyr)

hurricane2 <- hurricane2 %>%
  rowwise() %>%
  mutate(threat = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Threat", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```

```{r, impact}
hurricane2 <- hurricane2 %>%
  rowwise() %>%
  mutate(impact = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Impact", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```

```{r, cleanup}
hurricane2 <- hurricane2 %>%
  rowwise() %>%
  mutate(cleanup = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Cleanup", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```



#### create ptb and lbw

```{r, restrict lbw to term births}
hurricane2_lbw <- hurricane2
hurricane2_lbw <- subset(hurricane2_lbw, GESTATION_WEEKS >= 37)
```

```{r, create outcome variables}
hurricane2$ptb <- ifelse(hurricane2$GESTATION_WEEKS < 37, 1, 0)
hurricane2$vptb <- ifelse(hurricane2$GESTATION_WEEKS < 32, 1, 0)
hurricane2$lbw <- ifelse(hurricane2$BIRTH_WEIGHT_GRAMS < 2500, 1, 0)
hurricane2$vlbw <- ifelse(hurricane2$BIRTH_WEIGHT_GRAMS < 1500, 1, 0)
hurricane2_lbw$lbw <- ifelse(hurricane2_lbw$BIRTH_WEIGHT_GRAMS < 2500, 1, 0)
```

#### save data

```{r}
save(hurricane2, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane/hurricane2.Rdata")
save(hurricane2_lbw, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane/hurricane2_lbw.Rdata")
```


### 3rd trimester hurricane exposure

```{r, read in file}
hurricane3 <- read.csv("/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/All_3rd_Trimester_Storm_Effects.csv")
```


```{r, make missing data NA}
hurricane3[hurricane3 == 99] <- NA

hurricane3$DATE_OF_BIRTH <- as.character(hurricane3$DATE_OF_BIRTH)
hurricane3$MENSES_DATE <- as.character(hurricane3$MENSES_DATE)
# Only replace "NULL" with NA in character columns
hurricane3[] <- lapply(hurricane3, function(col) {
  if (is.character(col)) {
    col[col == "NULL"] <- NA
  }
  return(col)
})

hurricane3[] <- lapply(hurricane3, function(col) {
  if (is.character(col)) {
    col[col == "U"] <- NA
  }
  return(col)
})
hurricane3$DATE_OF_BIRTH <- as.Date(hurricane3$DATE_OF_BIRTH, format = "%Y-%m-%d") 
hurricane3$MENSES_DATE <- as.Date(hurricane3$MENSES_DATE, format = "%Y-%m-%d")

hurricane3$FATHER_EDCODE[hurricane3$FATHER_EDCODE == 9] <- NA
hurricane3$MOTHER_EDCODE[hurricane3$MOTHER_EDCODE == 9] <- NA
hurricane3$PRINCIPAL_SRCPAY_CODE[hurricane3$PRINCIPAL_SRCPAY_CODE == 9] <- NA
```

#### create covariates

```{r, combine father/mother race}
#father
hurricane3 <- hurricane3 %>%
  mutate(father_race = case_when(
    Father_CalculatedRace %in% c(01, 1) ~ 1,
    Father_CalculatedRace %in% c(02, 2) ~ 2,
    Father_CalculatedRace %in% c(03, 3) ~ 3,
    Father_CalculatedRace %in% c(04, 4, 05, 5, 07, 7, 08, 8, 10, 11, 12) ~ 4,
    Father_CalculatedRace %in% c(06, 6, 13, 14, 15) ~ 5,
    Father_CalculatedRace %in% c(09, 9) ~ 6,
    Father_CalculatedRace == 20 ~ 7,
  ))

#mother
hurricane3 <- hurricane3 %>%
  mutate(mother_race = case_when(
    Mother_CalculatedRace %in% c(01, 1) ~ 1,
    Mother_CalculatedRace %in% c(02, 2) ~ 2,
    Mother_CalculatedRace %in% c(03, 3) ~ 3,
    Mother_CalculatedRace %in% c(04, 4, 05, 5, 07, 7, 08, 8, 10, 11, 12) ~ 4,
    Mother_CalculatedRace %in% c(06, 6, 13, 14, 15) ~ 5,
    Mother_CalculatedRace %in% c(09, 9) ~ 6,
    Mother_CalculatedRace == 20 ~ 7,
  ))

hurricane3 <- hurricane3 %>%
  mutate(mother_his = case_when(
    Mother_CalculatedHisp %in% c(0, 6) ~ 0,
    Mother_CalculatedHisp %in% c(1, 2, 3, 5) ~ 1,
  ))
```

```{r, categorize BMI}
hurricane3$PrePregnancy_BMI <- as.numeric(hurricane3$PrePregnancy_BMI)
hurricane3$PPBMI_cat <- cut(hurricane3$PrePregnancy_BMI,
                              breaks = c(-Inf, 18.5, 25, 30, Inf),
                              labels = c(1, 2, 3, 4),
)

```


```{r, change class type}
hurricane3$FATHER_EDCODE <- as.character(hurricane3$FATHER_EDCODE)
hurricane3$father_race <- as.character(hurricane3$father_race)
hurricane3$mother_race <- as.character(hurricane3$mother_race)
hurricane3$MOTHER_EDCODE <- as.character(hurricane3$MOTHER_EDCODE)
hurricane3$Father_CalculatedHisp <- as.character(hurricane3$Father_CalculatedHisp)
hurricane3$Mother_CalculatedHisp <- as.character(hurricane3$Mother_CalculatedHisp)
```

#### create hurricane exposure variables

```{r, any hurricane impact}
hurricane3 <- hurricane3 %>%
  rowwise() %>%
  mutate(any_storm = ifelse(
    any(c_across(c(
      "Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
      "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
      "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
      "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12"
    )) %in% c("Threat", "Impact", "Cleanup"), na.rm = TRUE),
    1, 0
  )) %>%
  ungroup()
```

```{r, threat}
hurricane3 <- hurricane3 %>%
  rowwise() %>%
  mutate(threat = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Threat", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```

```{r, impact}
hurricane3 <- hurricane3 %>%
  rowwise() %>%
  mutate(impact = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Impact", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```

```{r, cleanup}
hurricane3 <- hurricane3 %>%
  rowwise() %>%
  mutate(cleanup = ifelse(
    any(c_across(c("Storm_Effect_1", "Storm_Effect_2", "Storm_Effect_3",
                   "Storm_Effect_4", "Storm_Effect_5", "Storm_Effect_6",
                   "Storm_Effect_7", "Storm_Effect_8", "Storm_Effect_9",
                   "Storm_Effect_10", "Storm_Effect_11", "Storm_Effect_12")) == "Cleanup", 
        na.rm = TRUE), 
    1, 0)) %>%
  ungroup()

```


#### create ptb and lbw

```{r, restrict lbw to term births}
hurricane3_lbw <- hurricane3
hurricane3_lbw <- subset(hurricane3_lbw, GESTATION_WEEKS >= 37)
```

```{r, create outcome variables}
hurricane3$ptb <- ifelse(hurricane3$GESTATION_WEEKS < 37, 1, 0)
hurricane3$vptb <- ifelse(hurricane3$GESTATION_WEEKS < 32, 1, 0)
hurricane3$lbw <- ifelse(hurricane3$BIRTH_WEIGHT_GRAMS < 2500, 1, 0)
hurricane3$vlbw <- ifelse(hurricane3$BIRTH_WEIGHT_GRAMS < 1500, 1, 0)
hurricane3_lbw$lbw <- ifelse(hurricane3_lbw$BIRTH_WEIGHT_GRAMS < 2500, 1, 0)
```

#### save data

```{r}
save(hurricane3, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane/hurricane3.Rdata")
save(hurricane3_lbw, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane/hurricane3_lbw.Rdata")
```


## multiple imputation

### 1st trimester

```{r}
#subset columns needed
h1imp <- hurricane1 %>% 
  select(ID, final_lat, final_lon, PrePregnancy_BMI, threat, impact, cleanup, any_storm, ALCOHOL_USE, FATHER_EDCODE, MENSES_DATE, MOTHER_EDCODE, MOTHER_AGE, MOTHER_WIC_YESNO, MR_PREV_PRETERM, PPBMI_cat, MOTHER_MARRIED, PRINCIPAL_SRCPAY_CODE, Father_CalculatedHisp, Mother_CalculatedHisp, MR_DIAB_GEST, MR_HYPERT_ECLAMPSIA, MR_HYPERT_PREG, mother_race, father_race, Birth_Year, BIRTH_WEIGHT_GRAMS, Birth_Date, GESTATION_WEEKS, ptb, vptb lbw, vlbw)

#factor character columns
h1imp$ALCOHOL_USE <- as.factor(h1imp$ALCOHOL_USE)
h1imp$FATHER_EDCODE <- as.factor(h1imp$FATHER_EDCODE)
h1imp$MOTHER_EDCODE <- as.factor(h1imp$MOTHER_EDCODE)
h1imp$MOTHER_WIC_YESNO <- as.factor(h1imp$MOTHER_WIC_YESNO)
h1imp$MR_PREV_PRETERM <- as.factor(h1imp$MR_PREV_PRETERM)
h1imp$MOTHER_MARRIED <- as.factor(h1imp$MOTHER_MARRIED)
h1imp$PRINCIPAL_SRCPAY_CODE <- as.factor(h1imp$PRINCIPAL_SRCPAY_CODE)
h1imp$Father_CalculatedHisp <- as.factor(h1imp$Father_CalculatedHisp)
h1imp$Mother_CalculatedHisp <- as.factor(h1imp$Mother_CalculatedHisp)
h1imp$mother_race <- as.factor(h1imp$mother_race)
h1imp$father_race <- as.factor(h1imp$father_race)

#imputation
imp1 <- mice(
  h1imp,
  m = 20, 
  maxit = 20,
  seed = 12345)
```


```{r}
#subset columns needed
h1imp_lbw <- hurricane1_lbw %>% 
  select(ID, final_lat, final_lon, PrePregnancy_BMI, threat, impact, cleanup, any_storm, ALCOHOL_USE, FATHER_EDCODE, MENSES_DATE, MOTHER_EDCODE, MOTHER_AGE, MOTHER_WIC_YESNO, MR_PREV_PRETERM, PPBMI_cat, MOTHER_MARRIED, PRINCIPAL_SRCPAY_CODE, Father_CalculatedHisp, Mother_CalculatedHisp, MR_DIAB_GEST, MR_HYPERT_ECLAMPSIA, MR_HYPERT_PREG, mother_race, father_race, Birth_Year, BIRTH_WEIGHT_GRAMS, Birth_Date, GESTATION_WEEKS, lbw)

#factor character columns
h1imp_lbw$ALCOHOL_USE <- as.factor(h1imp_lbw$ALCOHOL_USE)
h1imp_lbw$FATHER_EDCODE <- as.factor(h1imp_lbw$FATHER_EDCODE)
h1imp_lbw$MOTHER_EDCODE <- as.factor(h1imp_lbw$MOTHER_EDCODE)
h1imp_lbw$MOTHER_WIC_YESNO <- as.factor(h1imp_lbw$MOTHER_WIC_YESNO)
h1imp_lbw$MR_PREV_PRETERM <- as.factor(h1imp_lbw$MR_PREV_PRETERM)
h1imp_lbw$MOTHER_MARRIED <- as.factor(h1imp_lbw$MOTHER_MARRIED)
h1imp_lbw$PRINCIPAL_SRCPAY_CODE <- as.factor(h1imp_lbw$PRINCIPAL_SRCPAY_CODE)
h1imp_lbw$Father_CalculatedHisp <- as.factor(h1imp_lbw$Father_CalculatedHisp)
h1imp_lbw$Mother_CalculatedHisp <- as.factor(h1imp_lbw$Mother_CalculatedHisp)
h1imp_lbw$mother_race <- as.factor(h1imp_lbw$mother_race)
h1imp_lbw$father_race <- as.factor(h1imp_lbw$father_race)

#imputation
imp1_lbw <- mice(
  h1imp_lbw,
  m = 20, 
  maxit = 20,
  seed = 12345)
```


```{r}
save(hurricane1imp, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/imp1.Rdata")
save(hurricane1imp_lbw, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/imp1_lbw.Rdata")
```



### 2nd trimester

```{r}
#subset columns needed
h2imp <- hurricane2 %>% 
  select(ID, final_lat, final_lon, PrePregnancy_BMI, threat, impact, cleanup, any_storm, ALCOHOL_USE, FATHER_EDCODE, MENSES_DATE, MOTHER_EDCODE, MOTHER_AGE, MOTHER_WIC_YESNO, MR_PREV_PRETERM, PPBMI_cat, MOTHER_MARRIED, PRINCIPAL_SRCPAY_CODE, Father_CalculatedHisp, Mother_CalculatedHisp, MR_DIAB_GEST, MR_HYPERT_ECLAMPSIA, MR_HYPERT_PREG, mother_race, father_race, Birth_Year, BIRTH_WEIGHT_GRAMS, Birth_Date, GESTATION_WEEKS, ptb, vptb lbw, vlbw)

#factor character columns
h2imp$ALCOHOL_USE <- as.factor(h2imp$ALCOHOL_USE)
h2imp$FATHER_EDCODE <- as.factor(h2imp$FATHER_EDCODE)
h2imp$MOTHER_EDCODE <- as.factor(h2imp$MOTHER_EDCODE)
h2imp$MOTHER_WIC_YESNO <- as.factor(h2imp$MOTHER_WIC_YESNO)
h2imp$MR_PREV_PRETERM <- as.factor(h2imp$MR_PREV_PRETERM)
h2imp$MOTHER_MARRIED <- as.factor(h2imp$MOTHER_MARRIED)
h2imp$PRINCIPAL_SRCPAY_CODE <- as.factor(h2imp$PRINCIPAL_SRCPAY_CODE)
h2imp$Father_CalculatedHisp <- as.factor(h2imp$Father_CalculatedHisp)
h2imp$Mother_CalculatedHisp <- as.factor(h2imp$Mother_CalculatedHisp)
h2imp$mother_race <- as.factor(h2imp$mother_race)
h2imp$father_race <- as.factor(h2imp$father_race)

#imputation
imp2 <- mice(
  h2imp,
  m = 20, 
  maxit = 20,
  seed = 12345)
```


```{r}
#subset columns needed
h2imp_lbw <- hurricane2_lbw %>% 
  select(ID, final_lat, final_lon, PrePregnancy_BMI, threat, impact, cleanup, any_storm, ALCOHOL_USE, FATHER_EDCODE, MENSES_DATE, MOTHER_EDCODE, MOTHER_AGE, MOTHER_WIC_YESNO, MR_PREV_PRETERM, PPBMI_cat, MOTHER_MARRIED, PRINCIPAL_SRCPAY_CODE, Father_CalculatedHisp, Mother_CalculatedHisp, MR_DIAB_GEST, MR_HYPERT_ECLAMPSIA, MR_HYPERT_PREG, mother_race, father_race, Birth_Year, BIRTH_WEIGHT_GRAMS, Birth_Date, GESTATION_WEEKS, lbw)

#factor character columns
h2imp_lbw$ALCOHOL_USE <- as.factor(h2imp_lbw$ALCOHOL_USE)
h2imp_lbw$FATHER_EDCODE <- as.factor(h2imp_lbw$FATHER_EDCODE)
h2imp_lbw$MOTHER_EDCODE <- as.factor(h2imp_lbw$MOTHER_EDCODE)
h2imp_lbw$MOTHER_WIC_YESNO <- as.factor(h2imp_lbw$MOTHER_WIC_YESNO)
h2imp_lbw$MR_PREV_PRETERM <- as.factor(h2imp_lbw$MR_PREV_PRETERM)
h2imp_lbw$MOTHER_MARRIED <- as.factor(h2imp_lbw$MOTHER_MARRIED)
h2imp_lbw$PRINCIPAL_SRCPAY_CODE <- as.factor(h2imp_lbw$PRINCIPAL_SRCPAY_CODE)
h2imp_lbw$Father_CalculatedHisp <- as.factor(h2imp_lbw$Father_CalculatedHisp)
h2imp_lbw$Mother_CalculatedHisp <- as.factor(h2imp_lbw$Mother_CalculatedHisp)
h2imp_lbw$mother_race <- as.factor(h2imp_lbw$mother_race)
h2imp_lbw$father_race <- as.factor(h2imp_lbw$father_race)

#imputation
imp2_lbw <- mice(
  h2imp_lbw,
  m = 20, 
  maxit = 20,
  seed = 12345)
```


```{r}
save(hurricane2imp, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/imp2.Rdata")
save(hurricane2imp_lbw, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/imp2_lbw.Rdata")
```



### 3rd trimester

```{r}
#subset columns needed
h3imp <- hurricane3 %>% 
  select(ID, final_lat, final_lon, PrePregnancy_BMI, threat, impact, cleanup, any_storm, ALCOHOL_USE, FATHER_EDCODE, MENSES_DATE, MOTHER_EDCODE, MOTHER_AGE, MOTHER_WIC_YESNO, MR_PREV_PRETERM, PPBMI_cat, MOTHER_MARRIED, PRINCIPAL_SRCPAY_CODE, Father_CalculatedHisp, Mother_CalculatedHisp, MR_DIAB_GEST, MR_HYPERT_ECLAMPSIA, MR_HYPERT_PREG, mother_race, father_race, Birth_Year, BIRTH_WEIGHT_GRAMS, Birth_Date, GESTATION_WEEKS, ptb, vptb lbw, vlbw)

#factor character columns
h3imp$ALCOHOL_USE <- as.factor(h3imp$ALCOHOL_USE)
h3imp$FATHER_EDCODE <- as.factor(h3imp$FATHER_EDCODE)
h3imp$MOTHER_EDCODE <- as.factor(h3imp$MOTHER_EDCODE)
h3imp$MOTHER_WIC_YESNO <- as.factor(h3imp$MOTHER_WIC_YESNO)
h3imp$MR_PREV_PRETERM <- as.factor(h3imp$MR_PREV_PRETERM)
h3imp$MOTHER_MARRIED <- as.factor(h3imp$MOTHER_MARRIED)
h3imp$PRINCIPAL_SRCPAY_CODE <- as.factor(h3imp$PRINCIPAL_SRCPAY_CODE)
h3imp$Father_CalculatedHisp <- as.factor(h3imp$Father_CalculatedHisp)
h3imp$Mother_CalculatedHisp <- as.factor(h3imp$Mother_CalculatedHisp)
h3imp$mother_race <- as.factor(h3imp$mother_race)
h3imp$father_race <- as.factor(h3imp$father_race)

#imputation
imp3 <- mice(
  h3imp,
  m = 20, 
  maxit = 20,
  seed = 12345)
```


```{r}
#subset columns needed
h3imp_lbw <- hurricane3_lbw %>% 
  select(ID, final_lat, final_lon, PrePregnancy_BMI, threat, impact, cleanup, any_storm, ALCOHOL_USE, FATHER_EDCODE, MENSES_DATE, MOTHER_EDCODE, MOTHER_AGE, MOTHER_WIC_YESNO, MR_PREV_PRETERM, PPBMI_cat, MOTHER_MARRIED, PRINCIPAL_SRCPAY_CODE, Father_CalculatedHisp, Mother_CalculatedHisp, MR_DIAB_GEST, MR_HYPERT_ECLAMPSIA, MR_HYPERT_PREG, mother_race, father_race, Birth_Year, BIRTH_WEIGHT_GRAMS, Birth_Date, GESTATION_WEEKS, lbw)

#factor character columns
h3imp_lbw$ALCOHOL_USE <- as.factor(h3imp_lbw$ALCOHOL_USE)
h3imp_lbw$FATHER_EDCODE <- as.factor(h3imp_lbw$FATHER_EDCODE)
h3imp_lbw$MOTHER_EDCODE <- as.factor(h3imp_lbw$MOTHER_EDCODE)
h3imp_lbw$MOTHER_WIC_YESNO <- as.factor(h3imp_lbw$MOTHER_WIC_YESNO)
h3imp_lbw$MR_PREV_PRETERM <- as.factor(h3imp_lbw$MR_PREV_PRETERM)
h3imp_lbw$MOTHER_MARRIED <- as.factor(h3imp_lbw$MOTHER_MARRIED)
h3imp_lbw$PRINCIPAL_SRCPAY_CODE <- as.factor(h3imp_lbw$PRINCIPAL_SRCPAY_CODE)
h3imp_lbw$Father_CalculatedHisp <- as.factor(h3imp_lbw$Father_CalculatedHisp)
h3imp_lbw$Mother_CalculatedHisp <- as.factor(h3imp_lbw$Mother_CalculatedHisp)
h3imp_lbw$mother_race <- as.factor(h3imp_lbw$mother_race)
h3imp_lbw$father_race <- as.factor(h3imp_lbw$father_race)

#imputation
imp3_lbw <- mice(
  h3imp_lbw,
  m = 20, 
  maxit = 20,
  seed = 12345)
```


```{r}
save(imp3, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/imp3.Rdata")
save(imp3_lbw, file = "/Users/jessicabroach/Library/CloudStorage/OneDrive-TulaneUniversity/Research Assistant/GRP/hurricane and pregnancy/imp3_lbw.Rdata")
```


## models

#trimester 1

## Gestational Diabetes

```{r}
library(mice)

# THREAT ------------------------------------------------------------

# Unadjusted
m_threat_unadj <- with(
  imp1,
  glm(MR_DIAB_GEST ~ threat, family = binomial)
)

pool_threat_unadj <- pool(m_threat_unadj)

summary(
  pool_threat_unadj,
  conf.int = TRUE,
  exponentiate = TRUE
)


# Adjusted
m_threat_adj <- with(
  imp1,
  glm(
    MR_DIAB_GEST ~ threat +
      ALCOHOL_USE +
      FATHER_EDCODE +
      MOTHER_EDCODE +
      MOTHER_AGE +
      MOTHER_WIC_YESNO +
      PPBMI_cat +
      MOTHER_MARRIED +
      PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp +
      Mother_CalculatedHisp +
      mother_race +
      father_race,
    family = binomial
  )
)

pool_threat_adj <- pool(m_threat_adj)

summary(
  pool_threat_adj,
  conf.int = TRUE,
  exponentiate = TRUE
)

# IMPACT ------------------------------------------------------------

m_impact_unadj <- with(
  imp1,
  glm(MR_DIAB_GEST ~ impact, family = binomial)
)

summary(
  pool(m_impact_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_impact_adj <- with(
  imp1,
  glm(
    MR_DIAB_GEST ~ impact +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_impact_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)


# CLEANUP -----------------------------------------------------------

m_cleanup_unadj <- with(
  imp1,
  glm(MR_DIAB_GEST ~ cleanup, family = binomial)
)

summary(
  pool(m_cleanup_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_cleanup_adj <- with(
  imp1,
  glm(
    MR_DIAB_GEST ~ cleanup +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_cleanup_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)


# ANY STORM ---------------------------------------------------------

m_any_storm_unadj <- with(
  imp1,
  glm(MR_DIAB_GEST ~ any_storm, family = binomial)
)

summary(
  pool(m_any_storm_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_any_storm_adj <- with(
  imp1,
  glm(
    MR_DIAB_GEST ~ any_storm +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_any_storm_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)
```

#Gestational Hypertension

```{r}
library(mice)

# THREAT
# Unadjusted
m_threat_unadj <- with(imp1, glm(MR_HYPERT_PREG ~ threat, family = binomial))

summary(pool(m_threat_unadj), conf.int = TRUE, exponentiate = TRUE)

# Adjusted
m_threat_adj <- with(imp1, glm(MR_HYPERT_PREG ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT

# Unadjusted
m_impact_unadj <- with(imp1, glm(MR_HYPERT_PREG ~ impact, family = binomial))

summary(pool(m_impact_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_impact_adj <- with(imp1,glm(MR_HYPERT_PREG ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP 

# Unadjusted
m_cleanup_unadj <- with(imp1, glm(MR_HYPERT_PREG ~ cleanup, family = binomial))

summary(pool(m_cleanup_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_cleanup_adj <- with(imp1, glm(MR_HYPERT_PREG ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM 

# Unadjusted
m_any_storm_unadj <- with(imp1, glm(MR_HYPERT_PREG ~ any_storm, family = binomial))

summary(pool(m_any_storm_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_any_storm_adj <- with(imp1, glm(MR_HYPERT_PREG ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


##preterm birth

```{r}
library(mice)

# THREAT
m_threat <- with(imp1, glm(ptb ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp1, glm(ptb ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp1, glm(ptb ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp1, glm(ptb ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp1, glm(ptb ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp1, glm(ptb ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp1, glm(ptb ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp1, glm(ptb ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


##very preterm birth

```{r}
library(mice)

# THREAT
m_threat <- with(imp1, glm(vptb ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp1, glm(vptb ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp1, glm(vptb ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp1, glm(vptb ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp1, glm(vptb ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp1, glm(vptb ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp1, glm(vptb ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp1, glm(vptb ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```

###term lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp1_lbw, glm(lbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp1_lbw, glm(lbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp1_lbw, glm(lbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp1_lbw, glm(lbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp1_lbw, glm(lbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp1_lbw, glm(lbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp1_lbw, glm(lbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp1_lbw, glm(lbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


## lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp1, glm(lbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp1, glm(lbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp1, glm(lbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp1, glm(lbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp1, glm(lbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp1, glm(lbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp1, glm(lbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp1, glm(lbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```

## very lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp1, glm(vlbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp1, glm(vlbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp1, glm(vlbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp1, glm(vlbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp1, glm(vlbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp1, glm(vlbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp1, glm(vlbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp1, glm(vlbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```

#trimester 2
## Gestational Diabetes

```{r}
library(mice)

# THREAT ------------------------------------------------------------

# Unadjusted
m_threat_unadj <- with(
  imp2,
  glm(MR_DIAB_GEST ~ threat, family = binomial)
)

pool_threat_unadj <- pool(m_threat_unadj)

summary(
  pool_threat_unadj,
  conf.int = TRUE,
  exponentiate = TRUE
)


# Adjusted
m_threat_adj <- with(
  imp2,
  glm(
    MR_DIAB_GEST ~ threat +
      ALCOHOL_USE +
      FATHER_EDCODE +
      MOTHER_EDCODE +
      MOTHER_AGE +
      MOTHER_WIC_YESNO +
      PPBMI_cat +
      MOTHER_MARRIED +
      PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp +
      Mother_CalculatedHisp +
      mother_race +
      father_race,
    family = binomial
  )
)

pool_threat_adj <- pool(m_threat_adj)

summary(
  pool_threat_adj,
  conf.int = TRUE,
  exponentiate = TRUE
)

# IMPACT ------------------------------------------------------------

m_impact_unadj <- with(
  imp2,
  glm(MR_DIAB_GEST ~ impact, family = binomial)
)

summary(
  pool(m_impact_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_impact_adj <- with(
  imp2,
  glm(
    MR_DIAB_GEST ~ impact +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_impact_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)


# CLEANUP -----------------------------------------------------------

m_cleanup_unadj <- with(
  imp2,
  glm(MR_DIAB_GEST ~ cleanup, family = binomial)
)

summary(
  pool(m_cleanup_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_cleanup_adj <- with(
  imp2,
  glm(
    MR_DIAB_GEST ~ cleanup +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_cleanup_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)


# ANY STORM ---------------------------------------------------------

m_any_storm_unadj <- with(
  imp2,
  glm(MR_DIAB_GEST ~ any_storm, family = binomial)
)

summary(
  pool(m_any_storm_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_any_storm_adj <- with(
  imp2,
  glm(
    MR_DIAB_GEST ~ any_storm +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_any_storm_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)
```

#Gestational Hypertension

```{r}
library(mice)

# THREAT
# Unadjusted
m_threat_unadj <- with(imp2, glm(MR_HYPERT_PREG ~ threat, family = binomial))

summary(pool(m_threat_unadj), conf.int = TRUE, exponentiate = TRUE)

# Adjusted
m_threat_adj <- with(imp2, glm(MR_HYPERT_PREG ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT

# Unadjusted
m_impact_unadj <- with(imp2, glm(MR_HYPERT_PREG ~ impact, family = binomial))

summary(pool(m_impact_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_impact_adj <- with(imp2,glm(MR_HYPERT_PREG ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP 

# Unadjusted
m_cleanup_unadj <- with(imp2, glm(MR_HYPERT_PREG ~ cleanup, family = binomial))

summary(pool(m_cleanup_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_cleanup_adj <- with(imp2, glm(MR_HYPERT_PREG ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM 

# Unadjusted
m_any_storm_unadj <- with(imp2, glm(MR_HYPERT_PREG ~ any_storm, family = binomial))

summary(pool(m_any_storm_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_any_storm_adj <- with(imp2, glm(MR_HYPERT_PREG ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


##preterm birth

```{r}
library(mice)

# THREAT
m_threat <- with(imp2, glm(ptb ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp2, glm(ptb ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp2, glm(ptb ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp2, glm(ptb ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp2, glm(ptb ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp2, glm(ptb ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp2, glm(ptb ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp2, glm(ptb ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


##very preterm birth

```{r}
library(mice)

# THREAT
m_threat <- with(imp2, glm(vptb ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp2, glm(vptb ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp2, glm(vptb ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp2, glm(vptb ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp2, glm(vptb ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp2, glm(vptb ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp2, glm(vptb ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp2, glm(vptb ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```

###term lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp2_lbw, glm(lbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp2_lbw, glm(lbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp2_lbw, glm(lbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp2_lbw, glm(lbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp2_lbw, glm(lbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp2_lbw, glm(lbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp2_lbw, glm(lbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp2_lbw, glm(lbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


## lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp2, glm(lbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp2, glm(lbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp2, glm(lbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp2, glm(lbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp2, glm(lbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp2, glm(lbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp2, glm(lbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp2, glm(lbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```

## very lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp2, glm(vlbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp2, glm(vlbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp2, glm(vlbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp2, glm(vlbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp2, glm(vlbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp2, glm(vlbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp2, glm(vlbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp2, glm(vlbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


#trimester 3

## Gestational Diabetes

```{r}
library(mice)

# THREAT ------------------------------------------------------------

# Unadjusted
m_threat_unadj <- with(
  imp3,
  glm(MR_DIAB_GEST ~ threat, family = binomial)
)

pool_threat_unadj <- pool(m_threat_unadj)

summary(
  pool_threat_unadj,
  conf.int = TRUE,
  exponentiate = TRUE
)


# Adjusted
m_threat_adj <- with(
  imp3,
  glm(
    MR_DIAB_GEST ~ threat +
      ALCOHOL_USE +
      FATHER_EDCODE +
      MOTHER_EDCODE +
      MOTHER_AGE +
      MOTHER_WIC_YESNO +
      PPBMI_cat +
      MOTHER_MARRIED +
      PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp +
      Mother_CalculatedHisp +
      mother_race +
      father_race,
    family = binomial
  )
)

pool_threat_adj <- pool(m_threat_adj)

summary(
  pool_threat_adj,
  conf.int = TRUE,
  exponentiate = TRUE
)

# IMPACT ------------------------------------------------------------

m_impact_unadj <- with(
  imp3,
  glm(MR_DIAB_GEST ~ impact, family = binomial)
)

summary(
  pool(m_impact_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_impact_adj <- with(
  imp3,
  glm(
    MR_DIAB_GEST ~ impact +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_impact_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)


# CLEANUP -----------------------------------------------------------

m_cleanup_unadj <- with(
  imp3,
  glm(MR_DIAB_GEST ~ cleanup, family = binomial)
)

summary(
  pool(m_cleanup_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_cleanup_adj <- with(
  imp3,
  glm(
    MR_DIAB_GEST ~ cleanup +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_cleanup_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)


# ANY STORM ---------------------------------------------------------

m_any_storm_unadj <- with(
  imp3,
  glm(MR_DIAB_GEST ~ any_storm, family = binomial)
)

summary(
  pool(m_any_storm_unadj),
  conf.int = TRUE,
  exponentiate = TRUE
)


m_any_storm_adj <- with(
  imp3,
  glm(
    MR_DIAB_GEST ~ any_storm +
      ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE +
      MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat +
      MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE +
      Father_CalculatedHisp + Mother_CalculatedHisp +
      mother_race + father_race,
    family = binomial
  )
)

summary(
  pool(m_any_storm_adj),
  conf.int = TRUE,
  exponentiate = TRUE
)
```

#Gestational Hypertension

```{r}
library(mice)

# THREAT
# Unadjusted
m_threat_unadj <- with(imp3, glm(MR_HYPERT_PREG ~ threat, family = binomial))

summary(pool(m_threat_unadj), conf.int = TRUE, exponentiate = TRUE)

# Adjusted
m_threat_adj <- with(imp3, glm(MR_HYPERT_PREG ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT

# Unadjusted
m_impact_unadj <- with(imp3, glm(MR_HYPERT_PREG ~ impact, family = binomial))

summary(pool(m_impact_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_impact_adj <- with(imp3,glm(MR_HYPERT_PREG ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP 

# Unadjusted
m_cleanup_unadj <- with(imp3, glm(MR_HYPERT_PREG ~ cleanup, family = binomial))

summary(pool(m_cleanup_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_cleanup_adj <- with(imp3, glm(MR_HYPERT_PREG ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM 

# Unadjusted
m_any_storm_unadj <- with(imp3, glm(MR_HYPERT_PREG ~ any_storm, family = binomial))

summary(pool(m_any_storm_unadj), conf.int = TRUE, exponentiate = TRUE)


# Adjusted
m_any_storm_adj <- with(imp3, glm(MR_HYPERT_PREG ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


##preterm birth

```{r}
library(mice)

# THREAT
m_threat <- with(imp3, glm(ptb ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp3, glm(ptb ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp3, glm(ptb ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp3, glm(ptb ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp3, glm(ptb ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp3, glm(ptb ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp3, glm(ptb ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp3, glm(ptb ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


##very preterm birth

```{r}
library(mice)

# THREAT
m_threat <- with(imp3, glm(vptb ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp3, glm(vptb ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp3, glm(vptb ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp3, glm(vptb ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp3, glm(vptb ~ threat + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp3, glm(vptb ~ impact + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp3, glm(vptb ~ cleanup + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp3, glm(vptb ~ any_storm + ALCOHOL_USE + FATHER_EDCODE + MOTHER_EDCODE + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + PPBMI_cat + MOTHER_MARRIED + PRINCIPAL_SRCPAY_CODE + Father_CalculatedHisp + Mother_CalculatedHisp + mother_race + father_race, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```

###term lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp3_lbw, glm(lbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp3_lbw, glm(lbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp3_lbw, glm(lbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp3_lbw, glm(lbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp3_lbw, glm(lbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp3_lbw, glm(lbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp3_lbw, glm(lbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp3_lbw, glm(lbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```


## lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp3, glm(lbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp3, glm(lbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp3, glm(lbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp3, glm(lbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp3, glm(lbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp3, glm(lbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp3, glm(lbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp3, glm(lbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```

## very lbw

```{r}
library(mice)

# THREAT
m_threat <- with(imp3, glm(vlbw ~ threat, family = binomial))

summary(pool(m_threat), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact <- with(imp3, glm(vlbw ~ impact, family = binomial))

summary(pool(m_impact), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup <- with(imp3, glm(vlbw ~ cleanup, family = binomial))

summary(pool(m_cleanup), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm <- with(imp3, glm(vlbw ~ any_storm, family = binomial))

summary(pool(m_any_storm), conf.int = TRUE, exponentiate = TRUE)
```

```{r}
library(mice)

# THREAT
m_threat_adj <- with(imp3, glm(vlbw ~ threat + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_threat_adj), conf.int = TRUE, exponentiate = TRUE)


# IMPACT
m_impact_adj <- with(imp3, glm(vlbw ~ impact + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_impact_adj), conf.int = TRUE, exponentiate = TRUE)


# CLEANUP
m_cleanup_adj <- with(imp3, glm(vlbw ~ cleanup + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_cleanup_adj), conf.int = TRUE, exponentiate = TRUE)


# ANY STORM
m_any_storm_adj <- with(imp3, glm(vlbw ~ any_storm + mother_race + ALCOHOL_USE + FATHER_EDCODE + Father_CalculatedHisp + Mother_CalculatedHisp + father_race + MOTHER_AGE + MOTHER_WIC_YESNO + MR_PREV_PRETERM + MOTHER_EDCODE + MOTHER_MARRIED + PPBMI_cat + PRINCIPAL_SRCPAY_CODE, family = binomial))

summary(pool(m_any_storm_adj), conf.int = TRUE, exponentiate = TRUE)
```
