---
title: "All-Cause Mortality & Storms"
output: html_document
editor_options:
  chunk_output_type: console
---

## Analytics of deaths and storms. Start with non-spatial analytics

Read the data table to a csv file
```{r}
Deaths.dt <- data.table::fread(here::here("data", "outputs", "All_Deaths_Storm_Effects.csv"))
Deaths.df <- Deaths.dt |>
  as.data.frame()
```

```{r}
library(dplyr)

total_deaths <- nrow(Deaths.df)
exposed_total <- Deaths.dt[Storm_Effect %in% c("Cleanup", "Impact", "Threat"), .N]

# Calculate death counts and rates for specified storm effects
death_rates <- Deaths.df |>
  filter(Storm_Effect %in% c("Cleanup", "Impact", "Threat")) |>
  group_by(Storm_Effect) |>
  summarise(Deaths = n(), .groups = "drop") |>
  mutate(Death_Rate = Deaths / exposed_total)

# View result
print(death_rates)
```

This tells you, for example, that .304% of all deaths occurred under an Impact storm condition.

## Death rates per day

Step 1: Get number of deaths per day by Storm_Effect
```{r}
# Group deaths by date and storm effect
daily_deaths <- Deaths.df |>
  group_by(Death_Date, Storm_Effect) |>
  summarise(deaths = n(), .groups = "drop")
```

Step 2: Get number of days for each storm effect
This step assumes Storm_Effect applies to the whole day — i.e., each Death_Date has a unique Storm_Effect
```{r}
num_days <- daily_deaths |>
  distinct(Death_Date, Storm_Effect) |>
  count(Storm_Effect, name = "num_days")
```

Step 3: Total deaths per storm effect
```{r}
total_deaths <- daily_deaths |>
  group_by(Storm_Effect) |>
  summarise(total_deaths = sum(deaths), .groups = "drop")
```

Step 4: Calculate daily death rate
```{r}
death_rates <- left_join(total_deaths, num_days, by = "Storm_Effect") |>
  mutate(daily_death_rate = total_deaths / num_days)

print(death_rates)
```

Test it statistically
```{r}
model_data <- left_join(total_deaths, num_days, by = "Storm_Effect")
model_data$Storm_Effect <- relevel(factor(model_data$Storm_Effect), ref = "None")

poisson_model <- glm(total_deaths ~ Storm_Effect + offset(log(num_days)), data = model_data, family = poisson())
summary(poisson_model)
```

Create a nice table of the model results
```{r}
# Extract coefficients and 95% confidence intervals
coefs <- coef(summary(poisson_model))
confint_vals <- confint(poisson_model)  # May take a few seconds

# Build data frame of results
rate_ratios <- data.frame(
  Term = rownames(coefs),
  Estimate = coefs[, "Estimate"],
  Std_Error = coefs[, "Std. Error"],
  z_value = coefs[, "z value"],
  p_value = coefs[, "Pr(>|z|)"],
  RR = exp(coefs[, "Estimate"]),
  RR_lower = exp(confint_vals[, 1]),
  RR_upper = exp(confint_vals[, 2])
)

# Clean up names
rownames(rate_ratios) <- NULL
rate_ratios <- rate_ratios |>
  dplyr::select(Term, RR, RR_lower, RR_upper, p_value)

# Print the table
print(rate_ratios)

```


