---
title: "Daily ER visits relative to storm effects"
output: html_document
editor_options:
  chunk_output_type: console
---

## Create an empirical storm effects model

```{r}
library(sf)
library(dplyr)

storm_intensity <- 34 # all tropical storms and hurricanes
begin_year <- 2004
end_year <- 2022

L <- "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/IBTrACS.NA.list.v04r00.lines.zip"
  
if(!"IBTrACS.NA.list.v04r00.lines.zip" %in% list.files(here::here("data"))) {
download.file(url = L,
              destfile = here::here("data",
                                    "IBTrACS.NA.list.v04r00.lines.zip"))
unzip(here::here("data", "IBTrACS.NA.list.v04r00.lines.zip"),
      exdir = here::here("data"))
}

Tracks.sf <- st_read(dsn = here::here("data"), 
                     layer = "IBTrACS.NA.list.v04r00.lines") |>
  st_transform(crs = 32616)

Tracks.sf <- Tracks.sf |>
  filter(year >= begin_year & year <= end_year) |>
  filter(USA_WIND >= storm_intensity) |>
  select(SID, SEASON, year, month, day, hour, min,
         NAME, SUBBASIN, ISO_TIME, USA_WIND, USA_PRES, USA_RMW, USA_EYE, USA_ROCI)

Tracks.sf |>
  st_drop_geometry() |>
  summarize(avgRMW = mean(USA_RMW, na.rm = TRUE) * 1.852,
                   avgEYE = mean(USA_EYE, na.rm = TRUE) * 1.852,
                   avgROCI = mean(USA_ROCI, na.rm = TRUE) * 1.852)

# Fill in missing RMW values. Start with pedigree, then use minimum pressure, and finish again with pedigree
Tracks.sf <- Tracks.sf |>
  group_by(SID) |>  # pedigree
  mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks.sf <- Tracks.sf |>
  group_by(USA_PRES) |> # minimum pressure
  mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks.sf <- Tracks.sf |>
  group_by(SID) |> # pedigree
  mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

#Add a buffer to the tracks to make segmented wind swaths
Swaths.sf <- Tracks.sf |>
  st_buffer(dist = Tracks.sf$USA_RMW * 1852) # 1852 converts to meters

# Wind swaths that cross Florida. `USAboundaries` package no longer maintained on CRAN
#devtools::install_github("ropensci/USAboundariesData")
#devtools::install_github("ropensci/USAboundaries", force = TRUE)
#install.packages("USAboundariesData", repos = "https://ropensci.r-universe.dev", type = "source")
Boundaries.sf <- USAboundaries::us_states(resolution = "low", states = "FL") |> 
  st_transform(crs = 32616)

X <- Swaths.sf |>
  st_intersects(Boundaries.sf, sparse = FALSE) #Does the swath intersect the state border?
Swaths.sf <- Swaths.sf[X, ]
Swaths.sf <- Swaths.sf |>
  mutate(Date = lubridate::as_date(ISO_TIME)) #Add a date column

# Extract the boundaries of storm impacts by storm ID. Dissolve the overlap borders of the individual swaths by storm to create a single storm swath. Add the storm category (Saffir-Simpson) based on wind speed
Swaths.sf <- Swaths.sf |>
  group_by(SID) |>
  summarize(Date0 = first(Date),
            NAME = first(NAME),
            Wind = max(USA_WIND),
            geometry = st_union(geometry)) |>
  mutate(Storm_Category = case_when(
    Wind >= 34 & Wind <= 63 ~ 0,
    Wind >= 64 & Wind <= 82 ~ 1,
    Wind >= 83 & Wind <= 95 ~ 2,
    Wind >= 96 & Wind <= 112 ~ 3,
    Wind >= 113 & Wind <= 136 ~ 4,
    Wind >= 137 ~ 5
  ))

# Transform the geometry to a geographic CRS (4326) and unionize the swaths
sf_use_s2(TRUE)

Swaths.sf <- Swaths.sf |>
    st_transform(crs = 4326)
union_Swaths2.sfg <- Swaths.sf |>
  st_union()

sum(table(Swaths.sf$Storm_Category))
table(Swaths.sf$Storm_Category)
table(lubridate::year(Swaths.sf$Date0))
```

Expand `Swaths.sf` by adding rows
```{r}
library(tidyr)
library(lubridate)

deltaT <- 1   # One day increment
n_new <- 8    # Number of days before & after impact

TIC.sf <- Swaths.sf |>
    rowwise() |>
    mutate(new_features = list(tidyr::tibble(
           Date = Date0 + (-n_new:n_new) * deltaT))) |>
    unnest(new_features) |>
    ungroup() |>
    select(Date, Storm_Name = NAME, Storm_Category)

TIC <- rep(rep(c("Threat", "Impact", "Cleanup"), times = c(n_new, 1, n_new)), 
           times = nrow(Swaths.sf))
TIC.sf$Storm_Effect <- TIC

# Add month and year indicators change name of the geometry column (this is needed for spatial merges)
TIC.sf <- TIC.sf |>
  mutate(Month = month(Date),
         Year = year(Date))
storm_months <- unique(TIC.sf$Month)
storm_year_range <- range(TIC.sf$Year)

st_geometry(TIC.sf) <- "geom"
```

## Build a zip x date effect calendar

```{r}
library(data.table)
library(tigris)
options(tigris_use_cache = TRUE)

# --- ZCTAs for Florida (2020) ---
zcta_us <- zctas(cb = TRUE, year = 2020) %>%
  select(zip = ZCTA5CE20, geometry)

fl <- states(cb = TRUE, year = 2020) %>%
  filter(STUSPS == "FL") %>%
  st_transform(st_crs(zcta_us))

zctas_fl <- st_join(zcta_us, fl, join = st_intersects, left = FALSE) %>%
  select(zip, geometry)

#Align coordinate reference systems
tic_aligned  <- st_transform(TIC.sf,  st_crs(zctas_fl))

# Spatial join: which zips intersect each TIC polygon (by date & effect)
# Keep only the columns we need from TIC
tic_keep <- tic_aligned |> 
  select(Date, Storm_Effect, geom)

# Point-in-polygon for polygons
zcta_date_effect_sf <- st_join(
  zctas_fl,
  tic_keep,
  join = st_intersects,
  left = FALSE   # keep only zips that intersect a TIC polygon on that date
)

# Drop geometry for the calendar
zcta_date_effect <- as.data.table(st_drop_geometry(zcta_date_effect_sf))
setnames(zcta_date_effect, c("zip","Date","Storm_Effect"),
         c("zip","date","effect"))

# Normalize types
zcta_date_effect[, `:=`(
  zip    = as.character(zip),
  date   = as.IDate(date),
  effect = as.character(effect)
)]

# Resolve multiple effects on the same zip-date: Choose most severe ranking (None < Cleanup < Threat < Impact)
sev <- data.table(effect = c("None","Cleanup","Threat","Impact"),
                  rank   = 0:3)
zde <- sev[zcta_date_effect, on = "effect"]

# Keep the single most-severe label per ZIP–date
zde <- zde[ , .SD[which.max(rank)], by = .(zip, date)][ , .(zip, date, effect)]

# Build the full zip x date data frame & fill missing days with "None"
all_zips  <- sort(unique(zde$zip))
all_dates <- as.IDate(seq(as.Date("2004-12-31"), as.Date("2022-12-31"), by = "day"))
grid <- CJ(zip = as.character(all_zips), date = all_dates)

# Left-join the observed effects; fill NA → "None"
effect_calendar <- zde[grid, on = .(zip, date)]
effect_calendar[is.na(effect), effect := "None"]

# Keep effect as a factor with ordered levels
effect_calendar[, effect := factor(effect, levels = c("None","Cleanup","Threat","Impact"))]

# Exactly one effect per ZIP–day?
stopifnot(effect_calendar[, .N, by = .(zip, date)][, all(N == 1)])

# Effects present?
effect_calendar[, table(effect)]
```

## Randomize zip x month
```{r}
library(dplyr)
library(lubridate)
library(data.table)

set.seed(123)   # for reproducibility

effect_calendar_zipmonth_random <- effect_calendar %>%
  mutate(month = floor_date(as.Date(date), unit = "month")) %>%  # create month block
  group_by(zip, month) %>%
  mutate(effect = sample(effect)) %>%      # shuffle effect within ZIP × month
  ungroup() %>%
  select(-month) %>%
  as.data.table()
```

## Merge with emergency room visits

Data received from David Hsu on October 16, 2025 via email
```{r}
ER_visits.dt <- data.table::fread(here::here("data", "Combined_ER_Data.csv"))
ER_visits.df <- ER_visits.dt |>
  as.data.frame()
```

Clean and merge
```{r}
setDT(ER_visits.df)

# Pad ZipCode to 5 chars; drop non-sensical codes (0/NA/too short/too big)
ER_visits.df[
  , ZipCode := fifelse(ZipCode > 0 & ZipCode < 1e6, sprintf("%05d", ZipCode), NA_character_)
]
ER_visits.df[, Date := as.IDate(Date)]

ER_visits_clean <- ER_visits.df[!is.na(ZipCode) & nchar(ZipCode) == 5]

# Standardize names; optionally collapse duplicates
setnames(ER_visits_clean, c("ZipCode","Date","Total_Visits"),
                           c("zip",    "date","er_visits"))

# If there can be multiple ER rows per (zip,date), aggregate to daily total
ER_visits_clean <- ER_visits_clean[, .(er_visits = sum(er_visits, na.rm = TRUE)),
                                   by = .(zip, date)]

merged <- merge(
  effect_calendar_zipmonth_random,
  ER_visits_clean,
  by = c("zip", "date"),
  all.x = TRUE  # keep all ZIP–dates from the calendar
)

merged[is.na(er_visits), er_visits := 0L]

library(lubridate)   # for yearmonth(), wday()
setDT(merged)

# Add day of week (factor, ordered Monday–Sunday)
merged[, dow := lubridate::wday(date, label = TRUE, abbr = TRUE)]

# Add year-month (IDate stored as "YYYY-MM-01" for convenience)
merged[, year_month := as.IDate(floor_date(date, unit = "month"))]
```

## Model using logit and Poisson separately

A two-part “threshold (≤6) + truncated count (>6)” approximation (stays in `fixest`). The process is modeled as: Part A (logit): probability the true count is ≤6 (we observe 6) and part B (Poisson on truncated support): conditional mean for counts >6

```{r}
library(fixest)

dat <- merged %>% mutate(
  is_le6 = as.integer(er_visits == 6L)) %>% # proxy for "true ≤ 6"
  filter(month(date) %in% 5:11)  # hurricane season only

# Part A: logistic model for being ≤ 6
mA <- feglm(
  is_le6 ~ i(effect, ref="None") | zip + year_month + dow,
  data = dat, family = binomial("logit"),
  cluster = ~ zip + year_month
)

# Part B: Poisson on Y>6 (treat observed counts >6 as exact)
d_gt6 <- filter(dat, er_visits > 6L)
mB <- fepois(
  er_visits ~ i(effect, ref="None") | zip + year_month + dow,
  data = d_gt6, cluster = ~ zip + year_month
)

summary(mA); summary(mB)
```
