---
title: "Daily ER visits relative to storm effects"
output: html_document
editor_options:
  chunk_output_type: console
---

## Create an empirical storm effects model

```{r}
library(sf)
library(dplyr)

storm_intensity <- 34 # all tropical storms and hurricanes
begin_year <- 2004
end_year <- 2022

L <- "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/IBTrACS.NA.list.v04r00.lines.zip"
  
if(!"IBTrACS.NA.list.v04r00.lines.zip" %in% list.files(here::here("data"))) {
download.file(url = L,
              destfile = here::here("data",
                                    "IBTrACS.NA.list.v04r00.lines.zip"))
unzip(here::here("data", "IBTrACS.NA.list.v04r00.lines.zip"),
      exdir = here::here("data"))
}

Tracks.sf <- st_read(dsn = here::here("data"), 
                     layer = "IBTrACS.NA.list.v04r00.lines") |>
  st_transform(crs = 32616)

Tracks.sf <- Tracks.sf |>
  filter(year >= begin_year & year <= end_year) |>
  filter(USA_WIND >= storm_intensity) |>
  select(SID, SEASON, year, month, day, hour, min,
         NAME, SUBBASIN, ISO_TIME, USA_WIND, USA_PRES, USA_RMW, USA_EYE, USA_ROCI)

Tracks.sf |>
  st_drop_geometry() |>
  summarize(avgRMW = mean(USA_RMW, na.rm = TRUE) * 1.852,
                   avgEYE = mean(USA_EYE, na.rm = TRUE) * 1.852,
                   avgROCI = mean(USA_ROCI, na.rm = TRUE) * 1.852)

# Fill in missing RMW values. Start with pedigree, then use minimum pressure, and finish again with pedigree
Tracks.sf <- Tracks.sf |>
  group_by(SID) |>  # pedigree
  mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks.sf <- Tracks.sf |>
  group_by(USA_PRES) |> # minimum pressure
  mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks.sf <- Tracks.sf |>
  group_by(SID) |> # pedigree
  mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

#Add a buffer to the tracks to make segmented wind swaths
Swaths.sf <- Tracks.sf |>
  st_buffer(dist = Tracks.sf$USA_RMW * 1852) # 1852 converts to meters

# Wind swaths that cross Florida. `USAboundaries` package no longer maintained on CRAN
#devtools::install_github("ropensci/USAboundariesData")
#devtools::install_github("ropensci/USAboundaries", force = TRUE)
#install.packages("USAboundariesData", repos = "https://ropensci.r-universe.dev", type = "source")
Boundaries.sf <- USAboundaries::us_states(resolution = "low", states = "FL") |> 
  st_transform(crs = 32616)

X <- Swaths.sf |>
  st_intersects(Boundaries.sf, sparse = FALSE) #Does the swath intersect the state border?
Swaths.sf <- Swaths.sf[X, ]
Swaths.sf <- Swaths.sf |>
  mutate(Date = lubridate::as_date(ISO_TIME)) #Add a date column

# Extract the boundaries of storm impacts by storm ID. Dissolve the overlap borders of the individual swaths by storm to create a single storm swath. Add the storm category (Saffir-Simpson) based on wind speed
Swaths.sf <- Swaths.sf |>
  group_by(SID) |>
  summarize(Date0 = first(Date),
            NAME = first(NAME),
            Wind = max(USA_WIND),
            geometry = st_union(geometry)) |>
  mutate(Storm_Category = case_when(
    Wind >= 34 & Wind <= 63 ~ 0,
    Wind >= 64 & Wind <= 82 ~ 1,
    Wind >= 83 & Wind <= 95 ~ 2,
    Wind >= 96 & Wind <= 112 ~ 3,
    Wind >= 113 & Wind <= 136 ~ 4,
    Wind >= 137 ~ 5
  ))

# Transform the geometry to a geographic CRS (4326) and unionize the swaths
sf_use_s2(TRUE)

Swaths.sf <- Swaths.sf |>
    st_transform(crs = 4326)
union_Swaths2.sfg <- Swaths.sf |>
  st_union()

sum(table(Swaths.sf$Storm_Category))
table(Swaths.sf$Storm_Category)
table(lubridate::year(Swaths.sf$Date0))
```

44
 0  1  2  3  4  5 
30  3  3  4  2  2 
2004 2005 2006 2007 2008 2009 2010 2011 2012 2013 2016 2017 2018 2019 
   5    6    2    1    1    2    2    1    3    1    3    3    3    1 
2020 2021 2022 
   3    4    3

Expand `Swaths.sf` by adding rows based on the increment value of the attribute `Date0`. Dates prior to the impact date are threat dates and those after the impact date are cleanup dates
```{r}
library(tidyr)
library(lubridate)

deltaT <- 1   # One day increment
n_new <- 8    # Number of days before & after impact

TIC.sf <- Swaths.sf |>
    rowwise() |>
    mutate(new_features = list(tidyr::tibble(
           Date = Date0 + (-n_new:n_new) * deltaT))) |>
    unnest(new_features) |>
    ungroup() |>
    select(Date, Storm_Name = NAME, Storm_Category)

TIC <- rep(rep(c("Threat", "Impact", "Cleanup"), times = c(n_new, 1, n_new)), 
           times = nrow(Swaths.sf))
TIC.sf$Storm_Effect <- TIC

# Add month and year indicators change name of the geometry column (this is needed for spatial merges)
TIC.sf <- TIC.sf |>
  mutate(Month = month(Date),
         Year = year(Date))
storm_months <- unique(TIC.sf$Month)
storm_year_range <- range(TIC.sf$Year)

st_geometry(TIC.sf) <- "geom"
```

## Build a zip x date effect calendar

```{r}
library(data.table)
library(tigris)
options(tigris_use_cache = TRUE)

# --- ZCTAs for Florida (2020) ---
zcta_us <- zctas(cb = TRUE, year = 2020) %>%
  select(zip = ZCTA5CE20, geometry)

fl <- states(cb = TRUE, year = 2020) %>%
  filter(STUSPS == "FL") %>%
  st_transform(st_crs(zcta_us))

zctas_fl <- st_join(zcta_us, fl, join = st_intersects, left = FALSE) %>%
  select(zip, geometry)

#Align coordinate reference systems
tic_aligned  <- st_transform(TIC.sf,  st_crs(zctas_fl))

# Spatial join: which zips intersect each TIC polygon (by date & effect)
# Keep only the columns we need from TIC
tic_keep <- tic_aligned |> 
  select(Date, Storm_Effect, geom)

# Point-in-polygon for polygons
zcta_date_effect_sf <- st_join(
  zctas_fl,
  tic_keep,
  join = st_intersects,
  left = FALSE   # keep only zips that intersect a TIC polygon on that date
)

# Drop geometry for the calendar
zcta_date_effect <- as.data.table(st_drop_geometry(zcta_date_effect_sf))
setnames(zcta_date_effect, c("zip","Date","Storm_Effect"),
         c("zip","date","effect"))

# Normalize types
zcta_date_effect[, `:=`(
  zip    = as.character(zip),
  date   = as.IDate(date),
  effect = as.character(effect)
)]

# Resolve multiple effects on the same zip-date: Choose most severe ranking (None < Cleanup < Threat < Impact)
sev <- data.table(effect = c("None","Cleanup","Threat","Impact"),
                  rank   = 0:3)
zde <- sev[zcta_date_effect, on = "effect"]

# Keep the single most-severe label per ZIP–date
zde <- zde[ , .SD[which.max(rank)], by = .(zip, date)][ , .(zip, date, effect)]

# Build the full zip x date data frame & fill missing days with "None"
all_zips  <- sort(unique(zde$zip))
all_dates <- as.IDate(seq(as.Date("2004-12-31"), as.Date("2022-12-31"), by = "day"))
grid <- CJ(zip = as.character(all_zips), date = all_dates)

# Left-join the observed effects; fill NA → "None"
effect_calendar <- zde[grid, on = .(zip, date)]
effect_calendar[is.na(effect), effect := "None"]

# Keep effect as a factor with ordered levels
effect_calendar[, effect := factor(effect, levels = c("None","Cleanup","Threat","Impact"))]

# Exactly one effect per ZIP–day?
stopifnot(effect_calendar[, .N, by = .(zip, date)][, all(N == 1)])

# Effects present?
effect_calendar[, table(effect)]
```

effect
   None Cleanup  Threat  Impact 
6740950   64230   64240    8030 

## Merge with emergency room visits

Data received from David Hsu on October 16, 2025 via email
```{r}
ER_visits.dt <- data.table::fread(here::here("data", "Combined_ER_Data.csv"))
ER_visits.df <- ER_visits.dt |>
  as.data.frame()

setDT(ER_visits.df)

# Pad ZipCode to 5 chars; drop non-sensical codes (0/NA/too short/too big)
ER_visits.df[
  , ZipCode := fifelse(ZipCode > 0 & ZipCode < 1e6, sprintf("%05d", ZipCode), NA_character_)
]
ER_visits.df[, Date := as.IDate(Date)]

ER_visits_clean <- ER_visits.df[!is.na(ZipCode) & nchar(ZipCode) == 5]

# Standardize names; optionally collapse duplicates
setnames(ER_visits_clean, c("ZipCode","Date","Total_Visits"),
                           c("zip",    "date","er_visits"))

# If there can be multiple ER rows per (zip,date), aggregate to daily total
ER_visits_clean <- ER_visits_clean[, .(er_visits = sum(er_visits, na.rm = TRUE)),
                                   by = .(zip, date)]

merged <- merge(
  effect_calendar,
  ER_visits_clean,
  by = c("zip", "date"),
  all.x = TRUE  # keep all ZIP–dates from the calendar
)

merged[is.na(er_visits), er_visits := 0L]

library(lubridate)   # for yearmonth(), wday()
setDT(merged)

# Add day of week (factor, ordered Monday–Sunday)
merged[, dow := lubridate::wday(date, label = TRUE, abbr = TRUE)]

# Add year-month (IDate stored as "YYYY-MM-01" for convenience)
merged[, year_month := as.IDate(floor_date(date, unit = "month"))]
```

## Model using logit and Poisson separately

A two-part “threshold (≤6) + truncated count (>6)” approximation (stays in `fixest`). The process is modeled as: Part A (logit): probability the true count is ≤6 (we observe 6) and part B (Poisson on truncated support): conditional mean for counts >6

```{r}
library(fixest)

dat <- merged %>% mutate(
  is_le6 = as.integer(er_visits == 6L)) %>% # proxy for "true ≤ 6"
  filter(month(date) %in% 5:11)  # hurricane season only

# Part A: logistic model for being ≤ 6
mA <- feglm(
  is_le6 ~ i(effect, ref="None") | zip + year_month + dow,
  data = dat, family = binomial("logit"),
  cluster = ~ zip + year_month
)

# Part B: Poisson on Y>6 (treat observed counts >6 as exact)
d_gt6 <- filter(dat, er_visits > 6L)
mB <- fepois(
  er_visits ~ i(effect, ref="None") | zip + year_month + dow,
  data = d_gt6, cluster = ~ zip + year_month
)

summary(mA); summary(mB)
```

Binomial model:
GLM estimation, family = binomial, Dep. Var.: is_le6
Observations: 2,588,544
Fixed-effects: zip: 672,  year_month: 126,  dow: 7
Standard-errors: Clustered (zip & year_month) 
                 Estimate Std. Error   z value   Pr(>|z|)    
effect::Cleanup  0.007657   0.027531  0.278111 7.8093e-01    
effect::Threat  -0.004921   0.030158 -0.163164 8.7039e-01    
effect::Impact   0.194508   0.047288  4.113296 3.9005e-05 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
Log-Likelihood: -406,796.1   Adj. Pseudo R2: 0.150202
           BIC:  825,494.1     Squared Cor.: 0.052679

Interpretation
Cleanup and Threat days: No meaningful evidence that these periods change the probability of low ER visits.
Impact days (landfall): A positive and significant log-odds means ER visit counts are more likely to fall in the ≤6 range. In other words, ER utilization tends to drop during direct storm impact

This makes sense: Many people stay home. Clinics/hospitals reduce service hours. Travel becomes difficult.
This model focuses on the lower tail of demand and here, Impact days show a clear reduction

Poisson model
Poisson estimation, Dep. Var.: er_visits
Observations: 2,722,224
Fixed-effects: zip: 912,  year_month: 126,  dow: 7
Standard-errors: Clustered (zip & year_month) 
                 Estimate Std. Error  z value  Pr(>|z|)    
effect::Cleanup  0.023700   0.008545  2.77347 0.0055461 ** 
effect::Threat   0.018231   0.008404  2.16917 0.0300695 *  
effect::Impact  -0.085411   0.026034 -3.28079 0.0010352 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
Log-Likelihood: -8,500,999.7   Adj. Pseudo R2: 0.604913
           BIC: 17,017,498.0     Squared Cor.: 0.880416


Interpretation
Cleanup days: Shows a small but significant +2–3% ER increase. This makes intuitive sense as delayed care resumes injuries from debris removal reconnecting the healthcare system after outages. Threat days (pre-impact conditions): ER visits increase (+1–2%). People may seek care before the storm hits (filling prescriptions, addressing minor issues early)
Impact days: ER visits drop sharply (≈ –8%). This lines up perfectly with the binomial model’s finding: ER demand suppresses during the storm’s immediate impact

Jointly the models show storm Impact periods suppress ER utilization — likely due to mobility constraints, closures, and shelter-in-place behavior. Pre-storm Threat periods elevate ER visits, consistent with precautionary care. Cleanup periods also show elevated ER visits as expected with resuming delayed care, debris removal injuries and reconnecting with heathcare system after outages. This is exactly the kind of pattern you’d expect in a population with both anticipatory behavior and acute disruption around major storms

During hurricane season in Florida, patients ramp up ER usage slightly the week before storms (Threat). On the direct storm impact day, ER demand collapses (Impact). After storms, ER visits rebound and exceed normal levels (Cleanup). This is a classic displacement–disruption pattern. Pre-storm increase → mid-storm suppression → post-storm rebound. The fixed effects ensure this is not confounded by ZIP characteristics, seasonal cycles, or weekly patterns

* Compare results with a subset of the stronger storms 55+

GLM estimation, family = binomial, Dep. Var.: is_le6
Observations: 2,576,988
Fixed-effects: zip: 669,  year_month: 126,  dow: 7
Standard-errors: Clustered (zip & year_month) 
                 Estimate Std. Error   z value Pr(>|z|)    
effect::Cleanup -0.111066   0.075692 -1.467345 0.142282    
effect::Threat  -0.032148   0.056498 -0.569022 0.569341    
effect::Impact   0.261600   0.136040  1.922963 0.054485 .  
Log-Likelihood: -392,643.5   Adj. Pseudo R2: 0.153986
           BIC:  797,140.9     Squared Cor.: 0.053344
           
Poisson estimation, Dep. Var.: er_visits
Observations: 2,648,475
Fixed-effects: zip: 895,  year_month: 126,  dow: 7
Standard-errors: Clustered (zip & year_month) 
                 Estimate Std. Error   z value  Pr(>|z|)    
effect::Cleanup  0.061417   0.023028  2.667050 0.0076520 ** 
effect::Threat   0.014805   0.019471  0.760378 0.4470288    
effect::Impact  -0.235362   0.077399 -3.040881 0.0023589 ** 
Log-Likelihood: -8,280,629.7   Adj. Pseudo R2: 0.605412
           BIC: 16,576,477.7     Squared Cor.: 0.880557

## Function: two‑part g‑computation with clustered bootstrap + figure

```{r}
# --- Packages ---
library(data.table)
library(dplyr)
library(lubridate)
library(fixest)
library(ggplot2)

two_part_gcomp_bootstrap <- function(
  df,
  effect_levels = c("None", "Cleanup", "Threat", "Impact"),
  season_months = 5:11,          # May–Nov
  mean_le6      = 6,             # expected visits for the <=6 group (top-coded as 6)
  ref_level     = "None",
  n_boot        = 500L,          # bootstrap replicates
  cluster       = c("zip", "zip_year_month"),
  seed          = 123,
  verbose       = TRUE
) {
  cluster <- match.arg(cluster)

  # ---- 0) Prep data ----
  dat <- as.data.table(df)

  # Ensure types
  dat[, date := as.IDate(date)]
  dat[, zip  := as.character(zip)]

  # Recreate fixed effects (safe even if present already)
  dat[, dow        := lubridate::wday(date, label = TRUE, abbr = TRUE)]
  dat[, year_month := as.IDate(floor_date(date, unit = "month"))]

  # Hurricane season filter
  dat <- dat[month(date) %in% season_months]

  # Outcome helpers
  dat[, is_le6 := as.integer(er_visits == 6L)]
  dat[, effect := factor(effect, levels = effect_levels)]

  # Housekeeping id
  dat[, row_id := .I]

  # ---- 1) Fit two-part models on full sample ----
  # (Use i(effect, ref="None") + FE: zip + year_month + dow)
  fA <- as.formula(paste0(
    "is_le6 ~ i(effect, ref='", ref_level, "') | zip + year_month + dow"
  ))
  fB <- as.formula(paste0(
    "er_visits ~ i(effect, ref='", ref_level, "') | zip + year_month + dow"
  ))

  # Part A: logit for being <=6
  mA <- feglm(
    fA, data = dat,
    family = binomial("logit"),
    cluster = ~ zip + year_month
  )

  # Part B: Poisson on Y>6 (fit on the subset, but we'll predict on full)
  d_gt6 <- dat[er_visits > 6L]
  mB <- fepois(
    fB, data = d_gt6,
    cluster = ~ zip + year_month
  )

  # ---- 2) Counterfactual predictions on full data (point estimates) ----
  counterfactual_mean <- function(effect_level, modelA, modelB) {
    newdat <- copy(dat)
    newdat[, effect := factor(effect_level, levels = effect_levels)]
    # Predictions
    newdat[, prob_le6 := predict(modelA, type = "response", newdata = newdat)]
    newdat[, mu_gt6   := predict(modelB, type = "response", newdata = newdat)]
    # Combined expected ER visits
    newdat[, EY := prob_le6 * mean_le6 + (1 - prob_le6) * mu_gt6]
    mean(newdat$EY, na.rm = TRUE)
  }

  cf_levels <- effect_levels
  set.seed(seed)
  cf_point <- sapply(cf_levels, counterfactual_mean, modelA = mA, modelB = mB)
  names(cf_point) <- cf_levels

  # Differences vs reference ("None")
  diff_point <- cf_point - cf_point[ref_level]

  # ---- 3) Clustered bootstrap ----
  # Define bootstrap clusters
  if (cluster == "zip") {
    dat[, boot_cluster := zip]
  } else {
    dat[, boot_cluster := interaction(zip, year_month, drop = TRUE)]
  }

  unique_clusters <- unique(dat$boot_cluster)
  n_clusters <- length(unique_clusters)

  if (verbose) {
    message(sprintf("Bootstrap clustering on '%s' with %d clusters.", cluster, n_clusters))
  }

  # Function: one bootstrap replicate
  do_one_boot <- function(b) {
    # Resample clusters with replacement
    set.seed(seed + b)
    clust_draw <- sample(unique_clusters, size = n_clusters, replace = TRUE)
    dat_b <- dat[boot_cluster %in% clust_draw]
    # Keep all rows from each selected cluster (with multiplicity)
    # Implement multiplicity by binding copies when clusters repeat
    if (any(duplicated(clust_draw))) {
      # Count multiplicities
      mult <- as.data.table(table(clust_draw))
      setnames(mult, c("clust_draw", "n"))
      dat_b <- rbindlist(lapply(seq_len(nrow(mult)), function(i) {
        c_id <- as.character(mult$clust_draw[i])
        k    <- as.integer(mult$n[i])
        rbindlist(rep(list(dat[boot_cluster == c_id]), k))
      }), use.names = TRUE)
    }

    # Refit models
    # Guard in case some effect levels drop: keep factor levels fixed
    dat_b[, effect := factor(effect, levels = effect_levels)]
    d_gt6_b <- dat_b[er_visits > 6L]

    # If a replicate has no >6 rows (very unlikely), skip with NA
    if (nrow(d_gt6_b) == 0) {
      return(list(cf = rep(NA_real_, length(cf_levels))))
    }

    mA_b <- tryCatch(
      feglm(fA, data = dat_b, family = binomial("logit"), cluster = ~ zip + year_month),
      error = function(e) NULL
    )
    mB_b <- tryCatch(
      fepois(fB, data = d_gt6_b, cluster = ~ zip + year_month),
      error = function(e) NULL
    )

    if (is.null(mA_b) || is.null(mB_b)) {
      return(list(cf = rep(NA_real_, length(cf_levels))))
    }

    # Counterfactual means for this replicate
    cf_b <- sapply(cf_levels, counterfactual_mean, modelA = mA_b, modelB = mB_b)
    list(cf = cf_b)
  }

  if (verbose) message(sprintf("Running %d bootstrap replicates…", n_boot))
  boot_list <- lapply(seq_len(n_boot), do_one_boot)

  # Stack bootstrap results
  boot_cf <- rbindlist(lapply(boot_list, function(x) as.list(x$cf)), fill = TRUE)
  setnames(boot_cf, cf_levels)

  # Drop replicates that failed (all NA)
  boot_cf <- boot_cf[rowSums(is.na(boot_cf)) < length(cf_levels)]

  # Bootstrap diffs vs None
  boot_diff <- boot_cf[, lapply(.SD, function(col) col - boot_cf[[ref_level]])]

  # ---- 4) Build tidy outputs with 95% CIs (percentile) ----
  pct <- function(x, p) as.numeric(quantile(x, probs = p, na.rm = TRUE))
  ci_tbl_means <- data.table(
    effect   = cf_levels,
    estimate = as.numeric(cf_point),
    ci_lo    = sapply(cf_levels, function(k) pct(boot_cf[[k]], 0.025)),
    ci_hi    = sapply(cf_levels, function(k) pct(boot_cf[[k]], 0.975))
  )

  diff_levels <- setdiff(cf_levels, ref_level)
  ci_tbl_diffs <- data.table(
    effect   = diff_levels,
    estimate = as.numeric(diff_point[diff_levels]),
    ci_lo    = sapply(diff_levels, function(k) pct(boot_diff[[k]], 0.025)),
    ci_hi    = sapply(diff_levels, function(k) pct(boot_diff[[k]], 0.975))
  )

  # ---- 5) Publication-ready figure: differences vs "None" ----
plot_df <- ci_tbl_diffs %>%
  dplyr::mutate(effect = factor(effect, levels = c("Threat", "Impact", "Cleanup"))) %>%
  dplyr::arrange(effect)

fig <- ggplot(plot_df, aes(x = effect, y = estimate)) +
  geom_hline(yintercept = 0, linewidth = 0.6, color = "grey50") +
  geom_pointrange(aes(ymin = ci_lo, ymax = ci_hi),
                  linewidth = 0.6, fatten = 2, color = "#2C3E50") +
  labs(
    x = NULL,
    y = "Difference in expected ER visits vs None",
    title = "Combined Effect of Storm Phases on Expected ER Visits",
    subtitle = sprintf(
      "Two‑part model combined via g‑computation; %d bootstrap replicates (clustered by %s)",
      n_boot, ifelse(cluster == "zip","ZIP","ZIP×Year‑Month")
    ),
    caption = "Points: bootstrap means; bars: 95% percentile CIs"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title.position = "plot",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 11)
  )

  # Return everything
  list(
    models = list(mA = mA, mB = mB),
    point_estimates = list(
      expected_visits = ci_tbl_means,
      diff_vs_none    = ci_tbl_diffs
    ),
    bootstrap = list(
      cf = boot_cf,
      diff = boot_diff
    ),
    figure = fig
  )
}
```

## Combine models and create figure

```{r}
# Build the modeling dataset by joining:
dat_model <- merge(
  effect_calendar[, .(zip, date, effect)],
  ER_visits_clean[, .(zip, date, er_visits)],
  by = c("zip","date"),
  all.x = TRUE
)
dat_model[is.na(er_visits), er_visits := 0L]

# Run (ZIP-clustered bootstrap)
res <- two_part_gcomp_bootstrap(
  df        = dat_model,
  n_boot    = 10,          # increase to 500–1000 for publication
  cluster   = "zip",
  seed      = 123
)

# View the unified, combined effect (expected ER visits) with CIs
res$point_estimates$expected_visits

# View the key plot (differences vs "None")
print(res$figure)

# Save outputs
data.table::fwrite(res$point_estimates$expected_visits,
                   here::here("output","expected_visits_by_effect.csv"))
data.table::fwrite(res$point_estimates$diff_vs_none,
                   here::here("output","diff_vs_none_by_effect.csv"))
ggplot2::ggsave(filename = here::here("figs","combined_effect_diffs_vs_none.png"),
                plot = res$figure, width = 7, height = 4.5, dpi = 300)
```


