---
title: "Refit of Florida birth-outcome models using corrected tropical cyclone exposure"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  corrected_data: "data/outputs/All_Births_Corrected_Storm_Exposure-trimester-eligible-v2.rds"
  event_data: "data/derived/birth-storm-events-time-specific-trimester-eligible-v2.csv"
  birth_interval_data: "data/all_births_hurricane.csv"
  exposure_window_days: 1
  output_dir: "data/outputs/Models-trimester-eligible-v2-window1"
  figure_dir: "figs/model-comparison-trimester-eligible-v2-window1"
  imputations: 5
  mice_iterations: 5
  imputation_cores: 5
  polyreg_max_weights: 10000
  imputation_training_rows: 50000
  seed: 20260810
  fit_55kt_sensitivity: false
  fit_complete_case_sensitivity: true
  save_imputation_object: false
---

## Purpose and reconstructed manuscript model

This workflow refits the birth-outcome models described in `Final
Manuscript.pdf` using the corrected storm-time-specific exposure dataset. It
does not use the superseded threat, impact, or cleanup variables.

The manuscript describes:

1. Binary logistic regression with adjusted odds ratios (aOR) and 95%
   confidence intervals.
2. Outcomes: preterm birth (PTB), very preterm birth (vPTB), term low
   birthweight (tLBW), low birthweight (LBW), very low birthweight (vLBW),
   gestational diabetes (GDM), and hypertensive disorders of pregnancy (HDP).
3. Separate exposure models by trimester. PTB, vPTB, LBW, and vLBW were not
   modeled for the third trimester; tLBW was modeled in all three trimesters;
   GDM and HDP used first-trimester exposure only.
4. Adjustment for maternal and paternal race, ethnicity, and education;
   maternal age; categorical pre-pregnancy BMI; previous preterm infant; WIC
   status; and principal source of payment.
5. Multiple imputation by chained equations (MICE) for missing covariates.
   The manuscript reports five iterations but does not report the number of
   imputed datasets or variable-specific imputation methods.

For reproducibility, this refit uses `r params$imputations` imputed datasets and
`r params$mice_iterations` MICE iterations. In response to reviewer concerns
about temporal and spatial confounding, every corrected primary model includes
conception year-month and residence-county fixed effects, with standard errors
clustered by county. Eligibility is trimester-specific, so an early delivery
remains in every earlier gestational period it reached. The 34 kt ±7-day
exposure used previously is rebuilt here with the manuscript's ±1-day window.
The corrected residence-specific wind field and local impact date are otherwise
unchanged.

Alcohol was listed among measured covariates in the manuscript's Covariates
section but was not included in the Statistical Analysis adjustment list. It
is therefore not included in the primary refit.

## Setup

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("data.table", "fixest", "ggplot2", "mice", "scales")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0L) {
  stop(
    "Install required packages before running: ",
    paste(missing_packages, collapse = ", "),
    ". For example: install.packages(c(",
    paste(sprintf("'%s'", missing_packages), collapse = ", "), "))",
    call. = FALSE
  )
}

library(data.table)
library(fixest)
library(ggplot2)

set.seed(as.integer(params$seed))
window_days <- as.integer(params$exposure_window_days)
analysis_thresholds <- if (isTRUE(params$fit_55kt_sensitivity)) c(34L, 55L) else 34L
stopifnot(length(window_days) == 1L, !is.na(window_days), window_days >= 0L)
model_output_dir <- as.character(params$output_dir)
model_figure_dir <- as.character(params$figure_dir)
stopifnot(
  length(model_output_dir) == 1L, nzchar(model_output_dir),
  length(model_figure_dir) == 1L, nzchar(model_figure_dir)
)
dir.create(model_output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(model_figure_dir, recursive = TRUE, showWarnings = FALSE)

progress_file <- file.path(model_output_dir, "model-refit-progress.log")
if (file.exists(progress_file)) file.remove(progress_file)

log_progress <- function(stage, detail) {
  line <- sprintf(
    "%s | %-22s | %s",
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), stage, detail
  )
  message(line)
  write(line, progress_file, append = TRUE)
  invisible(line)
}

format_duration <- function(seconds) {
  seconds <- max(0, round(seconds))
  sprintf("%02d:%02d:%02d", seconds %/% 3600,
          (seconds %% 3600) %/% 60, seconds %% 60)
}

log_progress("setup", "model refit started")
```

## Read only model-required columns

An RDS must be read in full, but unnecessary columns are removed immediately
to reduce working memory.

```{r read-data}
stopifnot(file.exists(params$corrected_data))
births <- readRDS(params$corrected_data)
setDT(births)

source_columns <- c(
  "ID", "BIRTH_WEIGHT_GRAMS", "GESTATION_WEEKS", "PLURALITY_CODE",
  "MR_DIAB", "MR_DIAB_GEST", "MR_HYPERT_CHRONIC",
  "MR_HYPERT_ECLAMPSIA", "MR_HYPERT_PREG",
  "Mother_CalculatedRace", "Mother_CalculatedHisp", "MOTHER_EDCODE",
  "Father_CalculatedRace", "Father_CalculatedHisp", "FATHER_EDCODE",
  "MOTHER_AGE", "PrePregnancy_BMI", "MR_PREV_PRETERM",
  "MOTHER_WIC_YESNO", "PRINCIPAL_SRCPAY_CODE",
  "birth_year", "conception_year", "conception_month",
  "conception_season", "county_geoid", "county_name",
  "county_demo_year_lag1", "county_total_pop_lag1",
  "county_median_age_lag1", "county_poverty_rate_lag1",
  "county_median_hh_income_lag1", "county_median_home_value_lag1",
  "county_pct_white_lag1", "county_pct_black_lag1",
  "county_pct_asian_lag1", "county_pct_hispanic_lag1",
  "exposure_eligible", "exposure_eligible_T1",
  "exposure_eligible_T2", "exposure_eligible_T3"
)
exposure_columns <- as.vector(outer(
  paste0("exposed_window_T", 1:3), analysis_thresholds, paste, sep = "_"
))
required_columns <- c(source_columns, exposure_columns)
missing_columns <- setdiff(required_columns, names(births))
if (length(missing_columns) > 0L) {
  stop(
    "Corrected dataset is missing model-required columns: ",
    paste(missing_columns, collapse = ", "),
    ". Rerun HO-storms-exposure.Rmd after its latest update.",
    call. = FALSE
  )
}

births <- births[, ..required_columns]
gc()

# Rebuild only the gestational window assignment. The expensive spatial
# intersection and residence-specific local impact time are reused from the
# corrected event file. This makes the ±1-day comparison differ from the
# corrected conventional model only in window width, while retaining the
# corrected wind-field exposure and trimester-specific eligibility.
stopifnot(file.exists(params$event_data), file.exists(params$birth_interval_data))
interval_dates <- fread(
  params$birth_interval_data,
  select = c(
    "ID", "DATE_OF_BIRTH", "trimester1_s_date", "trimester1_e_date",
    "trimester2_s_date", "trimester2_e_date", "trimester3_s_date"
  )
)
interval_date_columns <- setdiff(names(interval_dates), "ID")
interval_dates[, (interval_date_columns) := lapply(.SD, as.IDate),
               .SDcols = interval_date_columns]
interval_dates[, `:=`(
  T1_start = trimester1_s_date,
  T1_end = pmin(trimester1_e_date, DATE_OF_BIRTH, na.rm = FALSE),
  T2_start = trimester2_s_date,
  T2_end = pmin(trimester2_e_date, DATE_OF_BIRTH, na.rm = FALSE),
  T3_start = trimester3_s_date,
  T3_end = DATE_OF_BIRTH
)]
interval_dates <- interval_dates[, .(
  ID, T1_start, T1_end, T2_start, T2_end, T3_start, T3_end
)]

event_impacts <- fread(
  params$event_data,
  select = c("ID", "SID", "threshold_kt", "local_impact_date")
)[threshold_kt %in% analysis_thresholds]
event_impacts[, `:=`(
  local_impact_date = as.IDate(local_impact_date),
  window_start_rebuilt = as.IDate(local_impact_date) - window_days,
  window_end_rebuilt = as.IDate(local_impact_date) + window_days
)]
event_impacts <- merge(
  event_impacts, interval_dates, by = "ID", all.x = TRUE, sort = FALSE
)

exposure_hits <- rbindlist(lapply(paste0("T", 1:3), function(trimester) {
  period_start <- paste0(trimester, "_start")
  period_end <- paste0(trimester, "_end")
  event_impacts[
    !is.na(get(period_start)) & !is.na(get(period_end)) &
      window_start_rebuilt <= get(period_end) &
      window_end_rebuilt >= get(period_start),
    .(ID, trimester, threshold_kt)
  ]
}))
exposure_hits <- unique(exposure_hits)

for (trimester_value in paste0("T", 1:3)) {
  eligibility <- paste0("exposure_eligible_", trimester_value)
  for (threshold in analysis_thresholds) {
    exposure <- paste0("exposed_window_", trimester_value, "_", threshold)
    births[, (exposure) := fifelse(get(eligibility) %in% TRUE, 0L, NA_integer_)]
    exposed_ids <- exposure_hits[
      trimester == trimester_value & threshold_kt == threshold, ID
    ]
    exposed_rows <- match(exposed_ids, births$ID)
    exposed_rows <- exposed_rows[
      !is.na(exposed_rows) & births[[eligibility]][exposed_rows] %in% TRUE
    ]
    set(births, exposed_rows, exposure, 1L)
  }
}
window_exposure_qa <- rbindlist(lapply(paste0("T", 1:3), function(trimester) {
  rbindlist(lapply(analysis_thresholds, function(threshold) {
    exposure <- paste0("exposed_window_", trimester, "_", threshold)
    data.table(
      window_days = window_days, trimester = trimester,
      threshold_kt = threshold,
      eligible = sum(!is.na(births[[exposure]])),
      exposed = sum(births[[exposure]] == 1L, na.rm = TRUE),
      prevalence = mean(births[[exposure]], na.rm = TRUE)
    )
  }))
}))
fwrite(window_exposure_qa,
       file.path(model_output_dir, "window1-exposure-rebuild-qa.csv"))
rm(interval_dates, event_impacts, exposure_hits)
gc()
log_progress(
  "read data",
  sprintf("loaded %s births and retained %s columns",
          format(nrow(births), big.mark = ","), ncol(births))
)
```

## Recode outcomes, eligibility, and covariates

```{r recode}
normalize_missing_code <- function(x, missing = c("", "NULL", "NA", "N/A", "U", "UNK", "UNKNOWN")) {
  value <- trimws(toupper(as.character(x)))
  value[value %in% missing] <- NA_character_
  value
}

binary_yes_no <- function(x) {
  value <- normalize_missing_code(x)
  fcase(
    value %in% c("Y", "YES", "1", "TRUE"), 1L,
    value %in% c("N", "NO", "0", "FALSE"), 0L,
    default = NA_integer_
  )
}

factor_missing_codes <- function(x, additional_missing = character()) {
  value <- normalize_missing_code(
    x, missing = c("", "NULL", "NA", "N/A", "U", "UNK", "UNKNOWN",
                   additional_missing)
  )
  factor(value)
}

collapse_race <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 1L, "White",
    value == 2L, "Black",
    !is.na(value) & value != 99L, "Other",
    default = NA_character_
  ), levels = c("White", "Black", "Other"))
}

collapse_ethnicity <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 0L, "Non-Hispanic",
    value %in% 1:98, "Hispanic",
    default = NA_character_
  ), levels = c("Non-Hispanic", "Hispanic"))
}

collapse_education <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value %in% 1:2, "Less than high school",
    value == 3L, "High school or GED",
    value %in% 4:5, "Some college or associate degree",
    value %in% 6:8, "Bachelor degree or higher",
    default = NA_character_
  ), levels = c(
    "Less than high school", "High school or GED",
    "Some college or associate degree", "Bachelor degree or higher"
  ))
}

collapse_payment <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 1L, "Medicaid",
    value == 2L, "Private insurance",
    value == 3L, "Self-pay",
    !is.na(value) & value != 9L, "Other",
    default = NA_character_
  ), levels = c("Medicaid", "Private insurance", "Self-pay", "Other"))
}

births[, `:=`(
  gest_weeks = suppressWarnings(as.numeric(GESTATION_WEEKS)),
  birthweight_g = suppressWarnings(as.numeric(BIRTH_WEIGHT_GRAMS)),
  singleton = suppressWarnings(as.integer(PLURALITY_CODE)) == 1L,
  maternal_race = collapse_race(Mother_CalculatedRace),
  maternal_ethnicity = collapse_ethnicity(Mother_CalculatedHisp),
  maternal_education = collapse_education(MOTHER_EDCODE),
  paternal_race = collapse_race(Father_CalculatedRace),
  paternal_ethnicity = collapse_ethnicity(Father_CalculatedHisp),
  paternal_education = collapse_education(FATHER_EDCODE),
  maternal_age = suppressWarnings(as.numeric(MOTHER_AGE)),
  previous_preterm = factor(binary_yes_no(MR_PREV_PRETERM), levels = 0:1),
  wic = factor(binary_yes_no(MOTHER_WIC_YESNO), levels = 0:1),
  payment = collapse_payment(PRINCIPAL_SRCPAY_CODE),
  preexisting_diabetes = binary_yes_no(MR_DIAB),
  gestational_diabetes_raw = binary_yes_no(MR_DIAB_GEST),
  chronic_hypertension = binary_yes_no(MR_HYPERT_CHRONIC),
  gestational_hypertension = binary_yes_no(MR_HYPERT_PREG),
  eclampsia = binary_yes_no(MR_HYPERT_ECLAMPSIA)
)]

bmi_numeric <- suppressWarnings(as.numeric(normalize_missing_code(
  births$PrePregnancy_BMI
)))
births[, bmi_category := cut(
  bmi_numeric,
  breaks = c(-Inf, 18.5, 25, 30, Inf),
  right = FALSE,
  labels = c("Underweight", "Normal", "Overweight", "Obesity")
)]

births[, hdp_raw := fcase(
  gestational_hypertension == 1L | eclampsia == 1L, 1L,
  gestational_hypertension == 0L & eclampsia == 0L, 0L,
  default = NA_integer_
)]

births[, `:=`(
  PTB = as.integer(gest_weeks < 37),
  vPTB = as.integer(gest_weeks < 32),
  LBW = as.integer(birthweight_g < 2500),
  vLBW = as.integer(birthweight_g < 1500),
  tLBW = as.integer(gest_weeks >= 37 & birthweight_g < 2500),
  GDM = gestational_diabetes_raw,
  HDP = hdp_raw,
  birth_year = suppressWarnings(as.integer(birth_year)),
  conception_year = suppressWarnings(as.integer(conception_year)),
  conception_month = suppressWarnings(as.integer(conception_month)),
  conception_year_month = factor(fifelse(
    !is.na(conception_year) & conception_month %in% 1:12,
    sprintf("%04d-%02d", conception_year, conception_month),
    NA_character_
  )),
  conception_season = factor(
    conception_season, levels = c("DJF", "MAM", "JJA", "SON")
  ),
  county_geoid = factor(county_geoid)
)]

# Union cohort for the conventional trimester-specific models. Each fitted
# model applies its own exposure_eligible_T* flag; no earlier-trimester model
# requires the pregnancy to have reached a later trimester.
model_births <- births[
  exposure_eligible %in% TRUE & singleton %in% TRUE &
    !is.na(gest_weeks) & !is.na(birthweight_g) &
    !is.na(conception_year_month) & !is.na(county_geoid)
]
model_births[, complication_eligible :=
               !is.na(GDM) & !is.na(HDP) &
               preexisting_diabetes == 0L & chronic_hypertension == 0L]

covariates <- c(
  "maternal_race", "maternal_ethnicity", "maternal_education",
  "paternal_race", "paternal_ethnicity", "paternal_education",
  "maternal_age", "bmi_category", "previous_preterm", "wic", "payment"
)

sample_qa <- rbindlist(lapply(c("T1", "T2", "T3"), function(trimester) {
  eligibility <- paste0("exposure_eligible_", trimester)
  analysis <- model_births[get(eligibility) %in% TRUE]
  analysis[, .(
    trimester = trimester,
    births = .N,
    counties = uniqueN(county_geoid),
    term_births = sum(gest_weeks >= 37),
    complication_eligible = sum(complication_eligible %in% TRUE),
    PTB_events = sum(PTB),
    vPTB_events = sum(vPTB),
    LBW_events = sum(LBW),
    vLBW_events = sum(vLBW),
    term_LBW_events = sum(tLBW[gest_weeks >= 37]),
    GDM_events = sum(GDM, na.rm = TRUE),
    HDP_events = sum(HDP, na.rm = TRUE),
    PTB_pct = 100 * mean(PTB),
    vPTB_pct = 100 * mean(vPTB),
    LBW_pct = 100 * mean(LBW),
    vLBW_pct = 100 * mean(vLBW),
    term_LBW_pct = 100 * mean(tLBW[gest_weeks >= 37]),
    GDM_pct = 100 * mean(GDM, na.rm = TRUE),
    HDP_pct = 100 * mean(HDP, na.rm = TRUE)
  )]
}))
print(sample_qa)

fwrite(sample_qa, file.path(model_output_dir, "corrected_model_sample_qa.csv"))
log_progress("recode", sprintf("model base cohort contains %s births",
                                format(nrow(model_births), big.mark = ",")))
```

## Corrected exposure prevalence

```{r prevalence}
prevalence_columns <- exposure_columns
corrected_prevalence <- rbindlist(lapply(prevalence_columns, function(variable) {
  data.table(
    exposure = variable,
    n_eligible = sum(!is.na(model_births[[variable]])),
    n_exposed = sum(model_births[[variable]] == 1L, na.rm = TRUE),
    prevalence = mean(model_births[[variable]], na.rm = TRUE)
  )
}))
corrected_prevalence[, manuscript_any_exposure_prevalence := 0.808]
corrected_prevalence[, prevalence_difference_percentage_points :=
                       100 * (prevalence - manuscript_any_exposure_prevalence)]
print(corrected_prevalence)
fwrite(corrected_prevalence,
       file.path(model_output_dir, "corrected_exposure_prevalence.csv"))
```

## Multiple imputation

The imputation model includes all analysis covariates, complete birthweight and
gestational-age measures, corrected exposures, county, conception year, and
conception month. The exact conception year-month fixed effect is applied in
the outcome models; using its 276-level interaction inside every chained
imputation regression adds substantial computation without changing the final
fixed-effect specification.
Only covariates are imputed; outcomes, exposures, and fixed effects are never
imputed. To make MICE feasible for more than four million births, its regression
equations are estimated on a reproducible random training sample of
`r format(params$imputation_training_rows, big.mark = ",")` records. MICE's
`ignore` mechanism then uses those fitted equations to impute all records. The
health-outcome models continue to use the complete eligible cohort.

```{r imputation}
imputation_predictors <- c(
  "birthweight_g", "gest_weeks", exposure_columns,
  "exposure_eligible_T1", "exposure_eligible_T2", "exposure_eligible_T3",
  "conception_year", "conception_month", "county_geoid"
)
imputation_columns <- unique(c(covariates, imputation_predictors))
imputation_source <- copy(model_births[, ..imputation_columns])
# Exposure is structurally missing outside its eligible trimester. For the
# imputation predictor matrix only, encode those cells as zero and retain the
# corresponding eligibility indicators. The outcome models continue to use the
# original NA exposure values and trimester-specific eligibility flags.
for (column in exposure_columns) {
  set(imputation_source, which(is.na(imputation_source[[column]])), column, 0L)
}
imputation_data <- as.data.frame(imputation_source)
rm(imputation_source)

training_rows <- min(
  as.integer(params$imputation_training_rows), nrow(imputation_data)
)
if (is.na(training_rows) || training_rows < 1000L) {
  stop("`imputation_training_rows` must be at least 1,000.", call. = FALSE)
}
set.seed(as.integer(params$seed) + 1L)
training_index <- sample.int(nrow(imputation_data), training_rows)
imputation_ignore <- rep(TRUE, nrow(imputation_data))
imputation_ignore[training_index] <- FALSE

method <- mice::make.method(imputation_data)
method[] <- ""
method["maternal_age"] <- "pmm"
binary_covariates <- c(
  "maternal_ethnicity", "paternal_ethnicity", "previous_preterm", "wic"
)
method[binary_covariates] <- "logreg"
multicategory_covariates <- setdiff(covariates,
                                    c("maternal_age", binary_covariates))
method[multicategory_covariates] <- "polyreg"

predictor_matrix <- mice::make.predictorMatrix(imputation_data)
predictor_matrix[,] <- 0L
predictor_matrix[covariates, setdiff(imputation_columns, covariates)] <- 1L
predictor_matrix[covariates, covariates] <- 1L
diag(predictor_matrix) <- 0L

log_progress(
  "imputation",
  sprintf("starting MICE: m=%s, maxit=%s, training rows=%s, imputed rows=%s",
          params$imputations, params$mice_iterations,
          format(training_rows, big.mark = ","),
          format(nrow(imputation_data), big.mark = ","))
)
imputation_started <- Sys.time()
imputation_cores <- min(
  as.integer(params$imputation_cores),
  as.integer(params$imputations)
)
stopifnot(imputation_cores >= 1L)
imputation_list <- parallel::mclapply(
  seq_len(as.integer(params$imputations)),
  mc.cores = imputation_cores,
  mc.preschedule = TRUE,
  mc.set.seed = TRUE,
  FUN = function(imputation_number) {
    mice::mice(
      imputation_data,
      m = 1L,
      maxit = as.integer(params$mice_iterations),
      method = method,
      predictorMatrix = predictor_matrix,
      ignore = imputation_ignore,
      seed = as.integer(params$seed) + imputation_number,
      printFlag = TRUE,
      nnet.MaxNWts = as.integer(params$polyreg_max_weights),
      nnet.trace = FALSE
    )
  }
)
stopifnot(
  length(imputation_list) == as.integer(params$imputations),
  all(vapply(imputation_list, inherits, logical(1), what = "mids"))
)
imputed <- Reduce(mice::ibind, imputation_list)
rm(imputation_list)
log_progress(
  "imputation",
  paste("completed in", format_duration(as.numeric(difftime(
    Sys.time(), imputation_started, units = "secs"
  ))))
)

if (isTRUE(params$save_imputation_object)) {
  saveRDS(imputed, file.path(model_output_dir, "corrected_birth_mice.rds"),
          compress = "xz")
}
```

## Model plan

```{r model-plan}
model_plan <- rbindlist(list(
  CJ(outcome = c("PTB", "vPTB", "LBW", "vLBW"), trimester = c("T1", "T2")),
  data.table(outcome = "tLBW", trimester = c("T1", "T2", "T3")),
  data.table(outcome = c("GDM", "HDP"), trimester = "T1")
))
model_plan[, sample := fcase(
  outcome == "tLBW", "term",
  outcome %in% c("GDM", "HDP"), "complication",
  default = "birth"
)]

thresholds <- analysis_thresholds
model_plan <- model_plan[, .(threshold_kt = thresholds),
                         by = .(outcome, trimester, sample)]
model_plan[, `:=`(
  exposure = paste0("exposed_window_", trimester, "_", threshold_kt),
  eligibility = paste0("exposure_eligible_", trimester)
)]
print(model_plan)
```

## Fit imputed logistic models

`fixest::feglm` fits the same binomial-logit likelihood as ordinary logistic
regression while absorbing conception year-month and county fixed effects
efficiently. County-clustered standard errors allow residual correlation among
births assigned to the same county.

```{r fit-models}
fit_one_model <- function(data, outcome, exposure, eligibility, sample) {
  keep <- rep(TRUE, nrow(data))
  if (sample == "term") keep <- data$gest_weeks >= 37
  if (sample == "complication") keep <- data$complication_eligible %in% TRUE

  analysis <- data[
    keep & data[[eligibility]] %in% TRUE &
      !is.na(data[[outcome]]) & !is.na(data[[exposure]])
  ]
  rhs <- paste(c(exposure, covariates), collapse = " + ")
  formula <- as.formula(paste0(
    outcome, " ~ ", rhs,
    " | conception_year_month + county_geoid"
  ))
  fit <- fixest::feglm(
    formula, data = analysis, family = "binomial",
    vcov = ~county_geoid
  )
  variance <- stats::vcov(fit)[exposure, exposure]

  data.table(
    beta = unname(coef(fit)[exposure]),
    within_variance = unname(variance),
    n = nobs(fit),
    events = sum(analysis[[outcome]] == 1L, na.rm = TRUE)
  )
}

model_results <- vector("list", nrow(model_plan) * params$imputations)
result_index <- 0L
model_started <- Sys.time()

for (imp in seq_len(params$imputations)) {
  log_progress("models", sprintf("starting imputation %s/%s",
                                  imp, params$imputations))
  completed_covariates <- as.data.table(mice::complete(imputed, imp))
  completed <- copy(model_births)
  completed[, (covariates) := completed_covariates[, covariates, with = FALSE]]

  for (j in seq_len(nrow(model_plan))) {
    result_index <- result_index + 1L
    spec <- model_plan[j]
    result <- fit_one_model(
      completed, spec$outcome, spec$exposure, spec$eligibility, spec$sample
    )
    model_results[[result_index]] <- cbind(
      data.table(imputation = imp), spec, result
    )
    log_progress(
      "models",
      sprintf(
        "imputation %s/%s, model %s/%s: %s %s %s kt",
        imp, params$imputations, j, nrow(model_plan),
        spec$outcome, spec$trimester, spec$threshold_kt
      )
    )
  }
  completed_results <- rbindlist(model_results[seq_len(result_index)])
  fwrite(
    completed_results,
    file.path(model_output_dir, "corrected_model_results_checkpoint.csv")
  )
  log_progress(
    "models",
    sprintf("checkpoint saved after imputation %s/%s",
            imp, params$imputations)
  )
  rm(completed, completed_covariates, completed_results)
  gc()
}

imputation_model_results <- rbindlist(model_results)
fwrite(imputation_model_results,
       file.path(model_output_dir, "corrected_model_results_by_imputation.csv"))
log_progress(
  "models",
  paste("all imputed models completed in", format_duration(as.numeric(difftime(
    Sys.time(), model_started, units = "secs"
  ))))
)
```

## Pool estimates using Rubin's rules

```{r pool}
pool_rubin <- function(beta, within_variance, n, events) {
  m <- length(beta)
  q_bar <- mean(beta)
  u_bar <- mean(within_variance)
  b <- if (m > 1L) var(beta) else 0
  total_variance <- u_bar + (1 + 1 / m) * b
  degrees_freedom <- if (b > 0) {
    (m - 1) * (1 + u_bar / ((1 + 1 / m) * b))^2
  } else {
    Inf
  }
  critical <- if (is.finite(degrees_freedom))
    qt(0.975, df = degrees_freedom) else qnorm(0.975)

  data.table(
    beta = q_bar,
    standard_error = sqrt(total_variance),
    degrees_freedom = degrees_freedom,
    OR = exp(q_bar),
    conf_low = exp(q_bar - critical * sqrt(total_variance)),
    conf_high = exp(q_bar + critical * sqrt(total_variance)),
    n = round(mean(n)),
    events = round(mean(events))
  )
}

pooled_results <- imputation_model_results[,
  pool_rubin(beta, within_variance, n, events),
  by = .(outcome, trimester, sample, threshold_kt, exposure)
]
pooled_results[, estimate := sprintf("%.3f (%.3f, %.3f)",
                                      OR, conf_low, conf_high)]
fwrite(pooled_results,
       file.path(model_output_dir, "corrected_model_estimates_pooled.csv"))
print(pooled_results)
```

## Complete-case sensitivity analysis

```{r complete-case}
complete_case_results <- data.table()
if (isTRUE(params$fit_complete_case_sensitivity)) {
  complete_case_results <- rbindlist(lapply(seq_len(nrow(model_plan)), function(j) {
    spec <- model_plan[j]
    complete <- complete.cases(model_births[, c(covariates, spec$exposure), with = FALSE])
    analysis <- model_births[complete]
    fit <- fit_one_model(
      analysis, spec$outcome, spec$exposure, spec$eligibility, spec$sample
    )
    se <- sqrt(fit$within_variance)
    cbind(spec, data.table(
      OR = exp(fit$beta),
      conf_low = exp(fit$beta - qnorm(0.975) * se),
      conf_high = exp(fit$beta + qnorm(0.975) * se),
      n = fit$n,
      events = fit$events
    ))
  }))
  fwrite(complete_case_results,
         file.path(model_output_dir, "corrected_model_estimates_complete_case.csv"))
}
```

## Comparison with estimates reported in the manuscript

The manuscript DOCX contains Tables 2-4. The comparison below transcribes every
adjusted `Any Storm Effect` estimate reported for the outcome/trimester models,
rather than relying only on estimates repeated in the Results prose.

```{r manuscript-comparison}
manuscript_results <- data.table(
  outcome = c(
    "PTB", "PTB", "vPTB", "vPTB", "tLBW", "tLBW", "tLBW",
    "LBW", "LBW", "vLBW", "vLBW", "GDM", "HDP"
  ),
  trimester = c(
    "T1", "T2", "T1", "T2", "T1", "T2", "T3",
    "T1", "T2", "T1", "T2", "T1", "T1"
  ),
  old_OR = c(
    1.012, 1.008, 0.985, 0.919, 1.023, 1.034, 0.966,
    1.004, 1.003, 0.985, 0.914, 1.054, 1.127
  ),
  old_low = c(
    1.004, 1.000, 0.966, 0.902, 1.006, 1.019, 0.950,
    0.994, 0.995, 0.965, 0.896, 1.041, 1.114
  ),
  old_high = c(
    1.020, 1.016, 1.005, 0.937, 1.040, 1.051, 0.982,
    1.013, 1.012, 1.005, 0.933, 1.068, 1.141
  )
)

comparison <- merge(
  pooled_results[threshold_kt == 34L],
  manuscript_results,
  by = c("outcome", "trimester"),
  all.x = TRUE
)
comparison[, `:=`(
  old_estimate = fifelse(
    !is.na(old_OR), sprintf("%.3f (%.3f, %.3f)", old_OR, old_low, old_high),
    NA_character_
  ),
  OR_absolute_difference = OR - old_OR,
  log_OR_difference = log(OR) - log(old_OR),
  percent_change_in_OR = 100 * (OR / old_OR - 1),
  old_direction = fcase(old_OR > 1, "above null", old_OR < 1, "below null",
                        default = NA_character_),
  corrected_direction = fcase(OR > 1, "above null", OR < 1, "below null",
                              default = "null"),
  direction_changed = !is.na(old_OR) & sign(log(OR)) != sign(log(old_OR)),
  corrected_excludes_null = conf_low > 1 | conf_high < 1,
  old_excludes_null = !is.na(old_OR) & (old_low > 1 | old_high < 1)
)]

fwrite(comparison,
       file.path(model_output_dir, "manuscript_vs_corrected_model_comparison.csv"))
print(comparison[, .(
  outcome, trimester, old_estimate, corrected = estimate,
  percent_change_in_OR, direction_changed,
  old_excludes_null, corrected_excludes_null
)])
```

## Comparison figure

```{r comparison-figure, fig.width=10, fig.height=7}
plot_data <- rbindlist(list(
  manuscript_results[, .(
    outcome, trimester, specification = "Manuscript (reported)",
    OR = old_OR, conf_low = old_low, conf_high = old_high
  )],
  pooled_results[threshold_kt == 34L &
                   paste(outcome, trimester) %in%
                   paste(manuscript_results$outcome, manuscript_results$trimester), .(
    outcome, trimester,
    specification = paste0(
      "Corrected 34 kt ±", window_days,
      " day + conception year-month/county FE"
    ),
    OR, conf_low, conf_high
  )]
))
plot_data[, label := paste(outcome, trimester)]

comparison_plot <- ggplot(
  plot_data,
  aes(x = OR, y = reorder(label, OR), color = specification)
) +
  geom_vline(xintercept = 1, linetype = 2, color = "grey50") +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high),
                 position = position_dodge(width = 0.5), height = 0.15) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_x_log10() +
  labs(
    x = "Adjusted odds ratio (log scale)", y = NULL, color = NULL,
    title = "Manuscript and corrected tropical-cyclone exposure estimates",
    subtitle = "All adjusted Any Storm Effect estimates from manuscript Tables 2–4"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
print(comparison_plot)

ggsave(
  file.path(model_figure_dir, "manuscript-vs-corrected-forest-plot.png"),
  comparison_plot, width = 10, height = 7, dpi = 300
)
```

## Automated summary

```{r summary}
reported_comparison <- comparison[!is.na(old_OR)]
summary_counts <- reported_comparison[, .(
  manuscript_estimates_compared = .N,
  direction_changes = sum(direction_changed),
  manuscript_intervals_excluding_null = sum(old_excludes_null),
  corrected_intervals_excluding_null = sum(corrected_excludes_null),
  median_absolute_percent_OR_change = median(abs(percent_change_in_OR))
)]
print(summary_counts)
fwrite(summary_counts, file.path(model_output_dir, "model_comparison_summary.csv"))

log_progress("complete", "model refit and comparison finished")
```

## Interpretation notes

Both columns use a ±1-day exposure window, but differences between manuscript
and corrected estimates still reflect several changes: correction of the
original space-time join, replacement of RMW track buffers with time-specific
34 kt wind radii, use of a residence-specific local impact date,
trimester-specific eligibility, and addition of conception year-month and
county fixed effects with county-clustered standard errors. The comparison
therefore isolates window width from the prior ±7-day corrected analysis, but
does not isolate the other exposure and model corrections from one another.

The manuscript's statement “five iterations” is insufficient to reproduce its
imputation exactly. This workflow records the number of imputed datasets,
iterations, seed, methods, and predictor matrix. If the original analysts can
provide their MICE specification or Tables 2-4, update those inputs before the
final manuscript revision.
