---
title: "Health Outcomes & Storms"
output: html_document
editor_options: 
  chunk_output_type: console
---

Task: Create a model that outlines areas of hurricane threats and impacts by date. Model output will be collated with health outcome data at the individual level within the areas for those dates

Get and import the IBTraCS data. Geometry type LINESTRING

https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/
https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/doc/IBTrACS_v04_Technical_Details.pdf

https://www.ncei.noaa.gov/sites/default/files/2021-07/IBTrACS_v04_column_documentation.pdf

Wind speed is in units of knots (nautical mile per hour)
```{r}
L <- "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/IBTrACS.NA.list.v04r00.lines.zip"
  
if(!"IBTrACS.NA.list.v04r00.lines.zip" %in% list.files(here::here("data"))) {
download.file(url = L,
              destfile = here::here("data",
                                    "IBTrACS.NA.list.v04r00.lines.zip"))
unzip(here::here("data", "IBTrACS.NA.list.v04r00.lines.zip"),
      exdir = here::here("data"))
}

Tracks.sf <- sf::st_read(dsn = here::here("data"), 
                         layer = "IBTrACS.NA.list.v04r00.lines") |>
  sf::st_transform(crs = 32616)
```

Keep only tropical cyclones having USA_WIND >= 64 occurring between 1981 and 2022
```{r}
Tracks2.sf <- Tracks.sf |>
  dplyr::filter(year >= 1981 & year <= 2022) |>
  dplyr::filter(USA_WIND >= 64) |> # change to 34 for tropical storms and hurricanes
  dplyr::select(SID, SEASON, year, month, day, hour, min,
                NAME, SUBBASIN, ISO_TIME,
                USA_WIND, USA_PRES, USA_RMW, USA_EYE, USA_ROCI)
```

Average radius to maximum wind 1981-2022. Values of USA_RMW (radius to maximum winds) and USA_EYE (eye diameter) are in nautical miles. ROCI: Radius of outer closed isobar. To convert to kilometers multiply by 1.852
```{r}
Tracks2.sf |>
  sf::st_drop_geometry() |>
#  dplyr::group_by(USA_PRES) |>
  dplyr::summarize(avgRMW = mean(USA_RMW, na.rm = TRUE) * 1.852,
                   avgEYE = mean(USA_EYE, na.rm = TRUE) * 1.852,
                   avgROCI = mean(USA_ROCI, na.rm = TRUE) * 1.852)
```

Fill in missing RMW values. Start with pedigree, then use minimum pressure, and finish again with pedigree
```{r}
Tracks2.sf <- Tracks2.sf |>
  dplyr::group_by(SID) |>  # pedigree
  dplyr::mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks2.sf <- Tracks2.sf |>
  dplyr::group_by(USA_PRES) |> # minimum pressure
  dplyr::mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks2.sf <- Tracks2.sf |>
  dplyr::group_by(SID) |> # pedigree
  dplyr::mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))
```
