---
title: "Validated storm-time-specific tropical cyclone exposure for Florida births"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  study_start: !r as.Date("2000-01-01")
  study_end: !r as.Date("2022-12-31")
  window_days: 7
  chunk_size: 50000
  progress_every: 5
  rebuild_ibtracs: false
  write_outputs: true
---

## Purpose

This workflow replaces the storm-wide, three-day exposure assignment in
`HO-storms.Rmd`. It uses individual IBTrACS storm observations and
quadrant-specific wind radii. A birth is exposed only when the residence at
delivery falls inside the wind field of the *same storm observation* used to
define the local impact date.

The primary direct-hazard threshold is 34 kt. A 55 kt sensitivity footprint is
estimated from the reported 34, 50, and 64 kt quadrant radii because IBTrACS
does not publish 55 kt radii. For each birth-storm pair, the local impact time
is the closest storm-center observation among observations whose wind field
contains the residence. The analysis window is `r params$window_days` days
before through `r params$window_days` days after that local impact date.

This is a residence-at-delivery environmental proxy. It does not establish the
mother's actual location, evacuation, displacement, flooding, power loss, or
personal storm experience.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(here)
library(lubridate)
library(sf)

options(stringsAsFactors = FALSE)
sf::sf_use_s2(TRUE)

dir.create(here("data", "derived"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("data", "outputs"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("figs", "exposure-validation"), recursive = TRUE,
           showWarnings = FALSE)

study_start <- as.Date(params$study_start)
study_end <- as.Date(params$study_end)

as_integer_parameter <- function(value, name) {
  parsed <- suppressWarnings(as.integer(sub("[Ll]$", "", as.character(value))))
  if (length(parsed) != 1L || is.na(parsed)) {
    stop("Parameter `", name, "` must be a single integer.", call. = FALSE)
  }
  parsed
}

window_days <- as_integer_parameter(params$window_days, "window_days")
chunk_size <- as_integer_parameter(params$chunk_size, "chunk_size")
progress_every <- as_integer_parameter(params$progress_every, "progress_every")
stopifnot(
  study_start <= study_end, window_days >= 0L, chunk_size > 0L,
  progress_every > 0L
)

progress_file <- here("data", "derived", "birth-exposure-progress.log")
if (file.exists(progress_file)) file.remove(progress_file)

log_progress <- function(stage, detail) {
  line <- sprintf(
    "%s | %-24s | %s",
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), stage, detail
  )
  message(line, appendLF = TRUE)
  write(line, file = progress_file, append = TRUE)
  invisible(line)
}

format_duration <- function(seconds) {
  if (!is.finite(seconds)) return("calculating")
  seconds <- max(0, round(seconds))
  sprintf(
    "%02d:%02d:%02d",
    seconds %/% 3600,
    (seconds %% 3600) %/% 60,
    seconds %% 60
  )
}

log_progress("setup", "workflow started")
```

## Download NOAA IBTrACS point observations

The point shapefile is required. The storm-level line shapefile already in the
repository cannot represent a time-varying wind field.

```{r ibtracs-download}
ibtracs_version <- "v04r01"
ibtracs_base <- paste0(
  "https://www.ncei.noaa.gov/data/",
  "international-best-track-archive-for-climate-stewardship-ibtracs/",
  ibtracs_version, "/access/shapefile"
)
ibtracs_zip_name <- paste0("IBTrACS.NA.list.", ibtracs_version, ".points.zip")
ibtracs_zip <- here("data", ibtracs_zip_name)
ibtracs_dir <- here("data", paste0("ibtracs-na-points-", ibtracs_version))

if (!file.exists(ibtracs_zip) || isTRUE(params$rebuild_ibtracs)) {
  download.file(
    paste(ibtracs_base, ibtracs_zip_name, sep = "/"),
    ibtracs_zip,
    mode = "wb"
  )
}

if (!dir.exists(ibtracs_dir) || isTRUE(params$rebuild_ibtracs)) {
  dir.create(ibtracs_dir, recursive = TRUE, showWarnings = FALSE)
  unzip(ibtracs_zip, exdir = ibtracs_dir)
}

ibtracs_shp <- list.files(
  ibtracs_dir, pattern = "[.]shp$", full.names = TRUE, ignore.case = TRUE
)
stopifnot(length(ibtracs_shp) == 1L)
```

## Read and validate births

Trimester definitions are weeks 1-13, weeks 14-28, and week 29 through
delivery. The supplied trimester dates are retained, rather than reconstructed
from delivery date, but their ordering and consistency are checked.

```{r births-read}
birth_file <- here("data", "all_births_hurricane.csv")
stopifnot(file.exists(birth_file))

births <- fread(
  birth_file,
  select = c(
    "ID", "DATE_OF_BIRTH", "final_lat", "final_lon",
    "trimester1_s_date", "trimester1_e_date",
    "trimester2_s_date", "trimester2_e_date",
    "trimester3_s_date"
  )
)

date_columns <- c(
  "DATE_OF_BIRTH", "trimester1_s_date", "trimester1_e_date",
  "trimester2_s_date", "trimester2_e_date", "trimester3_s_date"
)
births[, (date_columns) := lapply(.SD, as.IDate), .SDcols = date_columns]
births <- births[
  DATE_OF_BIRTH >= as.IDate(study_start) &
    DATE_OF_BIRTH <= as.IDate(study_end)
]

stopifnot(!anyNA(births$ID), uniqueN(births$ID) == nrow(births))

births[, valid_coordinates :=
         is.finite(final_lon) & is.finite(final_lat) &
         between(final_lon, -88.0, -79.0) &
         between(final_lat, 24.0, 32.0)]
births[, valid_gestation_dates :=
         !is.na(trimester1_s_date) & !is.na(trimester1_e_date) &
         !is.na(trimester2_s_date) & !is.na(trimester2_e_date) &
         !is.na(trimester3_s_date) &
         trimester1_s_date <= trimester1_e_date &
         trimester1_e_date < trimester2_s_date &
         trimester2_s_date <= trimester2_e_date &
         trimester2_e_date < trimester3_s_date &
         trimester3_s_date <= DATE_OF_BIRTH]

birth_qa <- births[, .(
  n_births = .N,
  n_invalid_coordinates = sum(!valid_coordinates),
  n_invalid_gestation_dates = sum(!valid_gestation_dates)
)]
print(birth_qa)

births_valid <- births[valid_coordinates & valid_gestation_dates]
birth_points <- st_as_sf(
  births_valid,
  coords = c("final_lon", "final_lat"),
  crs = 4326,
  remove = FALSE
)
```

## Read storm observations and calculate 55 kt radii

IBTrACS directly reports quadrant radii for 34, 50, and 64 kt winds. Within a
quadrant, the 55 kt radius is linearly interpolated between 50 and 64 kt when
both are available. If the 64 kt radius is unavailable, it is linearly
extrapolated inward from the 34-to-50 kt radial gradient. No RMW imputation or
storm-average radius is used.

```{r storms-read}
storm_points_raw <- st_read(ibtracs_shp, quiet = TRUE)
log_progress("storms", sprintf("read %s IBTrACS observations",
                                format(nrow(storm_points_raw), big.mark = ",")))

required_storm_columns <- c(
  "SID", "NAME", "ISO_TIME", "USA_WIND",
  paste0("USA_R34_", c("NE", "SE", "SW", "NW")),
  paste0("USA_R50_", c("NE", "SE", "SW", "NW")),
  paste0("USA_R64_", c("NE", "SE", "SW", "NW"))
)
missing_storm_columns <- setdiff(required_storm_columns, names(storm_points_raw))
stopifnot(length(missing_storm_columns) == 0L)

storm_points <- storm_points_raw |>
  select(all_of(required_storm_columns)) |>
  mutate(
    observation_time = ymd_hms(ISO_TIME, tz = "UTC", quiet = TRUE),
    observation_date = as.Date(observation_time),
    USA_WIND = as.numeric(USA_WIND)
  ) |>
  filter(
    observation_date >= study_start - days(window_days + 2L),
    observation_date <= study_end + days(window_days + 2L),
    USA_WIND >= 34
  )

radius_columns <- grep("^USA_R(34|50|64)_(NE|SE|SW|NW)$",
                       names(storm_points), value = TRUE)
for (column in radius_columns) {
  radius <- as.numeric(storm_points[[column]])
  radius[!is.finite(radius) | radius <= 0 | radius >= 999] <- NA_real_
  storm_points[[column]] <- radius
}

for (quadrant in c("NE", "SE", "SW", "NW")) {
  r34 <- storm_points[[paste0("USA_R34_", quadrant)]]
  r50 <- storm_points[[paste0("USA_R50_", quadrant)]]
  r64 <- storm_points[[paste0("USA_R64_", quadrant)]]

  r55_candidate <- ifelse(
    !is.na(r50) & !is.na(r64),
    r50 + (55 - 50) / (64 - 50) * (r64 - r50),
    ifelse(
      !is.na(r34) & !is.na(r50),
      pmax(0, r50 - (55 - 50) / (50 - 34) * (r34 - r50)),
      NA_real_
    )
  )
  # A 55 kt footprint cannot be more extensive than the observed 34 kt
  # footprint. Requiring R34 also guarantees that the sensitivity population
  # is nested within the primary exposure population.
  storm_points[[paste0("USA_R55_", quadrant)]] <- ifelse(
    is.na(r34) | is.na(r55_candidate), NA_real_, pmin(r34, r55_candidate)
  )
}

stopifnot(all(vapply(c("NE", "SE", "SW", "NW"), function(quadrant) {
  r34 <- storm_points[[paste0("USA_R34_", quadrant)]]
  r55 <- storm_points[[paste0("USA_R55_", quadrant)]]
  all(is.na(r55) | (!is.na(r34) & r55 <= r34))
}, logical(1))))

storm_radius_coverage <- rbindlist(lapply(c(34L, 55L), function(threshold) {
  columns <- paste0("USA_R", threshold, "_", c("NE", "SE", "SW", "NW"))
  data.table(
    threshold_kt = threshold,
    n_observations = nrow(storm_points),
    n_any_quadrant = sum(rowSums(!is.na(st_drop_geometry(storm_points)[columns])) > 0),
    n_all_quadrants = sum(rowSums(!is.na(st_drop_geometry(storm_points)[columns])) == 4)
  )
}))
storm_radius_coverage[, `:=`(
  pct_any_quadrant = 100 * n_any_quadrant / n_observations,
  pct_all_quadrants = 100 * n_all_quadrants / n_observations
)]
print(storm_radius_coverage)
```

## Construct storm-time-specific quadrant wind fields

Quadrants with missing radii remain missing; they are not silently treated as
zero-radius winds. The geometry uses a Florida-centered Albers equal-area CRS.
Each polygon retains its storm ID and observation timestamp.

```{r wind-field-functions}
analysis_crs <- 3086
nautical_mile_m <- 1852

sector_polygon <- function(center_xy, radius_m, bearing_start, bearing_end,
                           n = 36L) {
  if (!is.finite(radius_m) || radius_m <= 0) return(NULL)
  bearings <- seq(bearing_start, bearing_end, length.out = n + 1L)
  radians <- bearings * pi / 180
  arc <- cbind(
    center_xy[1] + radius_m * sin(radians),
    center_xy[2] + radius_m * cos(radians)
  )
  st_polygon(list(rbind(center_xy, arc, center_xy)))
}

make_wind_fields <- function(points_sf, threshold_kt) {
  projected <- st_transform(points_sf, analysis_crs)
  coordinates <- st_coordinates(projected)
  quadrant_spec <- data.table(
    quadrant = c("NE", "SE", "SW", "NW"),
    bearing_start = c(0, 90, 180, 270),
    bearing_end = c(90, 180, 270, 360)
  )

  geometries <- vector("list", nrow(projected))
  n_quadrants <- integer(nrow(projected))

  for (i in seq_len(nrow(projected))) {
    sectors <- vector("list", 0L)
    for (q in seq_len(nrow(quadrant_spec))) {
      radius_nm <- projected[[paste0(
        "USA_R", threshold_kt, "_", quadrant_spec$quadrant[q]
      )]][i]
      sector <- sector_polygon(
        coordinates[i, ], radius_nm * nautical_mile_m,
        quadrant_spec$bearing_start[q], quadrant_spec$bearing_end[q]
      )
      if (!is.null(sector)) sectors[[length(sectors) + 1L]] <- sector
    }

    n_quadrants[i] <- length(sectors)
    if (length(sectors) > 0L) {
      geometries[[i]] <- st_union(st_sfc(sectors, crs = analysis_crs))[[1]]
    } else {
      geometries[[i]] <- st_geometrycollection()
    }
  }

  st_sf(
    st_drop_geometry(projected) |>
      transmute(
        SID, NAME, observation_time, observation_date, USA_WIND,
        threshold_kt = as.integer(threshold_kt),
        observed_quadrants = n_quadrants
      ),
    geometry = st_sfc(geometries, crs = analysis_crs)
  ) |>
    filter(observed_quadrants > 0L, !st_is_empty(geometry))
}
```

Only observations whose coarse maximum-radius buffer could reach the study
residences are passed to the more expensive quadrant construction.

```{r wind-fields-build}
birth_extent <- st_bbox(st_transform(birth_points, analysis_crs))
birth_extent_polygon <- st_as_sfc(birth_extent) |> st_buffer(250 * 1000)

storm_points_projected <- st_transform(storm_points, analysis_crs)
near_florida <- lengths(st_intersects(storm_points_projected,
                                      birth_extent_polygon)) > 0L
storm_points_near <- storm_points[near_florida, ]

wind_fields_34 <- make_wind_fields(storm_points_near, 34L)
log_progress("wind fields", sprintf("built %s primary 34 kt observations",
                                     format(nrow(wind_fields_34), big.mark = ",")))
wind_fields_55 <- make_wind_fields(
  storm_points_near |> filter(USA_WIND >= 55), 55L
)
log_progress("wind fields", sprintf("built %s sensitivity 55 kt observations",
                                     format(nrow(wind_fields_55), big.mark = ",")))
wind_fields <- rbind(wind_fields_34, wind_fields_55) |>
  st_make_valid()

stopifnot(
  all(wind_fields$threshold_kt %in% c(34L, 55L)),
  !anyNA(wind_fields$SID),
  !anyNA(wind_fields$observation_time),
  all(!st_is_empty(wind_fields))
)
```

## Assign direct hazard and local impact time

The spatial join is performed against storm-time-specific polygons. Within
each birth, storm, and threshold, local impact is the closest storm-center
observation among the observations whose wind field contains the residence.

```{r exposure-assign}
storm_centers <- storm_points_near |>
  st_transform(analysis_crs) |>
  select(SID, observation_time, center_geometry = geometry)
storm_center_key <- paste(storm_centers$SID, storm_centers$observation_time)

assign_exposure_chunk <- function(index) {
  births_chunk <- st_transform(birth_points[index, ], analysis_crs) |>
    select(ID)

  matched <- st_join(
    births_chunk,
    wind_fields |>
      select(SID, NAME, observation_time, observation_date, USA_WIND,
             threshold_kt, observed_quadrants),
    join = st_within,
    left = FALSE
  )
  if (nrow(matched) == 0L) return(NULL)

  matched_key <- paste(matched$SID, matched$observation_time)
  center_index <- match(matched_key, storm_center_key)
  stopifnot(!anyNA(center_index))
  matched$center_distance_km <- as.numeric(
    st_distance(matched, storm_centers[center_index, ], by_element = TRUE)
  ) / 1000

  matched_dt <- matched |>
    st_drop_geometry() |>
    as.data.table()
  setorder(
    matched_dt, ID, SID, threshold_kt, center_distance_km,
    -USA_WIND, observation_time
  )
  matched_dt[, .SD[1L], by = .(ID, SID, threshold_kt)]
}

chunk_starts <- seq.int(1L, nrow(birth_points), by = chunk_size)
exposure_chunks <- vector("list", length(chunk_starts))
assignment_started <- Sys.time()
log_progress(
  "birth assignment",
  sprintf(
    "starting %s eligible births in %s chunks",
    format(nrow(birth_points), big.mark = ","),
    format(length(chunk_starts), big.mark = ",")
  )
)
for (j in seq_along(chunk_starts)) {
  chunk_started <- Sys.time()
  index <- chunk_starts[j]:min(
    chunk_starts[j] + chunk_size - 1L,
    nrow(birth_points)
  )
  exposure_chunks[[j]] <- assign_exposure_chunk(index)
  chunk_seconds <- as.numeric(difftime(Sys.time(), chunk_started, units = "secs"))
  elapsed_seconds <- as.numeric(
    difftime(Sys.time(), assignment_started, units = "secs")
  )
  completed_births <- max(index)
  completion_fraction <- j / length(chunk_starts)
  eta_seconds <- elapsed_seconds / completion_fraction - elapsed_seconds

  if (j == 1L || j %% progress_every == 0L || j == length(chunk_starts)) {
    log_progress(
      "birth assignment",
      sprintf(
        paste0(
          "chunk %s/%s | %s/%s births | %.1f%% | chunk %s | ",
          "elapsed %s | ETA %s | matched event rows %s"
        ),
        j, length(chunk_starts),
        format(completed_births, big.mark = ","),
        format(nrow(birth_points), big.mark = ","),
        100 * completion_fraction,
        format_duration(chunk_seconds),
        format_duration(elapsed_seconds),
        format_duration(eta_seconds),
        format(
          if (is.null(exposure_chunks[[j]])) 0L else nrow(exposure_chunks[[j]]),
          big.mark = ","
        )
      )
    )
  }
  gc()
}

birth_storm_events <- rbindlist(exposure_chunks, use.names = TRUE, fill = TRUE)

# Define local impact once from the primary 34 kt field. The 55 kt sensitivity
# indicates whether the residence also entered the stronger wind field, but it
# uses the same primary impact time. Threshold-specific impact times can differ
# by one or more IBTrACS observations and create artificial non-nesting at a
# trimester boundary.
primary_anchors <- birth_storm_events[
  threshold_kt == 34L,
  .(
    ID, SID,
    primary_observation_time = observation_time,
    primary_observation_date = observation_date,
    primary_center_distance_km = center_distance_km
  )
]

sensitivity_events <- merge(
  birth_storm_events[threshold_kt == 55L],
  primary_anchors,
  by = c("ID", "SID"),
  all.x = TRUE,
  sort = FALSE
)
stopifnot(
  !anyNA(sensitivity_events$primary_observation_time),
  !anyNA(sensitivity_events$primary_observation_date)
)
sensitivity_events[, `:=`(
  observation_time = primary_observation_time,
  observation_date = primary_observation_date,
  center_distance_km = primary_center_distance_km,
  primary_observation_time = NULL,
  primary_observation_date = NULL,
  primary_center_distance_km = NULL
)]

birth_storm_events <- rbindlist(
  list(
    birth_storm_events[threshold_kt == 34L],
    sensitivity_events
  ),
  use.names = TRUE,
  fill = TRUE
)
birth_storm_events[, `:=`(
  local_impact_date = as.IDate(observation_date),
  window_start = as.IDate(observation_date) - window_days,
  window_end = as.IDate(observation_date) + window_days
)]

stopifnot(
  birth_storm_events[, .N, by = .(ID, SID, threshold_kt)][, all(N == 1L)],
  all(birth_storm_events$window_start <= birth_storm_events$local_impact_date),
  all(birth_storm_events$local_impact_date <= birth_storm_events$window_end)
)
```

## Assign each event window to gestational periods

An event is assigned to a trimester when its residence-specific ±7-day window
overlaps that trimester. `acute_direct_*` separately indicates that the local
impact date itself occurred during the trimester.

```{r trimester-assign}
gestation_intervals <- rbindlist(list(
  births_valid[, .(ID, trimester = "T1", period_start = trimester1_s_date,
                   period_end = trimester1_e_date)],
  births_valid[, .(ID, trimester = "T2", period_start = trimester2_s_date,
                   period_end = trimester2_e_date)],
  births_valid[, .(ID, trimester = "T3", period_start = trimester3_s_date,
                   period_end = DATE_OF_BIRTH)]
))

event_windows <- merge(
  birth_storm_events,
  gestation_intervals,
  by = "ID",
  allow.cartesian = TRUE
)[window_start <= period_end & window_end >= period_start]

event_windows[, acute_direct :=
                local_impact_date >= period_start &
                local_impact_date <= period_end]

event_summary <- event_windows[, .(
  exposed_window = 1L,
  acute_direct = as.integer(any(acute_direct)),
  n_storms = uniqueN(SID),
  max_storm_wind_kt = max(USA_WIND, na.rm = TRUE),
  first_window_date = min(window_start),
  last_window_date = max(window_end)
), by = .(ID, trimester, threshold_kt)]

metric_columns <- c(
  "exposed_window", "acute_direct", "n_storms", "max_storm_wind_kt",
  "first_window_date", "last_window_date"
)
exposure_wide <- dcast(
  event_summary,
  ID ~ trimester + threshold_kt,
  value.var = metric_columns,
  sep = "_"
)

binary_count_columns <- grep(
  "^(exposed_window|acute_direct|n_storms)_", names(exposure_wide), value = TRUE
)
for (column in binary_count_columns) {
  set(exposure_wide, which(is.na(exposure_wide[[column]])), column, 0L)
}
```

## Merge by birth ID and add fixed-effect variables

This is a keyed merge, not an ordered column bind. `birth_year_fe` and
`conception_season_fe` are model-ready fixed effects. Calendar month is also
provided for a stricter seasonality specification.

```{r analytic-merge}
all_files <- list.files(
  here("data", "21_definition_final_re"),
  pattern = "[.]Rdata$", full.names = TRUE
)
stopifnot(length(all_files) > 0L)

load_one_object <- function(path) {
  environment <- new.env(parent = emptyenv())
  object_names <- load(path, envir = environment)
  stopifnot(length(object_names) == 1L)
  as.data.table(environment[[object_names]])
}

analytic_births <- rbindlist(
  lapply(all_files, load_one_object), use.names = TRUE, fill = TRUE
)
stopifnot(!anyNA(analytic_births$ID), uniqueN(analytic_births$ID) == nrow(analytic_births))

# Approved analytic columns sent by Jessica Broach on April 14, 2025. The
# corrected exposure variables, eligibility flag, and temporal fixed-effect
# variables are appended later in this workflow.
approved_birth_columns <- c(
  "ID", "final_lat", "final_lon", "ALCOHOL_DRINKS", "ALCOHOL_USE",
  "BIRTH_WEIGHT_GRAMS", "CalculatedGestationalAge", "DATE_OF_BIRTH",
  "GESTATION_WEEKS", "FATHER_AGE", "FATHER_EDCODE",
  "Father_CalculatedHisp", "Father_CalculatedRace", "MENSES_DATE",
  "MOTHER_AGE", "Mother_CalculatedHisp", "Mother_CalculatedRace",
  "MOTHER_EDCODE", "MOTHER_MARRIED", "MOTHER_RES_COUNTY",
  "MOTHER_WIC_YESNO", "MR_DIAB", "MR_DIAB_GEST", "MR_HYPERT_CHRONIC",
  "MR_HYPERT_ECLAMPSIA", "MR_HYPERT_PREG", "MR_PREV_PRETERM", "PLURALITY_CODE",
  "PrePregnancy_BMI", "PRINCIPAL_SOURCE_PAY", "PRINCIPAL_SRCPAY_CODE",
  "SEX", "m28_temperature", "lat_pm25", "lon_pm25", "date_pm25",
  "loc_data_pm25", "m28_pm25", "m21_pm25", "m14_pm25", "m7_pm25",
  "dday_pm25", "p7_pm25", "p14_pm25", "p21_pm25", "p28_pm25",
  "m20_pm25", "m19_pm25", "m18_pm25", "m17_pm25", "m16_pm25",
  "m15_pm25", "m14_pm25.1", "m13_pm25", "m12_pm25", "m11_pm25",
  "m10_pm25", "m9_pm25", "m8_pm25", "m7_pm25.1", "m6_pm25",
  "m5_pm25", "m4_pm25", "m3_pm25", "m2_pm25", "m1_pm25", "m0_pm25",
  "year_pm25r", "d_1_13_wk_90_1", "d_14_28_wk_90_1",
  "d_from29_wk_90_1", "d_1_13_wk_95_1", "d_14_28_wk_95_1",
  "d_from29_wk_95_1", "d_1_13_wk_98_1", "d_14_28_wk_98_1",
  "d_from29_wk_98_1", "d_1_13_wk_90_2", "d_14_28_wk_90_2",
  "d_from29_wk_90_2", "d_1_13_wk_95_2", "d_14_28_wk_95_2",
  "d_from29_wk_95_2", "d_1_13_wk_98_2", "d_14_28_wk_98_2",
  "d_from29_wk_98_2", "d_1_13_wk_90_3", "d_14_28_wk_90_3",
  "d_from29_wk_90_3", "d_1_13_wk_95_3", "d_14_28_wk_95_3",
  "d_from29_wk_95_3", "d_1_13_wk_98_3", "d_14_28_wk_98_3",
  "d_from29_wk_98_3", "trimester1_length", "trimester2_length",
  "trimester3_length"
)

missing_approved_columns <- setdiff(approved_birth_columns, names(analytic_births))
if (length(missing_approved_columns) > 0L) {
  stop(
    "Approved birth columns missing from the yearly files: ",
    paste(missing_approved_columns, collapse = ", "),
    call. = FALSE
  )
}
analytic_births <- analytic_births[, ..approved_birth_columns]
log_progress(
  "analytic merge",
  sprintf("retained %s approved/model-required source columns",
          format(length(approved_birth_columns), big.mark = ","))
)

parse_mixed_idate <- function(x) {
  # Year-specific files can store dates as POSIXct, Date, IDate, or character.
  # Normalize through an ISO date string and suppress low-level POSIX warnings;
  # unparseable values remain NA and are counted below.
  if (inherits(x, "POSIXt")) {
    text <- suppressWarnings(format(x, "%Y-%m-%d", tz = "UTC"))
  } else {
    text <- substr(as.character(x), 1L, 10L)
  }
  suppressWarnings(as.IDate(text, format = "%Y-%m-%d"))
}

analytic_births[, DATE_OF_BIRTH := parse_mixed_idate(DATE_OF_BIRTH)]
missing_birth_dates <- analytic_births[is.na(DATE_OF_BIRTH), .N]
log_progress(
  "analytic merge",
  sprintf("%s records have an unparseable birth date and will be excluded",
          format(missing_birth_dates, big.mark = ","))
)
analytic_births <- analytic_births[
  !is.na(DATE_OF_BIRTH) &
  DATE_OF_BIRTH >= as.IDate(study_start) &
    DATE_OF_BIRTH <= as.IDate(study_end)
]
analytic_births[, conception_date := parse_mixed_idate(MENSES_DATE)]
if ("GESTATION_WEEKS" %in% names(analytic_births)) {
  analytic_births[
    is.na(conception_date) & is.finite(GESTATION_WEEKS),
    conception_date := DATE_OF_BIRTH - as.integer(round(GESTATION_WEEKS * 7))
  ]
}

analytic_births[, `:=`(
  birth_year = year(DATE_OF_BIRTH),
  birth_month = month(DATE_OF_BIRTH),
  conception_year = year(conception_date),
  conception_month = month(conception_date)
)]
analytic_births[, conception_season := fcase(
  conception_month %in% c(12L, 1L, 2L), "DJF",
  conception_month %in% 3:5, "MAM",
  conception_month %in% 6:8, "JJA",
  conception_month %in% 9:11, "SON",
  default = NA_character_
)]

birth_exposure_eligibility <- births[, .(
  ID,
  exposure_eligible = valid_coordinates & valid_gestation_dates
)]

analytic_corrected <- merge(
  analytic_births,
  birth_exposure_eligibility,
  by = "ID",
  all.x = TRUE,
  sort = FALSE
)
analytic_corrected[is.na(exposure_eligible), exposure_eligible := FALSE]

analytic_corrected <- merge(
  analytic_corrected,
  exposure_wide,
  by = "ID",
  all.x = TRUE,
  sort = FALSE
)

expected_exposure_columns <- as.vector(outer(
  c("exposed_window", "acute_direct", "n_storms"),
  as.vector(outer(c("T1", "T2", "T3"), c(34L, 55L), paste, sep = "_")),
  paste, sep = "_"
))
for (column in expected_exposure_columns) {
  if (!column %in% names(analytic_corrected)) {
    analytic_corrected[, (column) := NA_integer_]
  }
  set(
    analytic_corrected,
    which(analytic_corrected$exposure_eligible &
            is.na(analytic_corrected[[column]])),
    column,
    0L
  )
}

analytic_corrected[, `:=`(
  birth_year_fe = factor(birth_year),
  conception_season_fe = factor(
    conception_season, levels = c("DJF", "MAM", "JJA", "SON")
  ),
  birth_month_fe = factor(birth_month)
)]

stopifnot(
  nrow(analytic_corrected) == nrow(analytic_births),
  uniqueN(analytic_corrected$ID) == nrow(analytic_corrected),
  all(analytic_corrected$ID %in% analytic_births$ID)
)

for (trimester in c("T1", "T2", "T3")) {
  primary <- analytic_corrected[[paste0("exposed_window_", trimester, "_34")]]
  sensitivity <- analytic_corrected[[paste0("exposed_window_", trimester, "_55")]]
  stopifnot(all(is.na(sensitivity) | is.na(primary) | sensitivity <= primary))
}
```

## Validation outputs

```{r validation}
exposure_prevalence <- melt(
  analytic_corrected,
  id.vars = c("ID", "birth_year"),
  measure.vars = patterns(exposed = "^exposed_window_T[123]_(34|55)$"),
  variable.name = "exposure_variable",
  value.name = "exposed"
)[, .(
  eligible_births = sum(!is.na(exposed)),
  exposed_births = sum(exposed, na.rm = TRUE),
  prevalence = mean(exposed, na.rm = TRUE)
), by = .(birth_year, exposure_variable)]

storm_event_qa <- event_windows[, .(
  exposed_births = uniqueN(ID),
  local_impact_start = min(local_impact_date),
  local_impact_end = max(local_impact_date),
  median_center_distance_km = median(center_distance_km, na.rm = TRUE),
  min_observed_quadrants = min(observed_quadrants),
  max_observed_quadrants = max(observed_quadrants)
), by = .(SID, NAME, threshold_kt, trimester)]

print(exposure_prevalence)
print(storm_event_qa[order(-exposed_births)][1:min(.N, 30L)])

prevalence_plot <- ggplot(
  exposure_prevalence,
  aes(birth_year, prevalence, color = exposure_variable)
) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Birth year", y = "Exposure prevalence",
    color = "Trimester / threshold",
    title = "Corrected residence-based TC exposure prevalence"
  ) +
  theme_minimal()
print(prevalence_plot)

ggsave(
  here("figs", "exposure-validation", "exposure-prevalence-by-year.png"),
  prevalence_plot, width = 10, height = 6, dpi = 300
)

stopifnot(
  all(exposure_prevalence$prevalence >= 0 &
        exposure_prevalence$prevalence <= 1),
  all(event_windows$threshold_kt %in% c(34L, 55L)),
  all(event_windows$window_start <= event_windows$period_end),
  all(event_windows$window_end >= event_windows$period_start)
)
```

## Save corrected birth data

The event-level file is retained for auditability. The analytic file contains
one row per birth and is the input for refitting all birth-outcome models.

```{r write-output}
if (isTRUE(params$write_outputs)) {
  fwrite(
    birth_storm_events,
    here("data", "derived", "birth-storm-events-time-specific.csv")
  )
  fwrite(
    storm_radius_coverage,
    here("data", "derived", "storm-wind-radius-coverage.csv")
  )
  fwrite(
    storm_event_qa,
    here("data", "derived", "birth-storm-exposure-validation.csv")
  )
  saveRDS(
    analytic_corrected,
    here("data", "outputs", "All_Births_Corrected_Storm_Exposure.rds"),
    compress = "xz"
  )
}
log_progress("complete", "corrected birth exposure workflow finished")
```

## Required model specification

All previously reported birth-outcome models must be refit. The 34 kt
`exposed_window` variable is primary; the corresponding 55 kt variable is the
sensitivity exposure. Every model must include calendar-year and season fixed
effects. The existing substantive covariates should be added to `covariates`
after their coding and missing-data strategy are finalized.

```{r model-template, eval=FALSE}
covariates <- c(
  "Mother_CalculatedRace", "Mother_CalculatedHisp", "MOTHER_EDCODE",
  "Father_CalculatedRace", "Father_CalculatedHisp", "FATHER_EDCODE"
)

fit_birth_model <- function(data, outcome, trimester = "T1",
                            threshold_kt = 34L) {
  exposure <- paste0("exposed_window_", trimester, "_", threshold_kt)
  model_formula <- reformulate(
    c(
      exposure,
      covariates,
      "birth_year_fe",
      "conception_season_fe"
      # Prefer birth_month_fe instead of conception_season_fe in a stricter
      # seasonality sensitivity model; do not include both simultaneously.
    ),
    response = outcome
  )
  glm(model_formula, data = data, family = binomial())
}

ptb_t1_primary <- fit_birth_model(
  analytic_corrected, outcome = "PTB", trimester = "T1", threshold_kt = 34L
)
ptb_t1_sensitivity <- fit_birth_model(
  analytic_corrected, outcome = "PTB", trimester = "T1", threshold_kt = 55L
)
```

## Interpretation constraints

1. The ±7-day variable represents a direct-wind event window around local
   impact, not 15 days of continuous wind exposure.
2. The 55 kt footprint is an empirical interpolation and must be labeled as a
   sensitivity analysis.
3. Missing wind radii can cause under-ascertainment; report the coverage table
   by threshold and study period.
4. Residence at delivery may differ from residence or actual location at
   impact. Evacuation and out-migration remain unmeasured.
5. Rainfall, flooding, storm surge, outages, and healthcare disruption are not
   represented by this direct-wind measure.
