---
title: "Health Outcomes & Storms"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Storm threat and impact model

Get and import the IBTraCS storm data

https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/
https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/doc/IBTrACS_v04_Technical_Details.pdf
https://www.ncei.noaa.gov/sites/default/files/2021-07/IBTrACS_v04_column_documentation.pdf

```{r}
L <- "https://www.ncei.noaa.gov/data/international-best-track-archive-for-climate-stewardship-ibtracs/v04r00/access/shapefile/IBTrACS.NA.list.v04r00.lines.zip"
  
if(!"IBTrACS.NA.list.v04r00.lines.zip" %in% list.files(here::here("data"))) {
download.file(url = L,
              destfile = here::here("data",
                                    "IBTrACS.NA.list.v04r00.lines.zip"))
unzip(here::here("data", "IBTrACS.NA.list.v04r00.lines.zip"),
      exdir = here::here("data"))
}

Tracks.sf <- sf::st_read(dsn = here::here("data"), 
                         layer = "IBTrACS.NA.list.v04r00.lines") |>
  sf::st_transform(crs = 32616)
```

Geometry type LINESTRING. Wind speed is in units of knots (nautical mile per hour)

Keep only storms having USA_WIND >= 64 occurring between 1981 and 2022
```{r}
Tracks2.sf <- Tracks.sf |>
  dplyr::filter(year >= 1981 & year <= 2022) |>
  dplyr::filter(USA_WIND >= 64) |> # change to 34 for tropical storms and hurricanes
  dplyr::select(SID, SEASON, year, month, day, hour, min,
                NAME, SUBBASIN, ISO_TIME,
                USA_WIND, USA_PRES, USA_RMW, USA_EYE, USA_ROCI)
```

Average radius to maximum wind 1981-2022. Values of USA_RMW (radius to maximum winds) and USA_EYE (eye diameter) are in nautical miles. ROCI: Radius of outer closed isobar. To convert to kilometers multiply by 1.852
```{r}
Tracks2.sf |>
  sf::st_drop_geometry() |>
#  dplyr::group_by(USA_PRES) |>
  dplyr::summarize(avgRMW = mean(USA_RMW, na.rm = TRUE) * 1.852,
                   avgEYE = mean(USA_EYE, na.rm = TRUE) * 1.852,
                   avgROCI = mean(USA_ROCI, na.rm = TRUE) * 1.852)
```

Fill in missing RMW values. Start with pedigree, then use minimum pressure, and finish again with pedigree
```{r}
Tracks2.sf <- Tracks2.sf |>
  dplyr::group_by(SID) |>  # pedigree
  dplyr::mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks2.sf <- Tracks2.sf |>
  dplyr::group_by(USA_PRES) |> # minimum pressure
  dplyr::mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))

Tracks2.sf <- Tracks2.sf |>
  dplyr::group_by(SID) |> # pedigree
  dplyr::mutate(USA_RMW = ifelse(is.na(USA_RMW), mean(USA_RMW, na.rm = TRUE), USA_RMW))
```

Add a buffer to the tracks to make segmented wind swaths
```{r}
Swaths.sf <- Tracks2.sf |>
  sf::st_buffer(dist = Tracks2.sf$USA_RMW * 1852) # 1852 converts to meters

tmap::tmap_mode("view")
tmap::tm_shape(Swaths.sf) +
  tmap::tm_borders()
```

Wind swaths that cross Florida. `USAboundaries` package no longer maintained on CRAN
```{r}
devtools::install_github("ropensci/USAboundariesData")
devtools::install_github("ropensci/USAboundaries")

Boundaries.sf <- USAboundaries::us_states(resolution = "low", states = "FL") |> 
  sf::st_transform(crs = 32616)

X <- Swaths.sf |>
  sf::st_intersects(Boundaries.sf, sparse = FALSE)
Swaths.sf <- Swaths.sf[X, ]
Swaths.sf <- Swaths.sf |>
  dplyr::mutate(Date = lubridate::as_date(ISO_TIME)) # add a m-y-d column

tmap::tm_shape(Swaths.sf) +
  tmap::tm_borders()
```

Extract the boundaries of storm impacts by storm ID
```{r}
Swaths2.sf <- Swaths.sf |>
  dplyr::group_by(SID) |>
  dplyr::summarize(Date0 = dplyr::first(Date),
                   NAME = dplyr::first(NAME),
                   geometry = sf::st_union(geometry))

tmap::tm_shape(Swaths2.sf) +
  tmap::tm_borders()
```

Each storm has a unique polygon (or multipolygon) and an impact date

Make the geometry have a geographic CRS 4326 and unionize the swaths
```{r}
Swaths2.sf <- Swaths2.sf |>
    sf::st_transform(crs = 4326)
union_Swaths2.sfg <- Swaths2.sf |>
  sf::st_union()
```

Expand `Swaths2.sf` by adding rows based on the increment value of the attribute `Date0`
```{r}
deltaT <- 1   # One day increment
n_new <- 1    # Number of new rows to create for each feature (row) times 2

TIC.sf <- Swaths2.sf |>
    dplyr::rowwise() |>
    dplyr::mutate(new_features = list(tidyr::tibble(
                  Date = Date0 + (-n_new:n_new) * deltaT))) |>
    tidyr::unnest(new_features) |>
    dplyr::ungroup() |>
    dplyr::select(Date, NAME)

TIC <- rep(rep(c("Threat", "Impact", "Cleanup"), times = c(1, 1, 1)), 
           times = nrow(Swaths2.sf))
TIC.sf$TIC <- TIC
```

## Merge with birth records

Email from Jihoon on December 11, 2024
1. ID 2. DATE_OF_BIRTH 3. final_lat  4. final_lon 5. trimester1_s_date: The first Trimester start date
6. trimester1_e_date : The first Trimester end date 7. trimester2_s_date : The second Trimester start date
8. trimester2_e_date : The second Trimester end date 9. trimester3_s_date: The third Trimester start date
The first Trimester: the weeks 1-13, The second Trimester: the weeks 14-28, The third Trimester: the weeks 29 ~
```{r}
Births.df <- read.csv(file = "data/all_births_hurricane.csv")
```

7.2 million births. Time period is 1981-1-1 until 2022-12-31. Only dates are given not times. Lat/lon ranges appear to include Texas through Florida 

Make the data frame a simple feature data frame with POINT geometry
```{r}
Births.sf <- Births.df |>
  dplyr::mutate(Date = lubridate::as_date(DATE_OF_BIRTH),
                DateT1s = lubridate::as_date(trimester1_s_date),
                DateT1e = lubridate::as_date(trimester1_e_date),
                DateT2s = lubridate::as_date(trimester2_s_date),
                DateT2e = lubridate::as_date(trimester2_e_date),
                DateT3s = lubridate::as_date(trimester3_s_date)
                ) |>
  sf::st_as_sf(coords = c("final_lon", "final_lat"),
               crs = 4326) |>
  dplyr::select(Birth_ID = ID, Birth_Date = Date, DateT1s, DateT1e, DateT2s, DateT2e, DateT3s)
```

Keep only births occurring inside the union of all threat/impact/cleanup polygons
```{r}
withIn <- Births.sf |>
  sf::st_within(union_Swaths2.sfg, sparse = FALSE)
BirthsWithin.sf <- Births.sf[withIn, ]
```

Check with a map. Sample 10,000 birth records
```{r}
nrows <- nrow(BirthsWithin.sf)
keep_rows <- sample.int(nrows, size = 10000)

tmap::tm_shape(Swaths2.sf) +
  tmap::tm_borders() +
tmap::tm_shape(BirthsWithin.sf$geometry[keep_rows]) +
  tmap::tm_dots()
```

Filter storms by third trimester date range. Test on a sample of 500 births
```{r}
keep_rows <- sample.int(nrow(BirthsWithin.sf), size = 500)
merge_Test.sf <- BirthsWithin.sf[keep_rows, ]

sf::st_geometry(TIC.sf) <- "geom" # can't have matching column names

# 1. Filter storms by trimester date range
births_storm_filtered <- merge_Test.sf |>
  dplyr::rowwise() |>
  dplyr::mutate(
    matched_storms = list(
      TIC.sf |>
        dplyr::filter(Date >= DateT3s & Date <= Birth_Date))) |>
  tidyr::unnest(matched_storms, keep_empty = TRUE)

# 2. Perform spatial join to check if birth locations fall within the storm polygons
births_with_impacts <- sf::st_join(births_storm_filtered, TIC.sf, join = st_within)

# 3. Keep only records where a birth was inside a storm impact area
X <- births_with_impacts |>
  dplyr::filter(!is.na(NAME.x)) |>
  dplyr::select(Birth_ID, Birth_Date, DateT3s,
                NAME = NAME.x, TIC = TIC.x, geometry) |>
  dplyr::distinct()
```

Check on a map
```{r}
tmap::tm_shape(Swaths2.sf) +
  tmap::tm_borders() +
tmap::tm_shape(births_with_impacts$geometry) +
  tmap::tm_dots()
```

## Merge with death records

File received from Jihoon on December 7, 2024 via email. From email:
1) Whether impacted or not when a hurricane passes on the date of death? What was impacted? 2) Whether impacted or not during a warning on the date of death
3) Whether impacted or not during a watch on the date of death. 4) Whether impacted or not during an advisory on the date of death
5) (If available) Whether impacted or not during an evacuation order on the date of death

```{r}
Deaths.df <- read.csv(file = "data/all_deaths_hurricane.csv")
```

6.2 million deaths. Time period is 1981-1-1 until 2022-12-31. Only dates are given not times. Note: Watch/warning data are only available starting in 2008. lat/lon ranges appear to include Texas through Florida
```{r}
Deaths.sf <- Deaths.df |>
  dplyr::mutate(Date = lubridate::as_date(DATE_OF_DEATH)) |>
  sf::st_as_sf(coords = c("final_lon", "final_lat"),
               crs = 4326) |>
  dplyr::select(Death_ID = ID, Death_Date = Date)
```

Keep only deaths occurring inside the union of all threat/impact/cleanup polygons
```{r}
withIn <- Deaths.sf |>
  sf::st_within(union_Swaths2.sfg, sparse = FALSE)
DeathsWithin.sf <- Deaths.sf[withIn, ]
```

Check with a map. Sample 10,000 death records
```{r}
nrows <- nrow(DeathsWithin.sf)
keep_rows <- sample.int(nrows, size = 10000)

tmap::tm_shape(Swaths2.sf) +
  tmap::tm_borders() +
tmap::tm_shape(DeathsWithin.sf$geometry[keep_rows]) +
  tmap::tm_dots()
```

Filter storms by death date. Test on a sample of 10000 deaths
```{r}
keep_rows <- sample.int(nrow(DeathsWithin.sf), size = 10000)
merge_Test.sf <- DeathsWithin.sf[keep_rows, ]

sf::st_geometry(TIC.sf) <- "geom" # can't have matching column names

# 1. Filter storms by death date
deaths_storm_filtered <- merge_Test.sf |>
  dplyr::rowwise() |>
  dplyr::mutate(
    matched_storms = list(
      TIC.sf |>
        dplyr::filter(Date == Death_Date))) |>
  tidyr::unnest(matched_storms, keep_empty = TRUE)

# 2. Perform spatial join to check if birth locations fall within the storm polygons
deaths_with_impacts <- sf::st_join(deaths_storm_filtered, TIC.sf, join = st_within)

# 3. Keep only records where a death was inside a storm impact area
deaths_with_impacts <- deaths_with_impacts |>
  dplyr::filter(!is.na(NAME.x)) |>
  dplyr::select(Death_ID, Death_Date,
                NAME = NAME.x, TIC = TIC.x, geometry) |>
  dplyr::distinct()
```

Check on a map
```{r}
tmap::tm_shape(Swaths2.sf) +
  tmap::tm_borders() +
tmap::tm_shape(deaths_with_impacts$geometry) +
  tmap::tm_dots()
```

Convert to regular data frame with X (Longitude) and Y (Latitude)
df <- sf_data %>%
  mutate(
    X = st_coordinates(.)[,1],  # Extract longitude
    Y = st_coordinates(.)[,2]   # Extract latitude
  ) %>%
  st_drop_geometry() 


## Toy example of a space-time merge using simple feature data frames

```{r}
library(sf)
library(dplyr)

# Example: Storm data (impact polygons and attributes)
storms <- st_sf(
  storm_name = c("Storm A", "Storm B", "Storm C"),
  impact_level = c(1, 2, 3),
  storm_date = as.Date(c("2023-01-15", "2023-02-20", "2023-03-10")),
  geometry = st_sfc(
    st_polygon(list(rbind(c(0, 0), c(0, 5), c(5, 5), c(5, 0), c(0, 0)))),
    st_polygon(list(rbind(c(3, 3), c(3, 8), c(8, 8), c(8, 3), c(3, 3)))),
    st_polygon(list(rbind(c(6, 6), c(6, 9), c(9, 9), c(9, 6), c(6, 6))))
  ),
  crs = 4326
)

# Example: Birth records (point locations and third trimester dates)
births <- st_sf(
  birth_date = as.Date(c("2023-05-01", "2023-03-01", "2023-01-20", "2023-09-01")),
  trimester_start = as.Date(c("2023-02-01", "2022-12-01", "2022-10-01", "2022-12-01")),
  trimester_end = as.Date(c("2023-04-30", "2023-02-28", "2023-01-19", "2023-08-30")),
  geometry = st_sfc(
    st_point(c(2, 2)),
    st_point(c(4, 4)),
    st_point(c(7, 7)),
    st_point(c(1, 7))
  ),
  crs = 4326
)

st_geometry(storms) <- "geom"

# 1. Filter storms by third trimester date range
births_storms_filtered <- births %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    matched_storms = list(
      storms %>%
        dplyr::filter(storm_date >= trimester_start & storm_date <= trimester_end)
    )
  ) %>%
  tidyr::unnest(matched_storms, keep_empty = TRUE)

# 2. Perform spatial join to check if birth locations fall within the storm polygons
births_with_impacts <- sf::st_join(births_storms_filtered, storms, join = st_within)

# 3. Keep only records where a birth was inside a storm impact area
births_with_impacts <- births_with_impacts %>%
  dplyr::filter(!is.na(storm_name.y)) %>%
  dplyr::select(birth_date, trimester_start, trimester_end, 
                storm_name = storm_name.y, impact_level = impact_level.y, geometry) |>
  dplyr::distinct()

# Display the final dataset
print(births_with_impacts)
```


