---
title: "Validated storm-time-specific tropical cyclone exposure for Florida births"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  study_start: !r as.Date("2000-01-01")
  study_end: !r as.Date("2022-12-31")
  window_days: 7
  chunk_size: 50000
  progress_every: 5
  output_tag: "trimester-eligible-v2"
  rebuild_ibtracs: false
  write_outputs: true
---

## Purpose

This workflow replaces the storm-wide, three-day exposure assignment in
`HO-storms.Rmd`. It uses individual IBTrACS storm observations and
quadrant-specific wind radii. A birth is exposed only when the residence at
delivery falls inside the wind field of the *same storm observation* used to
define the local impact date.

The primary direct-hazard threshold is 34 kt. A 55 kt sensitivity footprint is
estimated from the reported 34, 50, and 64 kt quadrant radii because IBTrACS
does not publish 55 kt radii. For each birth-storm pair, the local impact time
is the closest storm-center observation among observations whose wind field
contains the residence. The analysis window is `r params$window_days` days
before through `r params$window_days` days after that local impact date.

This is a residence-at-delivery environmental proxy. It does not establish the
mother's actual location, evacuation, displacement, flooding, power loss, or
personal storm experience.

## Setup

```{r setup, message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(ggplot2)
library(here)
library(lubridate)
library(sf)

options(stringsAsFactors = FALSE)
sf::sf_use_s2(TRUE)

dir.create(here("data", "derived"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("data", "outputs"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("figs", "exposure-validation"), recursive = TRUE,
           showWarnings = FALSE)

output_tag <- as.character(params$output_tag)
stopifnot(length(output_tag) == 1L, nzchar(output_tag))
output_suffix <- paste0("-", output_tag)

study_start <- as.Date(params$study_start)
study_end <- as.Date(params$study_end)

as_integer_parameter <- function(value, name) {
  parsed <- suppressWarnings(as.integer(sub("[Ll]$", "", as.character(value))))
  if (length(parsed) != 1L || is.na(parsed)) {
    stop("Parameter `", name, "` must be a single integer.", call. = FALSE)
  }
  parsed
}

window_days <- as_integer_parameter(params$window_days, "window_days")
chunk_size <- as_integer_parameter(params$chunk_size, "chunk_size")
progress_every <- as_integer_parameter(params$progress_every, "progress_every")
stopifnot(
  study_start <= study_end, window_days >= 0L, chunk_size > 0L,
  progress_every > 0L
)

progress_file <- here(
  "data", "derived", paste0("birth-exposure-progress", output_suffix, ".log")
)
if (file.exists(progress_file)) file.remove(progress_file)

log_progress <- function(stage, detail) {
  line <- sprintf(
    "%s | %-24s | %s",
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), stage, detail
  )
  message(line, appendLF = TRUE)
  write(line, file = progress_file, append = TRUE)
  invisible(line)
}

format_duration <- function(seconds) {
  if (!is.finite(seconds)) return("calculating")
  seconds <- max(0, round(seconds))
  sprintf(
    "%02d:%02d:%02d",
    seconds %/% 3600,
    (seconds %% 3600) %/% 60,
    seconds %% 60
  )
}

log_progress("setup", "workflow started")
```

## Download NOAA IBTrACS point observations

The point shapefile is required. The storm-level line shapefile already in the
repository cannot represent a time-varying wind field.

```{r ibtracs-download}
ibtracs_version <- "v04r01"
ibtracs_base <- paste0(
  "https://www.ncei.noaa.gov/data/",
  "international-best-track-archive-for-climate-stewardship-ibtracs/",
  ibtracs_version, "/access/shapefile"
)
ibtracs_zip_name <- paste0("IBTrACS.NA.list.", ibtracs_version, ".points.zip")
ibtracs_zip <- here("data", ibtracs_zip_name)
ibtracs_dir <- here("data", paste0("ibtracs-na-points-", ibtracs_version))

if (!file.exists(ibtracs_zip) || isTRUE(params$rebuild_ibtracs)) {
  download.file(
    paste(ibtracs_base, ibtracs_zip_name, sep = "/"),
    ibtracs_zip,
    mode = "wb"
  )
}

if (!dir.exists(ibtracs_dir) || isTRUE(params$rebuild_ibtracs)) {
  dir.create(ibtracs_dir, recursive = TRUE, showWarnings = FALSE)
  unzip(ibtracs_zip, exdir = ibtracs_dir)
}

ibtracs_shp <- list.files(
  ibtracs_dir, pattern = "[.]shp$", full.names = TRUE, ignore.case = TRUE
)
stopifnot(length(ibtracs_shp) == 1L)
```

## Read and validate births

Trimester definitions are weeks 1-13, weeks 14-28, and week 29 through
delivery. Eligibility is evaluated separately for each trimester. A pregnancy
that ends before a later trimester remains eligible for every earlier observed
period; it is not excluded merely because it never reached trimester 3.

```{r births-read}
birth_file <- here("data", "all_births_hurricane.csv")
stopifnot(file.exists(birth_file))

births <- fread(
  birth_file,
  select = c(
    "ID", "DATE_OF_BIRTH", "final_lat", "final_lon",
    "trimester1_s_date", "trimester1_e_date",
    "trimester2_s_date", "trimester2_e_date",
    "trimester3_s_date"
  )
)

date_columns <- c(
  "DATE_OF_BIRTH", "trimester1_s_date", "trimester1_e_date",
  "trimester2_s_date", "trimester2_e_date", "trimester3_s_date"
)
births[, (date_columns) := lapply(.SD, as.IDate), .SDcols = date_columns]
births <- births[
  DATE_OF_BIRTH >= as.IDate(study_start) &
    DATE_OF_BIRTH <= as.IDate(study_end)
]

stopifnot(!anyNA(births$ID), uniqueN(births$ID) == nrow(births))

births[, valid_coordinates :=
         is.finite(final_lon) & is.finite(final_lat) &
         between(final_lon, -88.0, -79.0) &
         between(final_lat, 24.0, 32.0)]

# Period ends are truncated at delivery. This preserves early deliveries in
# every trimester they actually reached and prevents post-delivery exposure.
births[, `:=`(
  T1_start = trimester1_s_date,
  T1_end = pmin(trimester1_e_date, DATE_OF_BIRTH, na.rm = FALSE),
  T2_start = trimester2_s_date,
  T2_end = pmin(trimester2_e_date, DATE_OF_BIRTH, na.rm = FALSE),
  T3_start = trimester3_s_date,
  T3_end = DATE_OF_BIRTH
)]

births[, `:=`(
  valid_T1_dates = !is.na(T1_start) & !is.na(T1_end) & T1_start <= T1_end,
  valid_T2_dates = !is.na(T2_start) & !is.na(T2_end) & T2_start <= T2_end,
  valid_T3_dates = !is.na(T3_start) & !is.na(T3_end) & T3_start <= T3_end
)]

birth_qa <- births[, .(
  n_births = .N,
  n_invalid_coordinates = sum(!valid_coordinates),
  n_valid_T1_dates = sum(valid_T1_dates),
  n_valid_T2_dates = sum(valid_T2_dates),
  n_valid_T3_dates = sum(valid_T3_dates),
  n_born_before_29_weeks = sum(
    as.numeric(DATE_OF_BIRTH - T1_start) < 28 * 7,
    na.rm = TRUE
  )
)]
print(birth_qa)

# County and storm assignment depend only on residence coordinates, not on
# whether a pregnancy reached a later trimester.
births_coordinate_valid <- births[valid_coordinates == TRUE]
birth_points <- st_as_sf(
  births_coordinate_valid,
  coords = c("final_lon", "final_lat"),
  crs = 4326,
  remove = FALSE
)
```

## Assign county polygons and load county demographics

County is assigned from the supplied Florida polygon archive using the birth
residence coordinates. Points outside all 67 county polygons are not assigned
to the nearest county and are marked ineligible for exposure modeling. Annual
ACS five-year county demographics are retained as one-year-lagged sensitivity
fields; the primary outcome models use county fixed effects instead.

```{r counties}
county_archive <- here("data", "FL_county_export.zip")
county_data_dir <- here("data", "derived", "FL_county_export")
county_gpkg <- file.path(county_data_dir, "FL_counties_boundaries.gpkg")
county_demographic_file <- file.path(
  county_data_dir, "FL_county_demographics_2000_2022.csv"
)
stopifnot(file.exists(county_archive))

if (!file.exists(county_gpkg) || !file.exists(county_demographic_file)) {
  unzip(county_archive, exdir = here("data", "derived"))
}
stopifnot(file.exists(county_gpkg), file.exists(county_demographic_file))

county_polygons <- st_read(county_gpkg, quiet = TRUE) |>
  st_transform(4326) |>
  transmute(
    county_geoid = as.character(GEOID),
    county_name = as.character(NAME)
  )
stopifnot(nrow(county_polygons) == 67L, uniqueN(county_polygons$county_geoid) == 67L)

birth_points <- st_join(
  birth_points, county_polygons,
  join = st_within,
  left = TRUE
)
stopifnot(nrow(birth_points) == nrow(births_coordinate_valid),
          uniqueN(birth_points$ID) == nrow(birth_points))

county_key <- as.data.table(st_drop_geometry(birth_points))[, .(
  ID, county_geoid, county_name
)]
births[, `:=`(county_geoid = NA_character_, county_name = NA_character_)]
births[county_key, on = "ID", `:=`(
  county_geoid = i.county_geoid,
  county_name = i.county_name
)]
births[, valid_county := !is.na(county_geoid)]
births[, `:=`(
  exposure_eligible_T1 = valid_coordinates & valid_county & valid_T1_dates,
  exposure_eligible_T2 = valid_coordinates & valid_county & valid_T2_dates,
  exposure_eligible_T3 = valid_coordinates & valid_county & valid_T3_dates
)]
births[, exposure_eligible :=
         exposure_eligible_T1 | exposure_eligible_T2 | exposure_eligible_T3]

county_qa <- births[, .(
  n_births = .N,
  n_valid_coordinates = sum(valid_coordinates),
  n_county_matched = sum(valid_county),
  n_county_unmatched = sum(valid_coordinates & !valid_county),
  n_eligible_T1 = sum(exposure_eligible_T1),
  n_eligible_T2 = sum(exposure_eligible_T2),
  n_eligible_T3 = sum(exposure_eligible_T3)
)]
print(county_qa)
fwrite(
  county_qa,
  here("data", "derived", paste0("birth-county-assignment-qa", output_suffix, ".csv"))
)

# Only polygon-matched births enter the spatial storm assignment.
birth_points <- birth_points[!is.na(birth_points$county_geoid), ]

county_demographics <- fread(county_demographic_file)
county_demographics[, county_geoid := sprintf("%05d", as.integer(GEOID))]
county_demographics <- county_demographics[source == "acs5"]
stopifnot(
  county_demographics[, .N, by = .(county_geoid, year)][, all(N == 1L)]
)
```

## Read storm observations and calculate 55 kt radii

IBTrACS directly reports quadrant radii for 34, 50, and 64 kt winds. Within a
quadrant, the 55 kt radius is linearly interpolated between 50 and 64 kt when
both are available. If the 64 kt radius is unavailable, it is linearly
extrapolated inward from the 34-to-50 kt radial gradient. No RMW imputation or
storm-average radius is used.

```{r storms-read}
storm_points_raw <- st_read(ibtracs_shp, quiet = TRUE)
log_progress("storms", sprintf("read %s IBTrACS observations",
                                format(nrow(storm_points_raw), big.mark = ",")))

required_storm_columns <- c(
  "SID", "NAME", "ISO_TIME", "USA_WIND",
  paste0("USA_R34_", c("NE", "SE", "SW", "NW")),
  paste0("USA_R50_", c("NE", "SE", "SW", "NW")),
  paste0("USA_R64_", c("NE", "SE", "SW", "NW"))
)
missing_storm_columns <- setdiff(required_storm_columns, names(storm_points_raw))
stopifnot(length(missing_storm_columns) == 0L)

storm_points <- storm_points_raw |>
  select(all_of(required_storm_columns)) |>
  mutate(
    observation_time = ymd_hms(ISO_TIME, tz = "UTC", quiet = TRUE),
    observation_date = as.Date(observation_time),
    USA_WIND = as.numeric(USA_WIND)
  ) |>
  filter(
    observation_date >= study_start - days(window_days + 2L),
    observation_date <= study_end + days(window_days + 2L),
    USA_WIND >= 34
  )

radius_columns <- grep("^USA_R(34|50|64)_(NE|SE|SW|NW)$",
                       names(storm_points), value = TRUE)
for (column in radius_columns) {
  radius <- as.numeric(storm_points[[column]])
  radius[!is.finite(radius) | radius <= 0 | radius >= 999] <- NA_real_
  storm_points[[column]] <- radius
}

for (quadrant in c("NE", "SE", "SW", "NW")) {
  r34 <- storm_points[[paste0("USA_R34_", quadrant)]]
  r50 <- storm_points[[paste0("USA_R50_", quadrant)]]
  r64 <- storm_points[[paste0("USA_R64_", quadrant)]]

  r55_candidate <- ifelse(
    !is.na(r50) & !is.na(r64),
    r50 + (55 - 50) / (64 - 50) * (r64 - r50),
    ifelse(
      !is.na(r34) & !is.na(r50),
      pmax(0, r50 - (55 - 50) / (50 - 34) * (r34 - r50)),
      NA_real_
    )
  )
  # A 55 kt footprint cannot be more extensive than the observed 34 kt
  # footprint. Requiring R34 also guarantees that the sensitivity population
  # is nested within the primary exposure population.
  storm_points[[paste0("USA_R55_", quadrant)]] <- ifelse(
    is.na(r34) | is.na(r55_candidate), NA_real_, pmin(r34, r55_candidate)
  )
}

stopifnot(all(vapply(c("NE", "SE", "SW", "NW"), function(quadrant) {
  r34 <- storm_points[[paste0("USA_R34_", quadrant)]]
  r55 <- storm_points[[paste0("USA_R55_", quadrant)]]
  all(is.na(r55) | (!is.na(r34) & r55 <= r34))
}, logical(1))))

storm_radius_coverage <- rbindlist(lapply(c(34L, 55L), function(threshold) {
  columns <- paste0("USA_R", threshold, "_", c("NE", "SE", "SW", "NW"))
  data.table(
    threshold_kt = threshold,
    n_observations = nrow(storm_points),
    n_any_quadrant = sum(rowSums(!is.na(st_drop_geometry(storm_points)[columns])) > 0),
    n_all_quadrants = sum(rowSums(!is.na(st_drop_geometry(storm_points)[columns])) == 4)
  )
}))
storm_radius_coverage[, `:=`(
  pct_any_quadrant = 100 * n_any_quadrant / n_observations,
  pct_all_quadrants = 100 * n_all_quadrants / n_observations
)]
print(storm_radius_coverage)
```

## Construct storm-time-specific quadrant wind fields

Quadrants with missing radii remain missing; they are not silently treated as
zero-radius winds. The geometry uses a Florida-centered Albers equal-area CRS.
Each polygon retains its storm ID and observation timestamp.

```{r wind-field-functions}
analysis_crs <- 3086
nautical_mile_m <- 1852

sector_polygon <- function(center_xy, radius_m, bearing_start, bearing_end,
                           n = 36L) {
  if (!is.finite(radius_m) || radius_m <= 0) return(NULL)
  bearings <- seq(bearing_start, bearing_end, length.out = n + 1L)
  radians <- bearings * pi / 180
  arc <- cbind(
    center_xy[1] + radius_m * sin(radians),
    center_xy[2] + radius_m * cos(radians)
  )
  st_polygon(list(rbind(center_xy, arc, center_xy)))
}

make_wind_fields <- function(points_sf, threshold_kt) {
  projected <- st_transform(points_sf, analysis_crs)
  coordinates <- st_coordinates(projected)
  quadrant_spec <- data.table(
    quadrant = c("NE", "SE", "SW", "NW"),
    bearing_start = c(0, 90, 180, 270),
    bearing_end = c(90, 180, 270, 360)
  )

  geometries <- vector("list", nrow(projected))
  n_quadrants <- integer(nrow(projected))

  for (i in seq_len(nrow(projected))) {
    sectors <- vector("list", 0L)
    for (q in seq_len(nrow(quadrant_spec))) {
      radius_nm <- projected[[paste0(
        "USA_R", threshold_kt, "_", quadrant_spec$quadrant[q]
      )]][i]
      sector <- sector_polygon(
        coordinates[i, ], radius_nm * nautical_mile_m,
        quadrant_spec$bearing_start[q], quadrant_spec$bearing_end[q]
      )
      if (!is.null(sector)) sectors[[length(sectors) + 1L]] <- sector
    }

    n_quadrants[i] <- length(sectors)
    if (length(sectors) > 0L) {
      geometries[[i]] <- st_union(st_sfc(sectors, crs = analysis_crs))[[1]]
    } else {
      geometries[[i]] <- st_geometrycollection()
    }
  }

  st_sf(
    st_drop_geometry(projected) |>
      transmute(
        SID, NAME, observation_time, observation_date, USA_WIND,
        threshold_kt = as.integer(threshold_kt),
        observed_quadrants = n_quadrants
      ),
    geometry = st_sfc(geometries, crs = analysis_crs)
  ) |>
    filter(observed_quadrants > 0L, !st_is_empty(geometry))
}
```

Only observations whose coarse maximum-radius buffer could reach the study
residences are passed to the more expensive quadrant construction.

```{r wind-fields-build}
birth_extent <- st_bbox(st_transform(birth_points, analysis_crs))
birth_extent_polygon <- st_as_sfc(birth_extent) |> st_buffer(250 * 1000)

storm_points_projected <- st_transform(storm_points, analysis_crs)
near_florida <- lengths(st_intersects(storm_points_projected,
                                      birth_extent_polygon)) > 0L
storm_points_near <- storm_points[near_florida, ]

wind_fields_34 <- make_wind_fields(storm_points_near, 34L)
log_progress("wind fields", sprintf("built %s primary 34 kt observations",
                                     format(nrow(wind_fields_34), big.mark = ",")))
wind_fields_55 <- make_wind_fields(
  storm_points_near |> filter(USA_WIND >= 55), 55L
)
log_progress("wind fields", sprintf("built %s sensitivity 55 kt observations",
                                     format(nrow(wind_fields_55), big.mark = ",")))
wind_fields <- rbind(wind_fields_34, wind_fields_55) |>
  st_make_valid()

stopifnot(
  all(wind_fields$threshold_kt %in% c(34L, 55L)),
  !anyNA(wind_fields$SID),
  !anyNA(wind_fields$observation_time),
  all(!st_is_empty(wind_fields))
)
```

## Assign direct hazard and local impact time

The spatial join is performed against storm-time-specific polygons. Within
each birth, storm, and threshold, local impact is the closest storm-center
observation among the observations whose wind field contains the residence.

```{r exposure-assign}
storm_centers <- storm_points_near |>
  st_transform(analysis_crs) |>
  select(SID, observation_time, center_geometry = geometry)
storm_center_key <- paste(storm_centers$SID, storm_centers$observation_time)

assign_exposure_chunk <- function(index) {
  births_chunk <- st_transform(birth_points[index, ], analysis_crs) |>
    select(ID, T1_start, DATE_OF_BIRTH)

  # The birth file is ordered chronologically. Restricting the candidate wind
  # fields to the dates represented in each chunk prevents a spatial match to
  # every Florida storm in the full study period. The row-level filter below
  # then enforces the exact pregnancy-specific risk interval.
  candidate_start <- min(births_chunk$T1_start, na.rm = TRUE) - window_days
  candidate_end <- max(births_chunk$DATE_OF_BIRTH, na.rm = TRUE) + window_days
  candidate_fields <- wind_fields[
    wind_fields$observation_date >= as.Date(candidate_start) &
      wind_fields$observation_date <= as.Date(candidate_end),
  ]
  if (nrow(candidate_fields) == 0L) return(NULL)

  matched <- st_join(
    births_chunk,
    candidate_fields |>
      select(SID, NAME, observation_time, observation_date, USA_WIND,
             threshold_kt, observed_quadrants),
    join = st_within,
    left = FALSE
  )
  if (nrow(matched) == 0L) return(NULL)

  matched <- matched[
    as.IDate(matched$observation_date) >= matched$T1_start - window_days &
      as.IDate(matched$observation_date) <= matched$DATE_OF_BIRTH + window_days,
  ]
  if (nrow(matched) == 0L) return(NULL)

  matched_key <- paste(matched$SID, matched$observation_time)
  center_index <- match(matched_key, storm_center_key)
  stopifnot(!anyNA(center_index))
  matched$center_distance_km <- as.numeric(
    st_distance(matched, storm_centers[center_index, ], by_element = TRUE)
  ) / 1000

  matched_dt <- matched |>
    st_drop_geometry() |>
    as.data.table()
  setorder(
    matched_dt, ID, SID, threshold_kt, center_distance_km,
    -USA_WIND, observation_time
  )
  matched_dt[, .SD[1L], by = .(ID, SID, threshold_kt)]
}

chunk_starts <- seq.int(1L, nrow(birth_points), by = chunk_size)
exposure_chunks <- vector("list", length(chunk_starts))
assignment_started <- Sys.time()
log_progress(
  "birth assignment",
  sprintf(
    "starting %s eligible births in %s chunks",
    format(nrow(birth_points), big.mark = ","),
    format(length(chunk_starts), big.mark = ",")
  )
)
for (j in seq_along(chunk_starts)) {
  chunk_started <- Sys.time()
  index <- chunk_starts[j]:min(
    chunk_starts[j] + chunk_size - 1L,
    nrow(birth_points)
  )
  exposure_chunks[[j]] <- assign_exposure_chunk(index)
  chunk_seconds <- as.numeric(difftime(Sys.time(), chunk_started, units = "secs"))
  elapsed_seconds <- as.numeric(
    difftime(Sys.time(), assignment_started, units = "secs")
  )
  completed_births <- max(index)
  completion_fraction <- j / length(chunk_starts)
  eta_seconds <- elapsed_seconds / completion_fraction - elapsed_seconds

  if (j == 1L || j %% progress_every == 0L || j == length(chunk_starts)) {
    log_progress(
      "birth assignment",
      sprintf(
        paste0(
          "chunk %s/%s | %s/%s births | %.1f%% | chunk %s | ",
          "elapsed %s | ETA %s | matched event rows %s"
        ),
        j, length(chunk_starts),
        format(completed_births, big.mark = ","),
        format(nrow(birth_points), big.mark = ","),
        100 * completion_fraction,
        format_duration(chunk_seconds),
        format_duration(elapsed_seconds),
        format_duration(eta_seconds),
        format(
          if (is.null(exposure_chunks[[j]])) 0L else nrow(exposure_chunks[[j]]),
          big.mark = ","
        )
      )
    )
  }
  gc()
}

birth_storm_events <- rbindlist(exposure_chunks, use.names = TRUE, fill = TRUE)

# Define local impact once from the primary 34 kt field. The 55 kt sensitivity
# indicates whether the residence also entered the stronger wind field, but it
# uses the same primary impact time. Threshold-specific impact times can differ
# by one or more IBTrACS observations and create artificial non-nesting at a
# trimester boundary.
primary_anchors <- birth_storm_events[
  threshold_kt == 34L,
  .(
    ID, SID,
    primary_observation_time = observation_time,
    primary_observation_date = observation_date,
    primary_center_distance_km = center_distance_km
  )
]

sensitivity_events <- merge(
  birth_storm_events[threshold_kt == 55L],
  primary_anchors,
  by = c("ID", "SID"),
  all.x = TRUE,
  sort = FALSE
)
stopifnot(
  !anyNA(sensitivity_events$primary_observation_time),
  !anyNA(sensitivity_events$primary_observation_date)
)
sensitivity_events[, `:=`(
  observation_time = primary_observation_time,
  observation_date = primary_observation_date,
  center_distance_km = primary_center_distance_km,
  primary_observation_time = NULL,
  primary_observation_date = NULL,
  primary_center_distance_km = NULL
)]

birth_storm_events <- rbindlist(
  list(
    birth_storm_events[threshold_kt == 34L],
    sensitivity_events
  ),
  use.names = TRUE,
  fill = TRUE
)
birth_storm_events[, `:=`(
  local_impact_date = as.IDate(observation_date),
  window_start = as.IDate(observation_date) - window_days,
  window_end = as.IDate(observation_date) + window_days
)]

stopifnot(
  birth_storm_events[, .N, by = .(ID, SID, threshold_kt)][, all(N == 1L)],
  all(birth_storm_events$window_start <= birth_storm_events$local_impact_date),
  all(birth_storm_events$local_impact_date <= birth_storm_events$window_end)
)
```

## Assign each event window to gestational periods

An event is assigned to a trimester when its residence-specific ±7-day window
overlaps that trimester. `acute_direct_*` separately indicates that the local
impact date itself occurred during the trimester.

```{r trimester-assign}
gestation_intervals <- rbindlist(list(
  births[exposure_eligible_T1 == TRUE, .(
    ID, trimester = "T1", period_start = T1_start, period_end = T1_end
  )],
  births[exposure_eligible_T2 == TRUE, .(
    ID, trimester = "T2", period_start = T2_start, period_end = T2_end
  )],
  births[exposure_eligible_T3 == TRUE, .(
    ID, trimester = "T3", period_start = T3_start, period_end = T3_end
  )]
))

stopifnot(
  !anyNA(gestation_intervals$period_start),
  !anyNA(gestation_intervals$period_end),
  all(gestation_intervals$period_start <= gestation_intervals$period_end),
  gestation_intervals[, .N, by = .(ID, trimester)][, all(N == 1L)]
)

event_windows <- merge(
  birth_storm_events,
  gestation_intervals,
  by = "ID",
  allow.cartesian = TRUE
)[window_start <= period_end & window_end >= period_start]

event_windows[, acute_direct :=
                local_impact_date >= period_start &
                local_impact_date <= period_end]

event_summary <- event_windows[, .(
  exposed_window = 1L,
  acute_direct = as.integer(any(acute_direct)),
  n_storms = uniqueN(SID),
  max_storm_wind_kt = max(USA_WIND, na.rm = TRUE),
  first_window_date = min(window_start),
  last_window_date = max(window_end)
), by = .(ID, trimester, threshold_kt)]

metric_columns <- c(
  "exposed_window", "acute_direct", "n_storms", "max_storm_wind_kt",
  "first_window_date", "last_window_date"
)
exposure_wide <- dcast(
  event_summary,
  ID ~ trimester + threshold_kt,
  value.var = metric_columns,
  sep = "_"
)

# Do not fill absent cells here. A birth represented in one trimester can be
# structurally ineligible for a later trimester. Eligible missing cells are
# set to zero only after trimester eligibility is joined below; ineligible
# cells must remain NA.
```

## Merge by birth ID and add fixed-effect variables

This is a keyed merge, not an ordered column bind. `birth_year_fe` and
`conception_season_fe` are model-ready fixed effects. Calendar month is also
provided for a stricter seasonality specification.

```{r analytic-merge}
# Approved analytic columns sent by Jessica Broach on April 14, 2025. The
# corrected exposure variables, eligibility flag, and temporal fixed-effect
# variables are appended later in this workflow.
approved_birth_columns <- c(
  "ID", "final_lat", "final_lon", "ALCOHOL_DRINKS", "ALCOHOL_USE",
  "BIRTH_WEIGHT_GRAMS", "CalculatedGestationalAge", "DATE_OF_BIRTH",
  "GESTATION_WEEKS", "FATHER_AGE", "FATHER_EDCODE",
  "Father_CalculatedHisp", "Father_CalculatedRace", "MENSES_DATE",
  "MOTHER_AGE", "Mother_CalculatedHisp", "Mother_CalculatedRace",
  "MOTHER_EDCODE", "MOTHER_MARRIED", "MOTHER_RES_COUNTY",
  "MOTHER_WIC_YESNO", "MR_DIAB", "MR_DIAB_GEST", "MR_HYPERT_CHRONIC",
  "MR_HYPERT_ECLAMPSIA", "MR_HYPERT_PREG", "MR_PREV_PRETERM", "PLURALITY_CODE",
  "PrePregnancy_BMI", "PRINCIPAL_SOURCE_PAY", "PRINCIPAL_SRCPAY_CODE",
  "SEX", "m28_temperature", "lat_pm25", "lon_pm25", "date_pm25",
  "loc_data_pm25", "m28_pm25", "m21_pm25", "m14_pm25", "m7_pm25",
  "dday_pm25", "p7_pm25", "p14_pm25", "p21_pm25", "p28_pm25",
  "m20_pm25", "m19_pm25", "m18_pm25", "m17_pm25", "m16_pm25",
  "m15_pm25", "m14_pm25.1", "m13_pm25", "m12_pm25", "m11_pm25",
  "m10_pm25", "m9_pm25", "m8_pm25", "m7_pm25.1", "m6_pm25",
  "m5_pm25", "m4_pm25", "m3_pm25", "m2_pm25", "m1_pm25", "m0_pm25",
  "year_pm25r", "d_1_13_wk_90_1", "d_14_28_wk_90_1",
  "d_from29_wk_90_1", "d_1_13_wk_95_1", "d_14_28_wk_95_1",
  "d_from29_wk_95_1", "d_1_13_wk_98_1", "d_14_28_wk_98_1",
  "d_from29_wk_98_1", "d_1_13_wk_90_2", "d_14_28_wk_90_2",
  "d_from29_wk_90_2", "d_1_13_wk_95_2", "d_14_28_wk_95_2",
  "d_from29_wk_95_2", "d_1_13_wk_98_2", "d_14_28_wk_98_2",
  "d_from29_wk_98_2", "d_1_13_wk_90_3", "d_14_28_wk_90_3",
  "d_from29_wk_90_3", "d_1_13_wk_95_3", "d_14_28_wk_95_3",
  "d_from29_wk_95_3", "d_1_13_wk_98_3", "d_14_28_wk_98_3",
  "d_from29_wk_98_3", "trimester1_length", "trimester2_length",
  "trimester3_length"
)

all_files <- list.files(
  here("data", "21_definition_final_re"),
  pattern = "[.]Rdata$", full.names = TRUE
)
stopifnot(length(all_files) > 0L)

load_one_object <- function(path) {
  environment <- new.env(parent = emptyenv())
  object_names <- load(path, envir = environment)
  stopifnot(length(object_names) == 1L)
  yearly_births <- as.data.table(environment[[object_names]])
  available_columns <- intersect(approved_birth_columns, names(yearly_births))
  yearly_births[, ..available_columns]
}

# Select within each yearly object before row-binding. Loading all source
# columns for every year before subsetting can exceed R's vector-memory limit.
analytic_births <- rbindlist(
  lapply(all_files, load_one_object), use.names = TRUE, fill = TRUE
)
stopifnot(!anyNA(analytic_births$ID), uniqueN(analytic_births$ID) == nrow(analytic_births))

missing_approved_columns <- setdiff(approved_birth_columns, names(analytic_births))
if (length(missing_approved_columns) > 0L) {
  stop(
    "Approved birth columns missing from the yearly files: ",
    paste(missing_approved_columns, collapse = ", "),
    call. = FALSE
  )
}
analytic_births <- analytic_births[, ..approved_birth_columns]
log_progress(
  "analytic merge",
  sprintf("retained %s approved/model-required source columns",
          format(length(approved_birth_columns), big.mark = ","))
)

parse_mixed_idate <- function(x) {
  # Year-specific files can store dates as POSIXct, Date, IDate, or character.
  # Normalize through an ISO date string and suppress low-level POSIX warnings;
  # unparseable values remain NA and are counted below.
  if (inherits(x, "POSIXt")) {
    text <- suppressWarnings(format(x, "%Y-%m-%d", tz = "UTC"))
  } else {
    text <- substr(as.character(x), 1L, 10L)
  }
  suppressWarnings(as.IDate(text, format = "%Y-%m-%d"))
}

analytic_births[, DATE_OF_BIRTH := parse_mixed_idate(DATE_OF_BIRTH)]
missing_birth_dates <- analytic_births[is.na(DATE_OF_BIRTH), .N]
log_progress(
  "analytic merge",
  sprintf("%s records have an unparseable birth date and will be excluded",
          format(missing_birth_dates, big.mark = ","))
)
analytic_births <- analytic_births[
  !is.na(DATE_OF_BIRTH) &
  DATE_OF_BIRTH >= as.IDate(study_start) &
    DATE_OF_BIRTH <= as.IDate(study_end)
]
analytic_births[, conception_date := parse_mixed_idate(MENSES_DATE)]
if ("GESTATION_WEEKS" %in% names(analytic_births)) {
  analytic_births[
    is.na(conception_date) & is.finite(GESTATION_WEEKS),
    conception_date := DATE_OF_BIRTH - as.integer(round(GESTATION_WEEKS * 7))
  ]
}

analytic_births[, `:=`(
  birth_year = year(DATE_OF_BIRTH),
  birth_month = month(DATE_OF_BIRTH),
  conception_year = year(conception_date),
  conception_month = month(conception_date)
)]
analytic_births[, conception_season := fcase(
  conception_month %in% c(12L, 1L, 2L), "DJF",
  conception_month %in% 3:5, "MAM",
  conception_month %in% 6:8, "JJA",
  conception_month %in% 9:11, "SON",
  default = NA_character_
)]

birth_exposure_eligibility <- births[, .(
  ID,
  exposure_eligible,
  exposure_eligible_T1,
  exposure_eligible_T2,
  exposure_eligible_T3,
  county_geoid,
  county_name
)]

analytic_corrected <- merge(
  analytic_births,
  birth_exposure_eligibility,
  by = "ID",
  all.x = TRUE,
  sort = FALSE
)

# Attach prior-year ACS five-year county measures for optional sensitivity
# analyses. Coverage begins with 2010 births because ACS5 starts in 2009.
lagged_county_demographics <- county_demographics[, .(
  county_geoid,
  birth_year = year + 1L,
  county_demo_year_lag1 = year,
  county_total_pop_lag1 = total_pop,
  county_median_age_lag1 = median_age,
  county_poverty_rate_lag1 = poverty_rate_pct,
  county_median_hh_income_lag1 = median_hh_income,
  county_median_home_value_lag1 = median_home_value,
  county_pct_white_lag1 = 100 * race_white / total_pop,
  county_pct_black_lag1 = 100 * race_black / total_pop,
  county_pct_asian_lag1 = 100 * race_asian / total_pop,
  county_pct_hispanic_lag1 = 100 * hispanic_any_race / total_pop
)]
stopifnot(
  lagged_county_demographics[, .N, by = .(county_geoid, birth_year)][,
    all(N == 1L)]
)

analytic_corrected <- merge(
  analytic_corrected,
  lagged_county_demographics,
  by = c("county_geoid", "birth_year"),
  all.x = TRUE,
  sort = FALSE
)
eligibility_columns <- c(
  "exposure_eligible", "exposure_eligible_T1",
  "exposure_eligible_T2", "exposure_eligible_T3"
)
for (column in eligibility_columns) {
  analytic_corrected[is.na(get(column)), (column) := FALSE]
}

analytic_corrected <- merge(
  analytic_corrected,
  exposure_wide,
  by = "ID",
  all.x = TRUE,
  sort = FALSE
)

expected_exposure_columns <- as.vector(outer(
  c("exposed_window", "acute_direct", "n_storms"),
  as.vector(outer(c("T1", "T2", "T3"), c(34L, 55L), paste, sep = "_")),
  paste, sep = "_"
))
for (column in expected_exposure_columns) {
  if (!column %in% names(analytic_corrected)) {
    analytic_corrected[, (column) := NA_integer_]
  }
  trimester <- sub(
    "^(exposed_window|acute_direct|n_storms)_(T[123])_.*$",
    "\\2", column
  )
  eligibility_column <- paste0("exposure_eligible_", trimester)
  set(
    analytic_corrected,
    which(analytic_corrected[[eligibility_column]] &
            is.na(analytic_corrected[[column]])),
    column,
    0L
  )
}

analytic_corrected[, `:=`(
  birth_year_fe = factor(birth_year),
  conception_season_fe = factor(
    conception_season, levels = c("DJF", "MAM", "JJA", "SON")
  ),
  birth_month_fe = factor(birth_month)
)]

stopifnot(
  nrow(analytic_corrected) == nrow(analytic_births),
  uniqueN(analytic_corrected$ID) == nrow(analytic_corrected),
  all(analytic_corrected$ID %in% analytic_births$ID)
)

for (trimester in c("T1", "T2", "T3")) {
  primary <- analytic_corrected[[paste0("exposed_window_", trimester, "_34")]]
  sensitivity <- analytic_corrected[[paste0("exposed_window_", trimester, "_55")]]
  eligible <- analytic_corrected[[paste0("exposure_eligible_", trimester)]]
  stopifnot(
    all(!eligible | (!is.na(primary) & !is.na(sensitivity))),
    all(eligible | (is.na(primary) & is.na(sensitivity))),
    all(is.na(sensitivity) | is.na(primary) | sensitivity <= primary)
  )
}
```

## Validation outputs

```{r validation}
exposure_prevalence <- melt(
  analytic_corrected,
  id.vars = c("ID", "birth_year"),
  measure.vars = patterns(exposed = "^exposed_window_T[123]_(34|55)$"),
  variable.name = "exposure_variable",
  value.name = "exposed"
)[, .(
  eligible_births = sum(!is.na(exposed)),
  exposed_births = sum(exposed, na.rm = TRUE),
  prevalence = mean(exposed, na.rm = TRUE)
), by = .(birth_year, exposure_variable)]

storm_event_qa <- event_windows[, .(
  exposed_births = uniqueN(ID),
  local_impact_start = min(local_impact_date),
  local_impact_end = max(local_impact_date),
  median_center_distance_km = median(center_distance_km, na.rm = TRUE),
  min_observed_quadrants = min(observed_quadrants),
  max_observed_quadrants = max(observed_quadrants)
), by = .(SID, NAME, threshold_kt, trimester)]

county_demographic_coverage <- analytic_corrected[, .(
  births = .N,
  county_matched = sum(!is.na(county_geoid)),
  lagged_acs5_available = sum(!is.na(county_demo_year_lag1)),
  lagged_acs5_coverage = mean(!is.na(county_demo_year_lag1))
), by = birth_year]

trimester_eligibility_qa <- rbindlist(lapply(c("T1", "T2", "T3"), function(trimester) {
  eligible_column <- paste0("exposure_eligible_", trimester)
  exposure_column <- paste0("exposed_window_", trimester, "_34")
  analytic_corrected[, .(
    trimester = trimester,
    births = .N,
    eligible = sum(get(eligible_column)),
    exposed = sum(get(exposure_column) == 1L, na.rm = TRUE),
    ineligible = sum(!get(eligible_column)),
    PTB_source = sum(
      suppressWarnings(as.numeric(GESTATION_WEEKS)) < 37 &
        suppressWarnings(as.integer(PLURALITY_CODE)) == 1L,
      na.rm = TRUE
    ),
    PTB_eligible = sum(
      get(eligible_column) &
        suppressWarnings(as.numeric(GESTATION_WEEKS)) < 37 &
        suppressWarnings(as.integer(PLURALITY_CODE)) == 1L,
      na.rm = TRUE
    ),
    vPTB_source = sum(
      suppressWarnings(as.numeric(GESTATION_WEEKS)) < 32 &
        suppressWarnings(as.integer(PLURALITY_CODE)) == 1L,
      na.rm = TRUE
    ),
    vPTB_eligible = sum(
      get(eligible_column) &
        suppressWarnings(as.numeric(GESTATION_WEEKS)) < 32 &
        suppressWarnings(as.integer(PLURALITY_CODE)) == 1L,
      na.rm = TRUE
    )
  )]
}))

print(exposure_prevalence)
print(storm_event_qa[order(-exposed_births)][1:min(.N, 30L)])
print(county_demographic_coverage)
print(trimester_eligibility_qa)
fwrite(
  county_demographic_coverage,
  here(
    "data", "derived",
    paste0("birth-county-demographic-coverage", output_suffix, ".csv")
  )
)
fwrite(
  trimester_eligibility_qa,
  here(
    "data", "derived",
    paste0("birth-trimester-eligibility-qa", output_suffix, ".csv")
  )
)

prevalence_plot <- ggplot(
  exposure_prevalence,
  aes(birth_year, prevalence, color = exposure_variable)
) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Birth year", y = "Exposure prevalence",
    color = "Trimester / threshold",
    title = "Corrected residence-based TC exposure prevalence"
  ) +
  theme_minimal()
print(prevalence_plot)

ggsave(
  here(
    "figs", "exposure-validation",
    paste0("exposure-prevalence-by-year", output_suffix, ".png")
  ),
  prevalence_plot, width = 10, height = 6, dpi = 300
)

stopifnot(
  all(exposure_prevalence$prevalence >= 0 &
        exposure_prevalence$prevalence <= 1),
  all(event_windows$threshold_kt %in% c(34L, 55L)),
  all(event_windows$window_start <= event_windows$period_end),
  all(event_windows$window_end >= event_windows$period_start)
)
```

## Save corrected birth data

The event-level file is retained for auditability. The analytic file contains
one row per birth and is the input for refitting all birth-outcome models.

```{r write-output}
if (isTRUE(params$write_outputs)) {
  fwrite(
    birth_storm_events,
    here(
      "data", "derived",
      paste0("birth-storm-events-time-specific", output_suffix, ".csv")
    )
  )
  fwrite(
    storm_radius_coverage,
    here(
      "data", "derived",
      paste0("storm-wind-radius-coverage", output_suffix, ".csv")
    )
  )
  fwrite(
    storm_event_qa,
    here(
      "data", "derived",
      paste0("birth-storm-exposure-validation", output_suffix, ".csv")
    )
  )
  saveRDS(
    analytic_corrected,
    here(
      "data", "outputs",
      paste0("All_Births_Corrected_Storm_Exposure", output_suffix, ".rds")
    ),
    compress = "xz"
  )
}
log_progress("complete", "corrected birth exposure workflow finished")
```

## Required model specification

All previously reported birth-outcome models must be refit. The 34 kt
`exposed_window` variable is primary; the corresponding 55 kt variable is the
sensitivity exposure. Every model must include conception year-month and
residence-county fixed effects, with county-clustered standard errors.
The existing substantive covariates should be added to `covariates` after their
coding and missing-data strategy are finalized. `HO-storms-models.Rmd` is the
authoritative implementation.

```{r model-template, eval=FALSE}
covariates <- c(
  "Mother_CalculatedRace", "Mother_CalculatedHisp", "MOTHER_EDCODE",
  "Father_CalculatedRace", "Father_CalculatedHisp", "FATHER_EDCODE"
)

fit_birth_model <- function(data, outcome, trimester = "T1",
                            threshold_kt = 34L) {
  exposure <- paste0("exposed_window_", trimester, "_", threshold_kt)
  eligibility <- paste0("exposure_eligible_", trimester)
  data <- data[get(eligibility) %in% TRUE]
  model_formula <- as.formula(paste0(
    outcome, " ~ ", paste(c(exposure, covariates), collapse = " + "),
    " | conception_year_month + county_geoid"
  ))
  fixest::feglm(
    model_formula, data = data, family = "binomial",
    vcov = ~county_geoid
  )
}

ptb_t1_primary <- fit_birth_model(
  analytic_corrected, outcome = "PTB", trimester = "T1", threshold_kt = 34L
)
ptb_t1_sensitivity <- fit_birth_model(
  analytic_corrected, outcome = "PTB", trimester = "T1", threshold_kt = 55L
)
```

## Interpretation constraints

1. The ±7-day variable represents a direct-wind event window around local
   impact, not 15 days of continuous wind exposure.
2. The 55 kt footprint is an empirical interpolation and must be labeled as a
   sensitivity analysis.
3. Missing wind radii can cause under-ascertainment; report the coverage table
   by threshold and study period.
4. Residence at delivery may differ from residence or actual location at
   impact. Evacuation and out-migration remain unmeasured.
5. Rainfall, flooding, storm surge, outages, and healthcare disruption are not
   represented by this direct-wind measure.
