---
title: "Refit of Florida birth-outcome models using corrected tropical cyclone exposure"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  corrected_data: "data/outputs/All_Births_Corrected_Storm_Exposure.rds"
  imputations: 5
  mice_iterations: 5
  polyreg_max_weights: 10000
  imputation_training_rows: 250000
  seed: 20260810
  fit_55kt_sensitivity: true
  fit_complete_case_sensitivity: true
  save_imputation_object: false
---

## Purpose and reconstructed manuscript model

This workflow refits the birth-outcome models described in `Final
Manuscript.pdf` using the corrected storm-time-specific exposure dataset. It
does not use the superseded threat, impact, or cleanup variables.

The manuscript describes:

1. Binary logistic regression with adjusted odds ratios (aOR) and 95%
   confidence intervals.
2. Outcomes: preterm birth (PTB), very preterm birth (vPTB), term low
   birthweight (tLBW), low birthweight (LBW), very low birthweight (vLBW),
   gestational diabetes (GDM), and hypertensive disorders of pregnancy (HDP).
3. Separate exposure models by trimester. PTB, vPTB, LBW, and vLBW were not
   modeled for the third trimester; tLBW was modeled in all three trimesters;
   GDM and HDP used first-trimester exposure only.
4. Adjustment for maternal and paternal race, ethnicity, and education;
   maternal age; categorical pre-pregnancy BMI; previous preterm infant; WIC
   status; and principal source of payment.
5. Multiple imputation by chained equations (MICE) for missing covariates.
   The manuscript reports five iterations but does not report the number of
   imputed datasets or variable-specific imputation methods.

For reproducibility, this refit uses `r params$imputations` imputed datasets and
`r params$mice_iterations` MICE iterations. In response to the reviewer concern
about temporal confounding, every corrected primary model includes birth-year
and conception-season fixed effects. The 34 kt ±7-day exposure is primary; the
55 kt nested exposure is a sensitivity analysis.

Alcohol was listed among measured covariates in the manuscript's Covariates
section but was not included in the Statistical Analysis adjustment list. It
is therefore not included in the primary refit.

## Setup

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("data.table", "fixest", "ggplot2", "mice", "scales")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0L) {
  stop(
    "Install required packages before running: ",
    paste(missing_packages, collapse = ", "),
    ". For example: install.packages(c(",
    paste(sprintf("'%s'", missing_packages), collapse = ", "), "))",
    call. = FALSE
  )
}

library(data.table)
library(fixest)
library(ggplot2)

set.seed(as.integer(params$seed))
dir.create("data/outputs/Models", recursive = TRUE, showWarnings = FALSE)
dir.create("figs/model-comparison", recursive = TRUE, showWarnings = FALSE)

progress_file <- "data/outputs/Models/model-refit-progress.log"
if (file.exists(progress_file)) file.remove(progress_file)

log_progress <- function(stage, detail) {
  line <- sprintf(
    "%s | %-22s | %s",
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), stage, detail
  )
  message(line)
  write(line, progress_file, append = TRUE)
  invisible(line)
}

format_duration <- function(seconds) {
  seconds <- max(0, round(seconds))
  sprintf("%02d:%02d:%02d", seconds %/% 3600,
          (seconds %% 3600) %/% 60, seconds %% 60)
}

log_progress("setup", "model refit started")
```

## Read only model-required columns

An RDS must be read in full, but unnecessary columns are removed immediately
to reduce working memory.

```{r read-data}
stopifnot(file.exists(params$corrected_data))
births <- readRDS(params$corrected_data)
setDT(births)

source_columns <- c(
  "ID", "BIRTH_WEIGHT_GRAMS", "GESTATION_WEEKS", "PLURALITY_CODE",
  "MR_DIAB", "MR_DIAB_GEST", "MR_HYPERT_CHRONIC",
  "MR_HYPERT_ECLAMPSIA", "MR_HYPERT_PREG",
  "Mother_CalculatedRace", "Mother_CalculatedHisp", "MOTHER_EDCODE",
  "Father_CalculatedRace", "Father_CalculatedHisp", "FATHER_EDCODE",
  "MOTHER_AGE", "PrePregnancy_BMI", "MR_PREV_PRETERM",
  "MOTHER_WIC_YESNO", "PRINCIPAL_SRCPAY_CODE",
  "birth_year", "conception_season", "exposure_eligible"
)
exposure_columns <- as.vector(outer(
  paste0("exposed_window_T", 1:3), c(34L, 55L), paste, sep = "_"
))
required_columns <- c(source_columns, exposure_columns)
missing_columns <- setdiff(required_columns, names(births))
if (length(missing_columns) > 0L) {
  stop(
    "Corrected dataset is missing model-required columns: ",
    paste(missing_columns, collapse = ", "),
    ". Rerun HO-storms-exposure.Rmd after its latest update.",
    call. = FALSE
  )
}

births <- births[, ..required_columns]
gc()
log_progress(
  "read data",
  sprintf("loaded %s births and retained %s columns",
          format(nrow(births), big.mark = ","), ncol(births))
)
```

## Recode outcomes, eligibility, and covariates

```{r recode}
normalize_missing_code <- function(x, missing = c("", "NULL", "NA", "N/A", "U", "UNK", "UNKNOWN")) {
  value <- trimws(toupper(as.character(x)))
  value[value %in% missing] <- NA_character_
  value
}

binary_yes_no <- function(x) {
  value <- normalize_missing_code(x)
  fcase(
    value %in% c("Y", "YES", "1", "TRUE"), 1L,
    value %in% c("N", "NO", "0", "FALSE"), 0L,
    default = NA_integer_
  )
}

factor_missing_codes <- function(x, additional_missing = character()) {
  value <- normalize_missing_code(
    x, missing = c("", "NULL", "NA", "N/A", "U", "UNK", "UNKNOWN",
                   additional_missing)
  )
  factor(value)
}

collapse_race <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 1L, "White",
    value == 2L, "Black",
    !is.na(value) & value != 99L, "Other",
    default = NA_character_
  ), levels = c("White", "Black", "Other"))
}

collapse_ethnicity <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 0L, "Non-Hispanic",
    value %in% 1:98, "Hispanic",
    default = NA_character_
  ), levels = c("Non-Hispanic", "Hispanic"))
}

collapse_education <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value %in% 1:2, "Less than high school",
    value == 3L, "High school or GED",
    value %in% 4:5, "Some college or associate degree",
    value %in% 6:8, "Bachelor degree or higher",
    default = NA_character_
  ), levels = c(
    "Less than high school", "High school or GED",
    "Some college or associate degree", "Bachelor degree or higher"
  ))
}

collapse_payment <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 1L, "Medicaid",
    value == 2L, "Private insurance",
    value == 3L, "Self-pay",
    !is.na(value) & value != 9L, "Other",
    default = NA_character_
  ), levels = c("Medicaid", "Private insurance", "Self-pay", "Other"))
}

births[, `:=`(
  gest_weeks = suppressWarnings(as.numeric(GESTATION_WEEKS)),
  birthweight_g = suppressWarnings(as.numeric(BIRTH_WEIGHT_GRAMS)),
  singleton = suppressWarnings(as.integer(PLURALITY_CODE)) == 1L,
  maternal_race = collapse_race(Mother_CalculatedRace),
  maternal_ethnicity = collapse_ethnicity(Mother_CalculatedHisp),
  maternal_education = collapse_education(MOTHER_EDCODE),
  paternal_race = collapse_race(Father_CalculatedRace),
  paternal_ethnicity = collapse_ethnicity(Father_CalculatedHisp),
  paternal_education = collapse_education(FATHER_EDCODE),
  maternal_age = suppressWarnings(as.numeric(MOTHER_AGE)),
  previous_preterm = factor(binary_yes_no(MR_PREV_PRETERM), levels = 0:1),
  wic = factor(binary_yes_no(MOTHER_WIC_YESNO), levels = 0:1),
  payment = collapse_payment(PRINCIPAL_SRCPAY_CODE),
  preexisting_diabetes = binary_yes_no(MR_DIAB),
  gestational_diabetes_raw = binary_yes_no(MR_DIAB_GEST),
  chronic_hypertension = binary_yes_no(MR_HYPERT_CHRONIC),
  gestational_hypertension = binary_yes_no(MR_HYPERT_PREG),
  eclampsia = binary_yes_no(MR_HYPERT_ECLAMPSIA)
)]

bmi_numeric <- suppressWarnings(as.numeric(normalize_missing_code(
  births$PrePregnancy_BMI
)))
births[, bmi_category := cut(
  bmi_numeric,
  breaks = c(-Inf, 18.5, 25, 30, Inf),
  right = FALSE,
  labels = c("Underweight", "Normal", "Overweight", "Obesity")
)]

births[, hdp_raw := fcase(
  gestational_hypertension == 1L | eclampsia == 1L, 1L,
  gestational_hypertension == 0L & eclampsia == 0L, 0L,
  default = NA_integer_
)]

births[, `:=`(
  PTB = as.integer(gest_weeks < 37),
  vPTB = as.integer(gest_weeks < 32),
  LBW = as.integer(birthweight_g < 2500),
  vLBW = as.integer(birthweight_g < 1500),
  tLBW = as.integer(gest_weeks >= 37 & birthweight_g < 2500),
  GDM = gestational_diabetes_raw,
  HDP = hdp_raw,
  birth_year = suppressWarnings(as.integer(birth_year)),
  conception_season = factor(
    conception_season, levels = c("DJF", "MAM", "JJA", "SON")
  )
)]

# Manuscript base cohort: accurate coordinates, singleton birth, and nonmissing
# gestational weeks and birthweight. Year/season are additionally required by
# the corrected specification.
model_births <- births[
  exposure_eligible %in% TRUE & singleton %in% TRUE &
    !is.na(gest_weeks) & !is.na(birthweight_g) &
    !is.na(birth_year) & !is.na(conception_season)
]
model_births[, complication_eligible :=
               !is.na(GDM) & !is.na(HDP) &
               preexisting_diabetes == 0L & chronic_hypertension == 0L]

covariates <- c(
  "maternal_race", "maternal_ethnicity", "maternal_education",
  "paternal_race", "paternal_ethnicity", "paternal_education",
  "maternal_age", "bmi_category", "previous_preterm", "wic", "payment"
)

sample_qa <- model_births[, .(
  births = .N,
  term_births = sum(gest_weeks >= 37),
  complication_eligible = sum(complication_eligible %in% TRUE),
  PTB_pct = 100 * mean(PTB),
  vPTB_pct = 100 * mean(vPTB),
  LBW_pct = 100 * mean(LBW),
  vLBW_pct = 100 * mean(vLBW),
  term_LBW_pct = 100 * mean(tLBW[gest_weeks >= 37]),
  GDM_pct = 100 * mean(GDM, na.rm = TRUE),
  HDP_pct = 100 * mean(HDP, na.rm = TRUE)
)]
print(sample_qa)

fwrite(sample_qa, "data/outputs/Models/corrected_model_sample_qa.csv")
log_progress("recode", sprintf("model base cohort contains %s births",
                                format(nrow(model_births), big.mark = ",")))
```

## Corrected exposure prevalence

```{r prevalence}
prevalence_columns <- exposure_columns
corrected_prevalence <- rbindlist(lapply(prevalence_columns, function(variable) {
  data.table(
    exposure = variable,
    n_eligible = sum(!is.na(model_births[[variable]])),
    n_exposed = sum(model_births[[variable]] == 1L, na.rm = TRUE),
    prevalence = mean(model_births[[variable]], na.rm = TRUE)
  )
}))
corrected_prevalence[, manuscript_any_exposure_prevalence := 0.808]
corrected_prevalence[, prevalence_difference_percentage_points :=
                       100 * (prevalence - manuscript_any_exposure_prevalence)]
print(corrected_prevalence)
fwrite(corrected_prevalence,
       "data/outputs/Models/corrected_exposure_prevalence.csv")
```

## Multiple imputation

The imputation model includes all analysis covariates, complete birthweight and
gestational-age measures, corrected exposures, and the temporal fixed effects.
Only covariates are imputed; outcomes, exposures, and fixed effects are never
imputed. To make MICE feasible for more than four million births, its regression
equations are estimated on a reproducible random training sample of
`r format(params$imputation_training_rows, big.mark = ",")` records. MICE's
`ignore` mechanism then uses those fitted equations to impute all records. The
health-outcome models continue to use the complete eligible cohort.

```{r imputation}
imputation_predictors <- c(
  "birthweight_g", "gest_weeks", exposure_columns,
  "birth_year", "conception_season"
)
imputation_columns <- unique(c(covariates, imputation_predictors))
imputation_data <- as.data.frame(model_births[, ..imputation_columns])

training_rows <- min(
  as.integer(params$imputation_training_rows), nrow(imputation_data)
)
if (is.na(training_rows) || training_rows < 1000L) {
  stop("`imputation_training_rows` must be at least 1,000.", call. = FALSE)
}
set.seed(as.integer(params$seed) + 1L)
training_index <- sample.int(nrow(imputation_data), training_rows)
imputation_ignore <- rep(TRUE, nrow(imputation_data))
imputation_ignore[training_index] <- FALSE

method <- mice::make.method(imputation_data)
method[] <- ""
method["maternal_age"] <- "pmm"
binary_covariates <- c(
  "maternal_ethnicity", "paternal_ethnicity", "previous_preterm", "wic"
)
method[binary_covariates] <- "logreg"
multicategory_covariates <- setdiff(covariates,
                                    c("maternal_age", binary_covariates))
method[multicategory_covariates] <- "polyreg"

predictor_matrix <- mice::make.predictorMatrix(imputation_data)
predictor_matrix[,] <- 0L
predictor_matrix[covariates, setdiff(imputation_columns, covariates)] <- 1L
predictor_matrix[covariates, covariates] <- 1L
diag(predictor_matrix) <- 0L

log_progress(
  "imputation",
  sprintf("starting MICE: m=%s, maxit=%s, training rows=%s, imputed rows=%s",
          params$imputations, params$mice_iterations,
          format(training_rows, big.mark = ","),
          format(nrow(imputation_data), big.mark = ","))
)
imputation_started <- Sys.time()
imputed <- mice::mice(
  imputation_data,
  m = as.integer(params$imputations),
  maxit = as.integer(params$mice_iterations),
  method = method,
  predictorMatrix = predictor_matrix,
  ignore = imputation_ignore,
  seed = as.integer(params$seed),
  printFlag = TRUE,
  nnet.MaxNWts = as.integer(params$polyreg_max_weights),
  nnet.trace = FALSE
)
log_progress(
  "imputation",
  paste("completed in", format_duration(as.numeric(difftime(
    Sys.time(), imputation_started, units = "secs"
  ))))
)

if (isTRUE(params$save_imputation_object)) {
  saveRDS(imputed, "data/outputs/Models/corrected_birth_mice.rds",
          compress = "xz")
}
```

## Model plan

```{r model-plan}
model_plan <- rbindlist(list(
  CJ(outcome = c("PTB", "vPTB", "LBW", "vLBW"), trimester = c("T1", "T2")),
  data.table(outcome = "tLBW", trimester = c("T1", "T2", "T3")),
  data.table(outcome = c("GDM", "HDP"), trimester = "T1")
))
model_plan[, sample := fcase(
  outcome == "tLBW", "term",
  outcome %in% c("GDM", "HDP"), "complication",
  default = "birth"
)]

thresholds <- 34L
if (isTRUE(params$fit_55kt_sensitivity)) thresholds <- c(34L, 55L)
model_plan <- model_plan[, .(threshold_kt = thresholds),
                         by = .(outcome, trimester, sample)]
model_plan[, exposure := paste0("exposed_window_", trimester, "_", threshold_kt)]
print(model_plan)
```

## Fit imputed logistic models

`fixest::feglm` fits the same binomial-logit likelihood as ordinary logistic
regression while absorbing birth-year and conception-season fixed effects
efficiently.

```{r fit-models}
fit_one_model <- function(data, outcome, exposure, sample) {
  keep <- rep(TRUE, nrow(data))
  if (sample == "term") keep <- data$gest_weeks >= 37
  if (sample == "complication") keep <- data$complication_eligible %in% TRUE

  analysis <- data[keep & !is.na(data[[outcome]]) & !is.na(data[[exposure]])]
  rhs <- paste(c(exposure, covariates), collapse = " + ")
  formula <- as.formula(paste0(
    outcome, " ~ ", rhs, " | birth_year + conception_season"
  ))
  fit <- fixest::feglm(formula, data = analysis, family = "binomial")
  variance <- fixest::vcov(fit)[exposure, exposure]

  data.table(
    beta = unname(coef(fit)[exposure]),
    within_variance = unname(variance),
    n = nobs(fit),
    events = sum(analysis[[outcome]] == 1L, na.rm = TRUE)
  )
}

model_results <- vector("list", nrow(model_plan) * params$imputations)
result_index <- 0L
model_started <- Sys.time()

for (imp in seq_len(params$imputations)) {
  log_progress("models", sprintf("starting imputation %s/%s",
                                  imp, params$imputations))
  completed_covariates <- as.data.table(mice::complete(imputed, imp))
  completed <- copy(model_births)
  completed[, (covariates) := completed_covariates[, ..covariates]]

  for (j in seq_len(nrow(model_plan))) {
    result_index <- result_index + 1L
    spec <- model_plan[j]
    result <- fit_one_model(
      completed, spec$outcome, spec$exposure, spec$sample
    )
    model_results[[result_index]] <- cbind(
      data.table(imputation = imp), spec, result
    )
    log_progress(
      "models",
      sprintf(
        "imputation %s/%s, model %s/%s: %s %s %s kt",
        imp, params$imputations, j, nrow(model_plan),
        spec$outcome, spec$trimester, spec$threshold_kt
      )
    )
  }
  completed_results <- rbindlist(model_results[seq_len(result_index)])
  fwrite(
    completed_results,
    "data/outputs/Models/corrected_model_results_checkpoint.csv"
  )
  log_progress(
    "models",
    sprintf("checkpoint saved after imputation %s/%s",
            imp, params$imputations)
  )
  rm(completed, completed_covariates, completed_results)
  gc()
}

imputation_model_results <- rbindlist(model_results)
fwrite(imputation_model_results,
       "data/outputs/Models/corrected_model_results_by_imputation.csv")
log_progress(
  "models",
  paste("all imputed models completed in", format_duration(as.numeric(difftime(
    Sys.time(), model_started, units = "secs"
  ))))
)
```

## Pool estimates using Rubin's rules

```{r pool}
pool_rubin <- function(beta, within_variance, n, events) {
  m <- length(beta)
  q_bar <- mean(beta)
  u_bar <- mean(within_variance)
  b <- if (m > 1L) var(beta) else 0
  total_variance <- u_bar + (1 + 1 / m) * b
  degrees_freedom <- if (b > 0) {
    (m - 1) * (1 + u_bar / ((1 + 1 / m) * b))^2
  } else {
    Inf
  }
  critical <- if (is.finite(degrees_freedom))
    qt(0.975, df = degrees_freedom) else qnorm(0.975)

  data.table(
    beta = q_bar,
    standard_error = sqrt(total_variance),
    degrees_freedom = degrees_freedom,
    OR = exp(q_bar),
    conf_low = exp(q_bar - critical * sqrt(total_variance)),
    conf_high = exp(q_bar + critical * sqrt(total_variance)),
    n = round(mean(n)),
    events = round(mean(events))
  )
}

pooled_results <- imputation_model_results[,
  pool_rubin(beta, within_variance, n, events),
  by = .(outcome, trimester, sample, threshold_kt, exposure)
]
pooled_results[, estimate := sprintf("%.3f (%.3f, %.3f)",
                                      OR, conf_low, conf_high)]
fwrite(pooled_results,
       "data/outputs/Models/corrected_model_estimates_pooled.csv")
print(pooled_results)
```

## Complete-case sensitivity analysis

```{r complete-case}
complete_case_results <- data.table()
if (isTRUE(params$fit_complete_case_sensitivity)) {
  complete_case_results <- rbindlist(lapply(seq_len(nrow(model_plan)), function(j) {
    spec <- model_plan[j]
    complete <- complete.cases(model_births[, c(covariates, spec$exposure), with = FALSE])
    analysis <- model_births[complete]
    fit <- fit_one_model(analysis, spec$outcome, spec$exposure, spec$sample)
    se <- sqrt(fit$within_variance)
    cbind(spec, data.table(
      OR = exp(fit$beta),
      conf_low = exp(fit$beta - qnorm(0.975) * se),
      conf_high = exp(fit$beta + qnorm(0.975) * se),
      n = fit$n,
      events = fit$events
    ))
  }))
  fwrite(complete_case_results,
         "data/outputs/Models/corrected_model_estimates_complete_case.csv")
}
```

## Comparison with estimates reported in the manuscript

The PDF does not contain Tables 2-4, so only estimates quoted in the Results
text can be compared numerically. Unquoted manuscript cells remain missing
rather than being reconstructed from rounded prose.

```{r manuscript-comparison}
manuscript_results <- data.table(
  outcome = c("PTB", "PTB", "vPTB", "tLBW", "tLBW", "tLBW",
              "vLBW", "GDM", "HDP"),
  trimester = c("T1", "T2", "T2", "T1", "T2", "T3", "T2", "T1", "T1"),
  old_OR = c(1.012, 1.008, 0.919, 1.023, 1.034, 0.966,
             0.914, 1.050, 1.127),
  old_low = c(1.004, 1.000, 0.902, 1.006, 1.019, 0.950,
              0.896, 1.041, 1.114),
  old_high = c(1.020, 1.016, 0.937, 1.040, 1.051, 0.982,
               0.933, 1.068, 1.141)
)

comparison <- merge(
  pooled_results[threshold_kt == 34L],
  manuscript_results,
  by = c("outcome", "trimester"),
  all.x = TRUE
)
comparison[, `:=`(
  old_estimate = fifelse(
    !is.na(old_OR), sprintf("%.3f (%.3f, %.3f)", old_OR, old_low, old_high),
    NA_character_
  ),
  OR_absolute_difference = OR - old_OR,
  log_OR_difference = log(OR) - log(old_OR),
  percent_change_in_OR = 100 * (OR / old_OR - 1),
  old_direction = fcase(old_OR > 1, "above null", old_OR < 1, "below null",
                        default = NA_character_),
  corrected_direction = fcase(OR > 1, "above null", OR < 1, "below null",
                              default = "null"),
  direction_changed = !is.na(old_OR) & sign(log(OR)) != sign(log(old_OR)),
  corrected_excludes_null = conf_low > 1 | conf_high < 1,
  old_excludes_null = !is.na(old_OR) & (old_low > 1 | old_high < 1)
)]

fwrite(comparison,
       "data/outputs/Models/manuscript_vs_corrected_model_comparison.csv")
print(comparison[, .(
  outcome, trimester, old_estimate, corrected = estimate,
  percent_change_in_OR, direction_changed,
  old_excludes_null, corrected_excludes_null
)])
```

## Comparison figure

```{r comparison-figure, fig.width=10, fig.height=7}
plot_data <- rbindlist(list(
  manuscript_results[, .(
    outcome, trimester, specification = "Manuscript (reported)",
    OR = old_OR, conf_low = old_low, conf_high = old_high
  )],
  pooled_results[threshold_kt == 34L &
                   paste(outcome, trimester) %in%
                   paste(manuscript_results$outcome, manuscript_results$trimester), .(
    outcome, trimester, specification = "Corrected 34 kt + year/season FE",
    OR, conf_low, conf_high
  )]
))
plot_data[, label := paste(outcome, trimester)]

comparison_plot <- ggplot(
  plot_data,
  aes(x = OR, y = reorder(label, OR), color = specification)
) +
  geom_vline(xintercept = 1, linetype = 2, color = "grey50") +
  geom_errorbarh(aes(xmin = conf_low, xmax = conf_high),
                 position = position_dodge(width = 0.5), height = 0.15) +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  scale_x_log10() +
  labs(
    x = "Adjusted odds ratio (log scale)", y = NULL, color = NULL,
    title = "Manuscript and corrected tropical-cyclone exposure estimates",
    subtitle = "Only manuscript estimates quoted in the Results text are shown"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
print(comparison_plot)

ggsave(
  "figs/model-comparison/manuscript-vs-corrected-forest-plot.png",
  comparison_plot, width = 10, height = 7, dpi = 300
)
```

## Automated summary

```{r summary}
reported_comparison <- comparison[!is.na(old_OR)]
summary_counts <- reported_comparison[, .(
  manuscript_estimates_compared = .N,
  direction_changes = sum(direction_changed),
  manuscript_intervals_excluding_null = sum(old_excludes_null),
  corrected_intervals_excluding_null = sum(corrected_excludes_null),
  median_absolute_percent_OR_change = median(abs(percent_change_in_OR))
)]
print(summary_counts)
fwrite(summary_counts, "data/outputs/Models/model_comparison_summary.csv")

log_progress("complete", "model refit and comparison finished")
```

## Interpretation notes

Differences between manuscript and corrected estimates reflect several changes
at once: correction of the original space-time join, replacement of RMW track
buffers with time-specific wind radii, replacement of the three-day phases
with a ±7-day local-impact window, use of 34 kt as the primary threshold, and
addition of calendar-year and season fixed effects. The comparison therefore
should not be interpreted as isolating a single modeling change.

The manuscript's statement “five iterations” is insufficient to reproduce its
imputation exactly. This workflow records the number of imputed datasets,
iterations, seed, methods, and predictor matrix. If the original analysts can
provide their MICE specification or Tables 2-4, update those inputs before the
final manuscript revision.
