---
title: "ThreatDays_ExcessMortality"
output: html_document
editor_options: 
  chunk_output_type: console
---

Read the population data, TIC model data, and Poisson model output (Zip_Population & ThreatDays)
```{r}
# Load population
zip_pop <- readRDS(
  "data/population/florida_zip_population_1985_2022.rds"
)

TIC.sf <- readRDS(
  "data/outputs/Results/TIC_34kt.rds"
)

# Add storm id

library(dplyr)

TIC <- TIC.sf %>%
  st_drop_geometry() %>%   # geometry not needed for ID assignment
  mutate(
    storm_id = paste0(Storm_Name, "_", Year)
  )

# Load statistical model results
m_event <- readRDS(
  "models/m_event_34kt.rds"
)

# Sanity checks
stopifnot(all(c("year", "zip_code", "population") %in% names(zip_pop)))
stopifnot(is.numeric(zip_pop$population))
stopifnot(nchar(zip_pop$zip_code[!is.na(zip_pop$zip_code)][1]) == 5)

# Store model coefficients separately for quick access:
#library(broom)
#coef_event <- tidy(m_event)
#saveRDS(
#  coef_event,
#  "models/m_event_34kt_tidy.rds"
#)
```

Read in the daily panel data. Make sure columns have same type
```{r}
panel_zip_day <- readr::read_csv("data/outputs/Deaths/daily_panel_34kt.csv")

library(dplyr)
library(stringr)

panel_zip_day <- panel_zip_day %>%
  mutate(
    zip  = sprintf("%05d", zip),  # numeric → character ZIP
    year = as.integer(format(date, "%Y"))
  )
```

Merge population onto ZIP–day panel
```{r}
library(stringr)

panel_zip_day <- panel_zip_day %>%
  mutate(
    zip  = str_pad(zip, width = 5, side = "left", pad = "0"),
    year = as.integer(format(date, "%Y"))
  ) %>%
  left_join(
    zip_pop,
    by = c("zip" = "zip_code", "year" = "year")
  ) %>%
  mutate(
    population = if_else(is.na(population), 0, population)
  )

# Sanity checks
stopifnot(all(nchar(panel_zip_day$zip) == 5))
stopifnot(all(panel_zip_day$population >= 0))
```

Add storm IDs to the panel_zip_day
```{r}
TIC_join <- TIC %>%
  select(Date, rel_day, storm_id)

# Ensure types match
panel_zip_day <- panel_zip_day %>%
  mutate(
    date    = as.Date(date),
    rel_day = as.integer(rel_day)
  )

# Join
panel_zip_day <- panel_zip_day %>%
  left_join(
    TIC_join,
    by = c("date" = "Date", "rel_day" = "rel_day")
  )

# Sanity check (should be zeros)

panel_zip_day %>%
  filter(rel_day_f != "None", is.na(storm_id)) %>%
  nrow()
panel_zip_day %>%
  filter(rel_day_f == "None", !is.na(storm_id)) %>%
  nrow()

# One storm ID per storm window (all ones)
TIC %>%
  distinct(Storm_Name, Year, storm_id) %>%
  count(storm_id) %>%
  summary()
```

Extract event‑time excess fractions from the model output
```{r}
library(dplyr)
library(stringr)

beta_hat <- stats::coef(m_event)

event_effects <- tibble(
  term = names(beta_hat),
  beta = beta_hat
) %>%
  filter(str_detect(term, "^rel_day_f::")) %>%
  mutate(
    rel_day = as.integer(str_remove(term, "rel_day_f::")),
    rr = exp(beta),
    excess_frac = rr - 1
  ) %>%
  select(rel_day, excess_frac)
```

Panel excess
```{r}
panel_excess <- panel_zip_day %>%
  left_join(event_effects, by = "rel_day") %>%
  mutate(
    excess_frac = if_else(is.na(excess_frac), 0, excess_frac),
    baseline_deaths = deaths / (1 + excess_frac),
    excess_deaths   = baseline_deaths * excess_frac
  )
```

Aggregate ZIP–day → ZIP–storm
```{r}
zip_storm_excess <- panel_excess %>%
#  filter(rel_day_f != "None") %>%
  filter(rel_day %in% -8:-1) %>%
  group_by(storm_id, zip) %>%
  summarise(
    excess_deaths_zip = sum(excess_deaths, na.rm = TRUE),
    .groups = "drop"
  )
```

Aggregate ZIP–storm → storm 
```{r}
storm_excess <- zip_storm_excess %>%
  group_by(storm_id) %>%
  summarise(
    excess_deaths = round(sum(excess_deaths_zip, na.rm = TRUE)),
    .groups = "drop"
  )
```

Include storm meta data
```{r}
storm_meta <- TIC.sf %>%
  st_drop_geometry() %>%
  filter(rel_day == 0) %>%
  transmute(
    storm_id = paste0(Storm_Name, "_", Year),
    storm_name = Storm_Name,
    storm_category = Storm_Category,
    impact_date = Date
  )

storm_excess_chrono <- storm_excess %>%
  left_join(storm_meta, by = "storm_id") %>%
  arrange(impact_date)
```

Normalize per 1M people
```{r}
storm_population <- panel_excess %>%
  filter(rel_day %in% -8:-1) %>%      # same window as excess
  group_by(storm_id, zip) %>%
  summarise(
    zip_population = max(population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(storm_id) %>%
  summarise(
    exposed_population = sum(zip_population, na.rm = TRUE),
    .groups = "drop"
  )

storm_excess_norm <- storm_excess_chrono %>%
  left_join(storm_population, by = "storm_id") %>%
  mutate(
    excess_per_1M = 1e6 * excess_deaths / exposed_population
  ) %>%
  arrange(impact_date)

# Write to CSV
fwrite(
  storm_excess_norm,
  "data/outputs/storm_excess_norm_chrono.csv"
)
```

Plot over time
```{r}
library(ggplot2)

ggplot(storm_excess_norm,
       aes(x = impact_date, y = excess_per_100k)) +
  geom_line(color = "steelblue", linewidth = 0.7) +
  geom_point(color = "steelblue", size = 1.8) +
  labs(
    x = "Storm impact date",
    y = "Excess deaths per 100,000 (days −8 to −1)",
    title = "Pre‑impact excess mortality per storm",
    subtitle = "Normalized by exposed ZIP‑code population"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
library(dplyr)
library(lubridate)

storm_year_avg <- storm_excess_norm %>%
  mutate(year = year(impact_date)) %>%
  group_by(year) %>%
  summarise(
    mean_excess_per_1M = mean(excess_per_1M, na.rm = TRUE),
    n_storms = n(),
    .groups = "drop"
  )

ggplot(storm_year_avg,
       aes(x = year, y = mean_excess_per_1M)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(aes(size = n_storms),
             color = "steelblue", alpha = 0.8) +
  scale_size_continuous(name = "Number of storms") +
  labs(
    x = "Year",
    y = "Average excess deaths per 1M (pre-impact)",
    title = "Annual average pre-impact excess mortality",
    subtitle = "Point size reflects number of storms per year"
  ) +
  theme_minimal(base_size = 13)
```


