---
title: "Time-to-event models for tropical-cyclone exposure and preterm delivery"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  corrected_data: "data/births/processed/All_Births_Corrected_Storm_Exposure-trimester-eligible-v2.rds"
  event_data: "data/births/derived/birth-storm-events-time-specific-trimester-eligible-v2.csv"
  output_dir: "models/births/time-to-event"
  entry_week: 20
  acute_days: [3, 7]
  fit_intensity_models: true
---

## Purpose

These models replace trimester-level ever/never exposure with gestational-age
time-to-event analyses. A pregnancy contributes unexposed person-time until the
residence-specific local storm impact. Storms after delivery never define
exposure. The primary time-varying covariate switches on at local impact and
remains on; acute sensitivity covariates are on from impact through either three
or seven days after impact. The pre-impact half of the descriptive ±7-day window is not used
as a causal time-varying exposure because it would condition current risk on a
future storm.

Pregnancies enter follow-up at 20 completed weeks. PTB follow-up ends at
delivery or 37 weeks; vPTB follow-up ends at delivery or 32 weeks. County and
conception year-month are controlled through joint Cox strata, and robust
standard errors are clustered by county. This first survival analysis uses
complete covariate cases so its purpose—diagnosing gestational-time selection—
is not mixed with another imputation exercise.

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("data.table", "survival")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Missing packages: ", paste(missing_packages, collapse = ", "))
}

library(data.table)
library(survival)

output_dir <- as.character(params$output_dir)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
progress_file <- file.path(output_dir, "survival-model-progress.log")
if (file.exists(progress_file)) file.remove(progress_file)

log_progress <- function(stage, detail) {
  line <- sprintf(
    "%s | %-20s | %s",
    format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z"), stage, detail
  )
  message(line)
  write(line, progress_file, append = TRUE)
  invisible(line)
}

entry_day <- as.integer(params$entry_week) * 7L
acute_windows <- sort(unique(as.integer(unlist(params$acute_days))))
stopifnot(entry_day >= 0L, length(acute_windows) >= 1L, all(acute_windows >= 0L))
log_progress("setup", "time-to-event workflow started")
```

## Read and recode the birth cohort

```{r births}
stopifnot(file.exists(params$corrected_data), file.exists(params$event_data))
births <- readRDS(params$corrected_data)
setDT(births)

source_columns <- c(
  "ID", "DATE_OF_BIRTH", "conception_date", "GESTATION_WEEKS",
  "PLURALITY_CODE", "county_geoid", "conception_year", "conception_month",
  "exposure_eligible_T1", "Mother_CalculatedRace",
  "Mother_CalculatedHisp", "MOTHER_EDCODE", "Father_CalculatedRace",
  "Father_CalculatedHisp", "FATHER_EDCODE", "MOTHER_AGE",
  "PrePregnancy_BMI", "MR_PREV_PRETERM", "MOTHER_WIC_YESNO",
  "PRINCIPAL_SRCPAY_CODE"
)
missing_columns <- setdiff(source_columns, names(births))
if (length(missing_columns)) {
  stop("Corrected data missing: ", paste(missing_columns, collapse = ", "))
}
births <- births[, ..source_columns]

normalize_missing_code <- function(
    x, missing = c("", "NULL", "NA", "N/A", "U", "UNK", "UNKNOWN")) {
  value <- trimws(toupper(as.character(x)))
  value[value %in% missing] <- NA_character_
  value
}

binary_yes_no <- function(x) {
  value <- normalize_missing_code(x)
  fcase(
    value %in% c("Y", "YES", "1", "TRUE"), 1L,
    value %in% c("N", "NO", "0", "FALSE"), 0L,
    default = NA_integer_
  )
}

collapse_race <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 1L, "White", value == 2L, "Black",
    !is.na(value) & value != 99L, "Other", default = NA_character_
  ), levels = c("White", "Black", "Other"))
}

collapse_ethnicity <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 0L, "Non-Hispanic", value %in% 1:98, "Hispanic",
    default = NA_character_
  ), levels = c("Non-Hispanic", "Hispanic"))
}

collapse_education <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value %in% 1:2, "Less than high school",
    value == 3L, "High school or GED",
    value %in% 4:5, "Some college or associate degree",
    value %in% 6:8, "Bachelor degree or higher",
    default = NA_character_
  ), levels = c(
    "Less than high school", "High school or GED",
    "Some college or associate degree", "Bachelor degree or higher"
  ))
}

collapse_payment <- function(x) {
  value <- suppressWarnings(as.integer(normalize_missing_code(x)))
  factor(fcase(
    value == 1L, "Medicaid", value == 2L, "Private insurance",
    value == 3L, "Self-pay", !is.na(value) & value != 9L, "Other",
    default = NA_character_
  ), levels = c("Medicaid", "Private insurance", "Self-pay", "Other"))
}

births[, `:=`(
  DATE_OF_BIRTH = as.IDate(DATE_OF_BIRTH),
  conception_date = as.IDate(conception_date),
  gest_weeks_reported = suppressWarnings(as.numeric(GESTATION_WEEKS)),
  singleton = suppressWarnings(as.integer(PLURALITY_CODE)) == 1L,
  maternal_race = collapse_race(Mother_CalculatedRace),
  maternal_ethnicity = collapse_ethnicity(Mother_CalculatedHisp),
  maternal_education = collapse_education(MOTHER_EDCODE),
  paternal_race = collapse_race(Father_CalculatedRace),
  paternal_ethnicity = collapse_ethnicity(Father_CalculatedHisp),
  paternal_education = collapse_education(FATHER_EDCODE),
  maternal_age = suppressWarnings(as.numeric(MOTHER_AGE)),
  previous_preterm = factor(binary_yes_no(MR_PREV_PRETERM), levels = 0:1),
  wic = factor(binary_yes_no(MOTHER_WIC_YESNO), levels = 0:1),
  payment = collapse_payment(PRINCIPAL_SRCPAY_CODE),
  county_geoid = factor(county_geoid),
  conception_year_month = factor(fifelse(
    !is.na(conception_year) & conception_month %in% 1:12,
    sprintf("%04d-%02d", as.integer(conception_year),
            as.integer(conception_month)),
    NA_character_
  ))
)]

bmi_numeric <- suppressWarnings(as.numeric(normalize_missing_code(
  births$PrePregnancy_BMI
)))
births[, bmi_category := cut(
  bmi_numeric, breaks = c(-Inf, 18.5, 25, 30, Inf), right = FALSE,
  labels = c("Underweight", "Normal", "Overweight", "Obesity")
)]

covariates <- c(
  "maternal_race", "maternal_ethnicity", "maternal_education",
  "paternal_race", "paternal_ethnicity", "paternal_education",
  "maternal_age", "bmi_category", "previous_preterm", "wic", "payment"
)

births[, gestational_day := as.integer(DATE_OF_BIRTH - conception_date)]
births[, reported_day := round(gest_weeks_reported * 7)]
births[, county_month_stratum := interaction(
  county_geoid, conception_year_month, drop = TRUE, lex.order = TRUE
)]

date_qa <- births[, .(
  births = .N,
  valid_gestational_days = sum(between(gestational_day, entry_day, 45L * 7L)),
  median_date_minus_reported_days = median(
    gestational_day - reported_day, na.rm = TRUE
  ),
  p01_date_minus_reported_days = quantile(
    gestational_day - reported_day, 0.01, na.rm = TRUE
  ),
  p99_date_minus_reported_days = quantile(
    gestational_day - reported_day, 0.99, na.rm = TRUE
  )
)]
fwrite(date_qa, file.path(output_dir, "gestational-date-qa.csv"))

complete <- complete.cases(births[, c(
  covariates, "gestational_day", "county_geoid",
  "conception_year_month", "county_month_stratum"
), with = FALSE])

analysis_births <- births[
  exposure_eligible_T1 %in% TRUE & singleton %in% TRUE & complete &
    between(gestational_day, entry_day, 45L * 7L)
]
analysis_births <- analysis_births[, c(
  "ID", "DATE_OF_BIRTH", "conception_date", "gestational_day",
  "gest_weeks_reported", "county_geoid", "conception_year_month",
  "county_month_stratum", covariates
), with = FALSE]
setkey(analysis_births, ID)

cohort_qa <- data.table(
  source_births = nrow(births),
  complete_case_survival_births = nrow(analysis_births),
  counties = uniqueN(analysis_births$county_geoid),
  county_month_strata = uniqueN(analysis_births$county_month_stratum)
)
fwrite(cohort_qa, file.path(output_dir, "survival-cohort-qa.csv"))
log_progress("births", sprintf("retained %s complete-case singleton births",
                                format(nrow(analysis_births), big.mark = ",")))
rm(births)
gc()
```

## Read local storm impacts and define intensity

The intensity model distinguishes tropical-storm observations (34–63 kt
center intensity) from category-1-or-higher observations (at least 64 kt).
Both indicators are time varying and are entered jointly, so their coefficients
can be contrasted directly.

```{r events}
events_34 <- fread(
  params$event_data,
  select = c("ID", "SID", "threshold_kt", "USA_WIND", "local_impact_date")
)[threshold_kt == 34L]
events_34[, local_impact_date := as.IDate(local_impact_date)]
events_34 <- merge(
  events_34,
  analysis_births[, .(ID, conception_date, DATE_OF_BIRTH)],
  by = "ID", all = FALSE, sort = FALSE
)
events_34[, event_day := as.integer(local_impact_date - conception_date)]

# Impacts on or after the delivery date cannot precede the delivery event.
events_34 <- events_34[
  !is.na(event_day) & !is.na(USA_WIND) &
    local_impact_date < DATE_OF_BIRTH & event_day >= 0L
]
events_34[, storm_class := fifelse(USA_WIND >= 64, "hu", "ts")]
events_34 <- unique(events_34[, .(ID, SID, event_day, storm_class)])

event_qa <- events_34[, .(
  birth_storm_impacts = .N,
  exposed_births = uniqueN(ID),
  tropical_storm_impacts = sum(storm_class == "ts"),
  hurricane_impacts = sum(storm_class == "hu"),
  first_event_day = min(event_day),
  last_event_day = max(event_day)
)]
fwrite(event_qa, file.path(output_dir, "survival-event-qa.csv"))
log_progress("events", sprintf("retained %s pre-delivery local impacts",
                                format(nrow(events_34), big.mark = ",")))
```

## Construct counting-process intervals

```{r intervals}
make_risk_base <- function(outcome, cutoff_week) {
  cutoff_day <- as.integer(cutoff_week) * 7L
  risk <- copy(analysis_births)
  risk[, `:=`(
    outcome = outcome,
    cutoff_day = cutoff_day,
    followup_stop = pmin(gestational_day, cutoff_day),
    status = as.integer(gestational_day < cutoff_day)
  )]
  risk[followup_stop > entry_day]
}

make_event_long <- function(risk, family) {
  relevant <- merge(
    events_34[, .(ID, event_day, storm_class)],
    risk[, .(ID, followup_stop)], by = "ID", all = FALSE, sort = FALSE
  )[event_day < followup_stop]

  if (family == "any") {
    relevant[, exposure_var := "any"]
  } else if (family == "intensity") {
    relevant[, exposure_var := storm_class]
  } else {
    stop("Unknown exposure family: ", family)
  }
  unique(relevant[, .(ID, event_day, exposure_var, followup_stop)])
}

build_intervals <- function(risk, family, structure, acute_days = NA_integer_) {
  exposure_vars <- if (family == "any") "any" else c("ts", "hu")
  event_long <- make_event_long(risk, family)

  if (structure == "cumulative") {
    event_long <- event_long[, .(event_day = min(event_day)),
                             by = .(ID, exposure_var, followup_stop)]
    event_long[, `:=`(
      start = pmax(event_day, entry_day),
      end = NA_integer_
    )]
    changes <- event_long[start < followup_stop,
      .(ID, time = start, exposure_var, delta = 1L)]
  } else if (structure == "acute") {
    acute_length <- as.integer(acute_days) + 1L
    stopifnot(length(acute_length) == 1L, !is.na(acute_length), acute_length >= 1L)
    event_long[, `:=`(
      start = pmax(event_day, entry_day),
      end = pmin(event_day + acute_length, followup_stop)
    )]
    event_long <- event_long[start < end]
    changes <- rbindlist(list(
      event_long[, .(ID, time = start, exposure_var, delta = 1L)],
      event_long[, .(ID, time = end, exposure_var, delta = -1L)]
    ))
  } else {
    stop("Unknown structure: ", structure)
  }

  if (nrow(changes)) {
    change_wide <- dcast(
      changes, ID + time ~ exposure_var, value.var = "delta",
      fun.aggregate = sum, fill = 0L
    )
  } else {
    change_wide <- data.table(ID = character(), time = integer())
  }
  for (variable in exposure_vars) {
    if (!variable %in% names(change_wide)) change_wide[, (variable) := 0L]
  }

  points <- rbindlist(list(
    risk[, .(ID, time = entry_day)],
    risk[, .(ID, time = followup_stop)],
    change_wide[, .(ID, time)]
  ))
  points <- unique(points)
  points <- merge(points, change_wide, by = c("ID", "time"),
                  all.x = TRUE, sort = FALSE)
  for (variable in exposure_vars) {
    set(points, which(is.na(points[[variable]])), variable, 0L)
  }
  setorder(points, ID, time)
  points[, (exposure_vars) := lapply(.SD, cumsum),
         by = ID, .SDcols = exposure_vars]
  points[, tstop := shift(time, type = "lead"), by = ID]
  points <- points[!is.na(tstop) & time < tstop]
  setnames(points, "time", "tstart")
  for (variable in exposure_vars) {
    points[, (paste0(variable, "_td")) := as.integer(get(variable) > 0L)]
  }
  points[, (exposure_vars) := NULL]

  intervals <- merge(
    points, risk, by = "ID", all.x = TRUE, sort = FALSE
  )
  intervals[, event := as.integer(status == 1L & tstop == followup_stop)]
  setorder(intervals, ID, tstart)

  duration_qa <- intervals[, .(
    observed_duration = sum(tstop - tstart),
    expected_duration = first(followup_stop) - entry_day,
    events = sum(event)
  ), by = ID]
  stopifnot(
    all(intervals$tstart < intervals$tstop),
    all(duration_qa$observed_duration == duration_qa$expected_duration),
    all(duration_qa$events <= 1L),
    sum(intervals$event) == sum(risk$status)
  )
  intervals
}
```

## Fit Cox models

```{r models}
covariate_rhs <- paste(covariates, collapse = " + ")

extract_cox <- function(fit, terms, contrast = NULL) {
  coefficients <- coef(fit)
  variance <- vcov(fit)
  rows <- rbindlist(lapply(terms, function(term) {
    beta <- unname(coefficients[term])
    se <- sqrt(variance[term, term])
    data.table(
      contrast = term, beta = beta, standard_error = se,
      HR = exp(beta), conf_low = exp(beta - qnorm(0.975) * se),
      conf_high = exp(beta + qnorm(0.975) * se),
      p_value = 2 * pnorm(-abs(beta / se))
    )
  }))
  if (!is.null(contrast)) {
    beta <- unname(coefficients[contrast[1]] - coefficients[contrast[2]])
    se <- sqrt(
      variance[contrast[1], contrast[1]] +
        variance[contrast[2], contrast[2]] -
        2 * variance[contrast[1], contrast[2]]
    )
    rows <- rbind(rows, data.table(
      contrast = "hu_vs_ts", beta = beta, standard_error = se,
      HR = exp(beta), conf_low = exp(beta - qnorm(0.975) * se),
      conf_high = exp(beta + qnorm(0.975) * se),
      p_value = 2 * pnorm(-abs(beta / se))
    ))
  }
  rows
}

cumulative_plan <- CJ(
  outcome = c("PTB", "vPTB"),
  structure = "cumulative",
  family = c("any", "intensity"),
  sorted = FALSE
)
cumulative_plan[, acute_days := NA_integer_]
acute_plan <- CJ(
  outcome = c("PTB", "vPTB"),
  structure = "acute",
  family = c("any", "intensity"),
  acute_days = acute_windows,
  sorted = FALSE
)
model_plan <- rbindlist(list(cumulative_plan, acute_plan), use.names = TRUE)
if (!isTRUE(params$fit_intensity_models)) {
  model_plan <- model_plan[family == "any"]
}

results <- vector("list", nrow(model_plan))
interval_qa <- vector("list", nrow(model_plan))

for (i in seq_len(nrow(model_plan))) {
  spec <- model_plan[i]
  cutoff_week <- if (spec$outcome == "PTB") 37L else 32L
  risk <- make_risk_base(spec$outcome, cutoff_week)
  intervals <- build_intervals(
    risk, spec$family, spec$structure, acute_days = spec$acute_days
  )

  exposure_terms <- if (spec$family == "any") {
    "any_td"
  } else {
    c("ts_td", "hu_td")
  }
  formula <- as.formula(paste0(
    "Surv(tstart, tstop, event) ~ ",
    paste(c(exposure_terms, covariates), collapse = " + "),
    " + strata(county_month_stratum) + cluster(county_geoid)"
  ))
  fit <- coxph(
    formula, data = intervals, ties = "efron",
    model = FALSE, x = FALSE, y = FALSE
  )
  extracted <- extract_cox(
    fit, exposure_terms,
    contrast = if (spec$family == "intensity") c("hu_td", "ts_td") else NULL
  )
  exposed_birth_count <- intervals[, .(
    ever_exposed = any(rowSums(.SD) > 0L)
  ), by = ID, .SDcols = exposure_terms][ever_exposed == TRUE, .N]
  extracted[, `:=`(
    outcome = spec$outcome,
    structure = spec$structure,
    family = spec$family,
    acute_days = spec$acute_days,
    births = uniqueN(intervals$ID),
    person_intervals = nrow(intervals),
    events = sum(intervals$event),
    exposed_births = exposed_birth_count
  )]
  results[[i]] <- extracted
  interval_qa[[i]] <- data.table(
    outcome = spec$outcome, structure = spec$structure,
    family = spec$family, acute_days = spec$acute_days,
    births = uniqueN(intervals$ID),
    intervals = nrow(intervals), events = sum(intervals$event),
    min_tstart = min(intervals$tstart), max_tstop = max(intervals$tstop)
  )
  fwrite(
    rbindlist(results[seq_len(i)]),
    file.path(output_dir, "time-to-event-estimates-checkpoint.csv")
  )
  log_progress(
    "models",
    sprintf("model %s/%s: %s %s %s%s", i, nrow(model_plan),
            spec$outcome, spec$structure, spec$family,
            ifelse(is.na(spec$acute_days), "", paste0(" day0-", spec$acute_days)))
  )
  rm(fit, intervals, risk)
  gc()
}

survival_results <- rbindlist(results)
survival_results[, estimate := sprintf(
  "%.3f (%.3f, %.3f)", HR, conf_low, conf_high
)]
setcolorder(survival_results, c(
  "outcome", "structure", "acute_days", "family", "contrast", "HR", "conf_low",
  "conf_high", "p_value", "births", "person_intervals", "events",
  "exposed_births", "estimate", "beta", "standard_error"
))
fwrite(survival_results, file.path(output_dir, "time-to-event-estimates.csv"))
fwrite(rbindlist(interval_qa), file.path(output_dir, "time-to-event-interval-qa.csv"))

acute_comparison <- dcast(
  survival_results[structure == "acute"],
  outcome + family + contrast ~ paste0("day0_", acute_days),
  value.var = c("HR", "conf_low", "conf_high", "p_value", "exposed_births")
)
if (all(c("HR_day0_3", "HR_day0_7") %in% names(acute_comparison))) {
  acute_comparison[, percent_HR_difference_0_3_vs_0_7 :=
    100 * (HR_day0_3 / HR_day0_7 - 1)]
}
fwrite(acute_comparison, file.path(output_dir, "acute-window-comparison.csv"))
print(survival_results)
log_progress("complete", "time-to-event workflow finished")
```

## Interpretation

The cumulative model estimates the delivery hazard after any prior local storm
impact. The acute models estimate the hazard during days 0–3 and days 0–7 after
impact, changing only the duration of the active exposure indicator.
`ts_td` and `hu_td` compare each storm-intensity class with no currently active
exposure; `hu_vs_ts` is their direct log-hazard contrast. Because these are
time-varying Cox models, they do not classify births occurring before a later
trimester storm as unexposed controls for that future storm.
